### 2018-10-15 last_paper_AIWC_OpenCL-based_Architecture-Independent_Workload_Characterization.Rmd

---
title: "AIWC: OpenCL-based Architecture-Independent Workload Characterization"
abstract: "

Measuring performance-critical characteristics of application workloads is important both for developers, who must understand and optimize the performance of codes, as well as designers and integrators of HPC systems, who must ensure that compute architectures are suitable for the intended workloads.
However, if these workload characteristics are tied to architectural features that are specific to a particular system, they may not generalize well to alternative or future systems.
An architecture-independent method ensures an accurate characterization of inherent program behaviour, without bias due to architecture-dependent features that vary widely between different types of accelerators.

This work presents the first architecture-independent workload characterization framework for heterogeneous compute platforms, proposing a set of metrics determining the suitability and performance of an application on any parallel HPC architecture.
The tool, AIWC, is a plugin for the open-source Oclgrind simulator.
It supports parallel workloads and is capable of characterizing OpenCL codes currently in use in the supercomputing setting.
AIWC simulates an OpenCL device by directly interpreting LLVM instructions, and the resulting metrics may be used for performance prediction and developer feedback to guide device-specific optimizations.
An evaluation of the metrics collected over a subset of the Extended OpenDwarfs Benchmark Suite is also presented.
"
keywords: "workload characterization, benchmarking, HPC"
bibliography: ./bibliography/bibliography.bib
output:
    bookdown::html_document2:
        fig_caption: yes
    bookdown::pdf_document2:
        fig_caption: yes
---

<!--IEEE needs the keywords to be set here :(-->
\iftoggle{IEEE-BUILD}{
\begin{IEEEkeywords}
workload characterisation, analysis
\end{IEEEkeywords}
}{}

<!--
LLVM-HPC2018 - The Fifth Workshop on the LLVM Compiler Infrastructure in HPC
http://llvm-hpc5-workshop.github.io/

Dallas, Texas, November 12, 2018

LLVM is increasingly used in work targeted at high-performance computing. Research in, and implementation of, program analysis, compilation, execution, and profiling has clearly benefited from the availability of a high-quality, freely-available infrastructure on which to build.

 * Tools for optimization, profiling, and feedback
 * Source code transformation and analysis
 * Gap analyses of open-source LLVM-based tools

IEEE 8-11 page limit
-->

#Introduction

Modern high-performance computing (HPC) systems are typically heterogeneous, with a single node comprising a traditional CPU and an accelerator such as a GPU or many-integrated-core device (MIC).
High bandwidth, low latency interconnects such as the Cray XC50 *Aries*, Fujitsu Post-K *Tofu* and IBM Power9 *Bluelink*, support tighter integration between compute devices on a node.
Some interconnects support multiple different kinds of devices on a single node, for example, *Bluelink* features both NVLink support for Nvidia GPUs and CAPI for other emerging accelerators such as DSPs, FPGAs and MICs.

The OpenCL programming framework is well-suited to such heterogeneous computing environments, as a single OpenCL code may be executed on multiple different device types.
When combined with autotuning, an OpenCL code may exhibit good performance across varied devices. [@spafford2010maestro, @chaimov2014toward, @nugteren2015cltune, @price2017analyzing]

Application codes differ in resource requirements, control structure and available parallelism.
Similarly, compute devices differ in number and capabilities of execution units, processing model, and available resources.
Given performance measurements for particular combinations of codes and devices, it is difficult to generalize to novel combinations.
Hardware designers and HPC integrators would benefit from accurate and systematic performance prediction, for example, in designing an HPC system, to choose a mix of accelerators that are well-suited to the expected workload.

To this end, we present the Architecture Independent Workload Characterization (AIWC) tool.
AIWC simulates the execution of OpenCL kernels to collect architecture-independent features that characterize each code, which may also be used in performance prediction.

AIWC is the first workload characterization tool to support multi-threaded or parallel workloads, which it achieves by collecting metrics that indicate both instruction and thread-level parallelism.
Exploitable coarse-grained parallelism is measured by counting the number of work-items and barriers encountered.
Instructions To Barrier (ITB) and Instructions per Thread (IPT) can be used to indicate workload irregularity or imbalance.

We demonstrate the use of AIWC to characterize a variety of codes in the Extended OpenDwarfs Benchmark Suite [@johnston18opendwarfs].

#Related Work

Oclgrind is an OpenCL device simulator developed by Price and McIntosh-Smith [@price:15] capable of performing simulated kernel execution.
It operates on a restricted LLVM IR known as Standard Portable Intermediate Representation (SPIR) [@kessenich2015], thereby simulating OpenCL kernel code in a hardware agnostic manner.
This architecture independence allows the tool to uncover many portability issues when migrating OpenCL code between devices.
Additionally, Oclgrind comes with a set of tools to detect runtime API errors, race conditions and invalid memory accesses, and generate instruction histograms.
AIWC is added as a tool to Oclgrind and leverages its ability to simulate OpenCL device execution using LLVM IR codes.

AIWC relies on the selection of the instruction set architecture (ISA)-independent features determined by Shao and Brooks [@shao2013isa], which in turn builds on earlier work in microarchitecture-independent workload characterization.
Hoste and Eeckout [@hoste2007microarchitecture] show that although conventional microarchitecture-dependent characteristics are useful in locating performance bottlenecks [@ganesan2008performance,@prakash2008performance], they are misleading when used as a basis on which to differentiate benchmark applications.
Microarchitecture-independent workload characterization and the associated analysis tool, known as MICA, was proposed to collect metrics to characterize an application independent of particular microarchitectural characteristics.
Architecture-dependent characteristics typically include instructions per cycle (IPC) and miss rates -- cache, branch misprediction and translation look-aside buffer (TLB) -- and are collected from hardware performance counter results, typically PAPI.
However, these characteristics fail to distinguish between inherent program behaviour and its mapping to specific hardware features, ignoring critical differences between architectures such as pipeline depth and cache size.
The MICA framework collects independent features including instruction mix, instruction-level parallelism (ILP), register traffic, working-set size, data stream strides and branch predictability.
These feature results are collected using the Pin [@luk2005pin] binary instrumentation tool.
In total 47 microarchitecture-independent metrics are used to characterize an application code.
To simplify analysis and understanding of the data, the authors combine principal component analysis with a genetic algorithm to select eight metrics which account for approximately 80% of the variance in the data set.

A caveat in the MICA approach is that the results presented are not ISA-independent nor independent from differences in compilers.
Additionally, since the metrics collected rely heavily on Pin instrumentation, characterization of multi-threaded workloads or accelerators are not supported.
As such, it is unsuited to conventional supercomputing workloads which make heavy use of parallelism and accelerators.

Shao and Brooks [@shao2013isa] have since extended the generality of the MICA to be ISA independent.
The primary motivation for this work was in evaluating the suitability of benchmark suites when targeted on general purpose accelerator platforms.
The proposed framework briefly evaluates eleven SPEC benchmarks and examines 5 ISA-independent features/metrics.
Namely, number of opcodes (e.g., add, mul), the value of branch entropy -- a measure of the randomness of branch behaviour, the value of memory entropy -- a metric based on the lack of memory locality when examining accesses, the unique number of static instructions, and the unique number of data addresses.

Related to the paper, Shao also presents a proof of concept implementation (WIICA) which uses an LLVM IR Trace Profiler to generate an execution trace, from which a python script collects the ISA independent metrics.
Any results gleaned from WIICA are easily reproducible, the execution trace is generated by manually selecting regions of code built from the LLVM IR Trace Profiler.
Unfortunately, use of the tool is non-trivial given the complexity of the toolchain and the nature of dependencies (LLVM 3.4 and Clang 3.4).
Additionally, WIICA operates on `C` and `C++` code, which cannot be executed directly on any accelerator device aside from the CPU.
Our work extends this implementation to the broader OpenCL setting to collect architecture independent metrics from a hardware-agnostic language -- OpenCL.
Additional metrics, such as Instructions To Barrier (ITB), Vectorization (SIMD) indicators and Instructions Per Operand (SIMT) were also determined and added by us in order to perform a similar analysis for concurrent and accelerator workloads.

The branch entropy measure used by Shao and Brooks [@shao2013isa] was initially proposed by Yokota [@yokota2007introducing] and uses Shannon's information entropy to determine a score of Branch History Entropy.
De Pestel, Eyerman and Eeckhout [@depestel2017linear] proposed an alternative metric, average linear branch entropy metric, to allow accurate prediction of miss rates across a range of branch predictors.
As their metric is more suitable for architecture-independent studies, we adopt it for this work.

Caparrós Cabezas and Stanley-Marbell [@CaparrosCabezas:2011:PDM:1989493.1989506] present a framework for characterizing instruction- and thread-level parallelism, thread parallelism, and data movement, based on cross-compilation to a MIPS-IV simulator of an ideal machine with perfect caches and branch prediction and unlimited functional units.
Instruction-level and thread-level parallelism are identified through analysis of data dependencies between instructions and basic blocks.
The current version of AIWC does not perform dependency analysis for characterizing parallelism, however, we hope to include such metrics in future versions.

In contrast to our multidimensional workload characterization, models such as Roofline [@williams2009roofline] and Execution-Cache-Memory [@hager2013exploring] seek to characterize an application based on one or two limiting factors such as memory bandwidth.
The advantage of these approaches is the simplicity of analysis and interpretation.
We view these models as capturing a 'principal component' of a more complex performance space; we claim that by allowing the capture of additional dimensions, AIWC supports performance prediction for a greater range of applications.

#Metrics

For each OpenCL kernel invocation, the Oclgrind simulator **AIWC** tool collects a set of metrics, which are listed in [Table @tbl:aiwc-metrics].

\iftoggle{ACM-BUILD}{
\begin{table*}[t]
\caption{Metrics collected by the \textbf{AIWC} tool ordered by type. \label{tbl:aiwc-metrics}}
}{}
\iftoggle{IEEE-BUILD}{
\begin{table*}[t]
\caption{Metrics collected by the \textbf{AIWC} tool ordered by type. \label{tbl:aiwc-metrics}}
}{}
\iftoggle{LNCS-BUILD}{
\begin{table}[tb]
\caption{Metrics collected by the \textbf{AIWC} tool ordered by type. \label{tbl:aiwc-metrics}}
\begin{adjustbox}{max width=\textwidth}
}{}

\centering

\begin{tabular}{@{}cll@{}}
\toprule

{Type} & {Metric} & {Description}\\\hline

Compute & Opcode & total \# of unique opcodes required to cover 90\% of dynamic
instructions\\
Compute & Total Instruction Count & total \# of instructions executed\\
Parallelism & Work-items & total \# of work-items or threads executed\\
Parallelism & Total Barriers Hit & total \# of barrier instructions\\
Parallelism & Min ITB & minimum \# of instructions executed until a barrier\\
Parallelism & Max ITB & maximum \# of instructions executed until a barrier\\
Parallelism & Median ITB & median \# of instructions executed until a barrier\\
Parallelism & Min IPT & minimum \# of instructions executed per thread\\
Parallelism & Max IPT & maximum \# of instructions executed per thread\\
Parallelism & Median IPT & median \# of instructions executed per thread\\
Parallelism & Max SIMD Width & maximum \# of data items operated on during an instruction\\
Parallelism & Mean SIMD Width & mean \# of data items operated on during an instruction\\
Parallelism & SD SIMD Width & standard deviation across \# of data items affected\\
Memory & Total Memory Footprint & total \# of unique memory addresses accessed\\
Memory & 90\% Memory Footprint & \# of unique memory addresses that cover 90\% of memory accesses\\
Memory & Unique Reads & total \# of unique memory addresses read\\
Memory & Unique Writes & total \# of unique memory addresses written\\
Memory & Unique Read/Write Ratio & indication of workload being (unique reads / unique writes) \\
Memory & Total Reads & total \# of memory addresses read\\
Memory & Total Writes & total \# of memory addresses written\\
Memory & Reread Ratio & indication of memory reuse for reads (unique reads/total reads)\\
Memory & Rewrite Ratio & indication of memory reuse for writes (unique writes/total writes)\\
Memory & Global Memory Address Entropy & measure of the randomness of memory addresses\\
Memory & Local Memory Address Entropy & measure of the spatial locality of memory addresses\\
Control & Total Unique Branch Instructions & total \# of unique branch instructions\\
Control & 90\% Branch Instructions & \# of unique branch instructions that cover 90\%
of branch instructions\\
Control & Yokota Branch Entropy & branch history entropy using Shannon's information entropy\\
Control & Average Linear Branch Entropy & branch history entropy score using the
average linear branch entropy\\
\hline
\end{tabular}

\iftoggle{ACM-BUILD}{
\end{table*}
}{}
\iftoggle{IEEE-BUILD}{
\end{table*}
}{}
\iftoggle{LNCS-BUILD}{
\end{adjustbox}
\end{table}
}{}

The **Opcode**, **total memory footprint** and **90% memory footprint** measures are simple counts.
Likewise, **total instruction count** is the number of instructions achieved during a kernel execution.
The **global memory address entropy** is a positive real number that corresponds to the randomness of memory addresses accessed.
The **local memory address entropy** is computed as 10 separate values according to increasing number of Least Significant Bits (LSB), or low order bits, omitted in the calculation.
The number of bits skipped ranges from 1 to 10, and a steeper drop in entropy with increasing number of bits indicates greater spatial locality in the address stream.

Both **unique branch instructions** and the associated **90% branch instructions** are counts indicating the count of logical control flow branches encountered during kernel execution.
**Yokota branch entropy** ranges between 0 and 1, and offers an indication of a program's predictability as a floating point entropy value. [@yokota2007introducing] 
The **average linear branch entropy** metric is proportional to the miss rate in program execution; $p=0$ for branches always taken or not-taken but $p=0.5$ for the most unpredictable control flow.
All branch-prediction metrics were computed using a fixed history of 16-element branch strings, each of which is composed of 1-bit branch results (taken/not-taken).

As the OpenCL programming model is targeted at parallel architectures, any workload characterization must consider exploitable parallelism and associated communication and synchronization costs.
We characterize thread-level parallelism (TLP) by the number of **work-items** executed by each kernel, which indicates the maximum number of threads that can be executed concurrently.

Work-item communication hinders TLP, and in the OpenCL setting, takes the form of either local communication (within a work-group) using local synchronization (barriers) or globally by dividing the kernel and invoking the smaller kernels on the command queue.
Both local and global synchronization can be measured in **instructions to barrier** (ITB) by performing a running tally of instructions executed per work-item until a barrier is encountered under which the count is saved and resets; this count will naturally include the final (implicit) barrier at the end of the kernel.
**Min**, **max** and **median ITB** are reported to understand synchronization overheads, as a large difference between min and max ITB may indicate an irregular workload.

**Instructions per thread** (IPT) based metrics are generated by performing a running tally of instructions executed per work-item until completion.
The count is saved and resets.
**Min**, **max** and **median IPT** are reported to understand load imbalance.

To characterize data parallelism, we examine the number and width of vector operands in the generated LLVM IR, reported as **max SIMD width**, **mean SIMD width** and standard deviation -- **SD SIMD width**.
Further characterisation of parallelism is presented in the **work-items** and **total barriers hit** metrics.

Some of the other metrics are highly dependent on workload scale, so **work-items** may be used to normalize between different scales.
For example, **total memory footprint** can be divided by **work-items** to give the total memory footprint per work-item, which indicates the memory required per processing element.

Finally, unique verses absolute reads and writes can indicate shared and local memory reuse between work-items within a work-group, and globally, which shows the predictability of a workload.
To present these characteristics the **unique reads**, **unique writes**, **unique read/write ratio**, **total reads**, **total writes**, **reread ratio**, **rewrite ratio** metrics are proposed.
The **unique read/write ratio** shows that the workload is balanced, read intensive or write intensive.
They are computed by storing read and write memory accesses separately and are later combined, to compute the **global memory address entropy** and **local memory address entropy** scores.

#Methodology -- Workload Characterization by tooling Oclgrind

AIWC verifies the architecture independent metrics since they are collected on a toolchain and in a language actively executed on a wide range of accelerators -- the OpenCL runtime supports execution on CPU, GPU, DSP, FPGA, MIC and ASIC hardware architectures.
The intermediate representation of the OpenCL kernel code is a subset of LLVM IR known as SPIR -- Standard Portable Intermediate Representation.
This IR forms a basis for Oclgrind to perform OpenCL device simulation, which interprets LLVM IR instructions.

Migrating the metrics presented in the ISA-independent workload characterization paper [@shao2013isa] to the Oclgrind tool offers an accessible, high-accuracy and reproducible method to acquire these AIWC features.
Namely:

* Accessibility: since the Oclgrind OpenCL kernel debugging tool is one of the most adopted OpenCL debugging tools freely available to date, having AIWC metric generation included as an Oclgrind plugin allows rapid workload characterization.
* High-Accuracy: evaluating the low level optimized IR does not suffer from a loss of precision since each instruction is instrumented during its execution in the simulator, unlike with the conventional metrics generated by measuring architecture driven events -- such as PAPI and MICA analysis.
* Reproducibility: each instruction is instrumented by the AIWC tool during execution, there is no variance in the metric results presented between OpenCL kernel runs.

The caveat with this approach is the overhead imposed by executing full solution HPC codes on a slower simulator device.
However, since AIWC metrics do not vary between runs, this is still a shorter execution time than the typical number of iterations required to get a reasonable statistical sample when compared to a MICA or architecture dependent analysis.

<!--
Oclgrind OpenCL kernel debugging tool is one of the most adopted OpenCL debugging tools freely available to date, and increases the accessibility of workload characterization -- by having these features as an instrumentation plugin.
-->

#Implementation

AIWC is implemented as a plugin for Oclgrind, which simulates kernel execution on an ideal compute device.
OpenCL kernels are executed in series, and Oclgrind generates notification events which AIWC handles to populate data structures for each workload metric.
Once each kernel has completed execution, AIWC performs statistical summaries of the collected metrics by examining these data structures.

The **Opcode** diversity metric updates a counter on an unordered map during each \texttt{workItemBegin} event, the type of operation is determined by examining the opcode name using the LLVM Instruction API.

The number of **work-items** is computed by incrementing a global counter -- accessible by all work-item threads -- once a \texttt{workItemBegin} notification event occurs.

TLP metrics require barrier events to be instrumented within each thread.
Instructions To Barrier **ITB** metrics require each thread to increment a local counter once every \texttt{instructionExecuted} has occurred, this counter is added to a vector and reset once the work-item encounters a barrier.
The **Total Barriers Hit** counter also increments on the same condition.
Work-items are executed sequentially within all work-items in a work-group.
If a barrier is hit the queue moves onto all other available work-items in a ready state.
Collection of the metrics post barrier resumes during the \texttt{workItemClearBarrier} event.

ILP **SIMD** metrics examine the size of the result variable provided from the \texttt{instructionExecuted} notification, the width is then added to a vector for the statistics to be computed once the kernel execution has completed.

**Total Memory Footprint** **90% Memory Footprint** and Local Memory Address Entropy **LMAE** metrics require the address accessed to be stored during kernel execution and occurs during the \texttt{memoryLoad}, \texttt{memoryStore}, \texttt{memoryAtomicLoad} and \texttt{memoryAtomicStore} notifications.

Branch entropy measurements require a check during \texttt{instructionExecuted} event on whether the instruction is a branch instruction, if so a flag indicating a branch operation has occurred is set and both LLVM IR labels -- which correspond to branch targets -- are recorded.
On the next \texttt{instructionExecuted} the flag is queried and reset while the current instruction label is compared against which of the two targets were taken, the result is stored in the branch history trace.
The implementation of this is shown in Listing \ref{lst:instructionExecuted}.
Note the `instructionExecuted` callback is propagated from Oclgrind during every OpenCL kernel instruction -- emulated in LLVM IR.
This function also updates variables to track instruction diversity by counting the occurrences of each instruction, instructions to barrier and other parallelism metrics by running a counter until a barrier is hit, finally, the vectorization -- as part of the parallelism metrics -- are updated by recording the width of executed instructions.
The `m_state` variable is shared between all work-items in a work-group and these are stored into a global set of variables using a mutex lock once the work-group has completed execution.

The branch metrics are then computed by evaluating the full history of combined branch's taken and not-taken.

\begin{lstlisting}[float=*t,language=C++, caption={The Instruction Executed callback function collects specific program metrics and adds them to a history trace for later analysis.},label={lst:instructionExecuted}]
void WorkloadCharacterisation::instructionExecuted(...,  const llvm::Instruction *instruction, ...){
    unsigned opcode = instruction->getOpcode();
    std::string opcode_name = llvm::Instruction::getOpcodeName(opcode);
    //update key-value pair of instruction name and its occurrence in the kernel
    (*m_state.computeOps)[opcode_name]++;
    std::string Str = "";
    //if a conditional branch which has labels, store the labels to track
    //in the next instruction which of the two lines we end up in
    if (opcode == llvm::Instruction::Br && instruction->getNumOperands() == 3){
        if(instruction->getOperand(1)->getType()->isLabelTy() &&
                instruction->getOperand(2)->getType()->isLabelTy()){
            m_state.previous_instruction_is_branch = true;
            llvm::raw_string_ostream OS(Str);
            instruction->getOperand(1)->printAsOperand(OS,false);
            m_state.target1 = Str;
            Str = "";
            instruction->getOperand(2)->printAsOperand(OS,false);
            m_state.target2 = Str;
            llvm::DebugLoc loc = instruction->getDebugLoc();
            m_state.branch_loc = loc.getLine();
         }
    }
    //if the last instruction was a branch, log which of the two targets were taken
    else if (m_state.previous_instruction_is_branch == true){
        llvm::raw_string_ostream OS(Str);
        instruction->getParent()->printAsOperand(OS,false);
        if(Str == m_state.target1)
            (*m_state.branchOps)[m_state.branch_loc].push_back(true);//taken
        else if(Str == m_state.target2){
            (*m_state.branchOps)[m_state.branch_loc].push_back(false);//not taken
        }
        m_state.previous_instruction_is_branch = false;
    }
    //counter for instructions to barrier and other parallelism metrics
    m_state.instruction_count++;
    m_state.workitem_instruction_count++;
    //SIMD instruction width metrics use the following
    m_state.instructionWidth->push_back(result.num);
\end{lstlisting}

The **Total Unique Branch Instructions** is a count of the absolute number of unique locations that branching occurred, while the **90% Branch Instructions** indicates the number of unique branch locations that cover 90% of all branches.
**Yokota** from Shao [@shao2013isa], and **Average Linear Branch Entropy**, from De Pestel [@depestel2017linear] and have been computed and are also presented based on this implementation.
\texttt{workGroupComplete} events trigger the collection of the intermediate work-item and work-group counter variables to be added to the global suite, while \texttt{workGroupBegin} events reset all the local/intermediate counters.

Finally, \texttt{kernelBegin} initializes the global counters and \texttt{kernelEnd} triggers the generation and presentation of all the statistics listed in [Table @tbl:aiwc-metrics].
The source code is available at the GitHub Repository [@beau_johnston_2017_1134175].

#Demonstration

```{r, load_libraries, include=FALSE, cache=FALSE, echo=FALSE}
library(gridGraphics)
library(cowplot)
source('../analysis_tools/kiviat_plotter.R')

```

```{r, load_aiwc_data, include=FALSE, cache=FALSE, echo=FALSE}

load('../analysis_tools/newfeat.Rdf')

#source('../analysis_tools/load_aiwc_dataframes.R')
source('../analysis_tools/restructure_old_aiwc_data.R')

featdata.all <- drop_metrics_for_simple_kiviat(featdata.all)

featdata.kmeans <- subset(featdata.all,application=="kmeans")
featdata.lud    <- subset(featdata.all,application=="lud")
featdata.csr    <- subset(featdata.all,application=="csr")
featdata.fft    <- subset(featdata.all,application=="fft")
featdata.gem    <- subset(featdata.all,application=="gem")
featdata.dwt    <- subset(featdata.all,application=="dwt")
featdata.srad   <- subset(featdata.all,application=="srad")
featdata.bfs    <- subset(featdata.all,application=="bfs")
featdata.hmm    <- subset(featdata.all,application=="hmm")
featdata.nw     <- subset(featdata.all,application=="nw")
featdata.crc    <- subset(featdata.all,application=="crc")

source('../analysis_tools/stacked_features.R')

```

```{r eval=TRUE, dev='pdf', echo=FALSE, cache=FALSE,include=FALSE, generate_lud_perimeter_all_kiviat}
# TODO: load new aiwc results with load imbalance as newfeatdata.lud
source('../analysis_tools/restructure_aiwc_data.R')
rm(featdata.lud)
source('../analysis_tools/load_aiwc_lud_updated_dataframes.R')
newfeatdata.lud <- featdata.lud
newfeatdata.lud <- drop_metrics_for_simple_kiviat(newfeatdata.lud)
x <- reorder_and_subset(featdata=newfeatdata.lud,size='tiny',kernel="lud_perimeter")
x <- data.frame(t(colMeans(x)))
y <- x
x <- reorder_and_subset(featdata=newfeatdata.lud,size='small',kernel="lud_perimeter")
x <- data.frame(t(colMeans(x)))
y <- rbind(y,x)
x <- reorder_and_subset(featdata=newfeatdata.lud,size='medium',kernel="lud_perimeter")
x <- data.frame(t(colMeans(x)))
y <- rbind(y,x)
x <- reorder_and_subset(featdata=newfeatdata.lud,size='large',kernel="lud_perimeter")
x <- data.frame(t(colMeans(x)))
y <- rbind(y,x)
#TODO: newfeatdata.lud should be featdata.all for normalization vector -- instead this just removes the uncomputed load_imbalance results, since only the lud application has been updated with new awic metrics to measure load imbalance
featdata.all <- subset(featdata.all, application!="lud")
featdata.all$metric <- as.character(featdata.all$metric)
featdata.all$metric[featdata.all$metric == "max instructions to barrier"] <- "max instructions executed by a work-item"
featdata.all$metric[featdata.all$metric == "min instructions to barrier"] <- "min instructions executed by a work-item"
featdata.all$metric <- as.factor(featdata.all$metric)
featdata.all <- rbind(featdata.all,featdata.lud)
nv <- new_reorder_features(featdata.all)
nv$load_imbalance[which(nv$application!="lud")] <- 0
nv <- subset(nv,select=-c(size,kernel,application,invocation))
plot_kiviat(y,nv,colour="black",colour_spokes=TRUE,all_sizes=TRUE)
z1 <- recordPlot()

x <- reorder_and_subset(featdata=newfeatdata.lud,size='tiny',kernel="lud_internal")
x <- data.frame(t(colMeans(x)))
y <- x
x <- reorder_and_subset(featdata=newfeatdata.lud,size='small',kernel="lud_internal")
x <- data.frame(t(colMeans(x)))
y <- rbind(y,x)
x <- reorder_and_subset(featdata=newfeatdata.lud,size='medium',kernel="lud_internal")
x <- data.frame(t(colMeans(x)))
y <- rbind(y,x)
x <- reorder_and_subset(featdata=newfeatdata.lud,size='large',kernel="lud_internal")
x <- data.frame(t(colMeans(x)))
y <- rbind(y,x)

#pdf(file='../figure/tmp.pdf')
plot_kiviat(y,nv,colour="black",colour_spokes=TRUE,all_sizes=TRUE)
z2 <- recordPlot()
#plot_kiviat(y,nv,colour="black",colour_spokes=TRUE,all_sizes=TRUE)
#z2 <- recordPlot()
#dev.off()
#z2 <- image_read('../figure/tmp.pdf')
#z2 <- ggdraw() + draw_image(image_trim(z2))

x <- reorder_and_subset(featdata=newfeatdata.lud,size='tiny',kernel="lud_diagonal")
x <- data.frame(t(colMeans(x)))
y <- x
x <- reorder_and_subset(featdata=newfeatdata.lud,size='small',kernel="lud_diagonal")
x <- data.frame(t(colMeans(x)))
y <- rbind(y,x)
x <- reorder_and_subset(featdata=newfeatdata.lud,size='medium',kernel="lud_diagonal")
x <- data.frame(t(colMeans(x)))
y <- rbind(y,x)
x <- reorder_and_subset(featdata=newfeatdata.lud,size='large',kernel="lud_diagonal")
x <- data.frame(t(colMeans(x)))
y <- rbind(y,x)

plot_kiviat(y,nv,colour="black",colour_spokes=TRUE,all_sizes=TRUE)
z3 <- recordPlot()

source('../analysis_tools/generate_simple_lud_lines.R')

```

We now demonstrate the use of **AIWC** on several scientific application kernels selected from the Extended OpenDwarfs Benchmark Suite [@johnston18opendwarfs].
These benchmarks were extracted from and are representative of general scientific application codes.
Our selection is not intended to be exhaustive, rather, it is meant to illustrate how key properties of the codes are reflected in the metrics collected by **AIWC**.

AIWC is run on full application codes, but it is difficult to present an entire summary due to the nature of OpenCL.
Computationally intensive kernels are simply selected regions of the full application codes and are invoked separately for device execution.
As such, the AIWC metrics can either be shown per kernel run on a device, or as the summation of all metrics for a kernel for a full application at a given problem size -- we chose the latter.
Additionally, given the number of kernels presented we believe AIWC will generalize to full codes in other domains.

We present metrics for `r length(unique(featdata.all$application))` different application codes -- which includes `r length(unique(featdata.all$kernel))` kernels in total.
Each code was run with four different problem sizes, called **tiny**, **small**, **medium** and **large** in the Extended OpenDwarfs Benchmark Suite; these correspond respectively to problems that would fit in the L1, L2 and L3 cache or main memory of a typical current-generation CPU architecture.
As simulation within Oclgrind is deterministic, all results presented are for a single run for each combination of code and problem size.

In a cursory breakdown, four selected metrics are presented in [Figure @fig:stacked_plots].
One metric was chosen from each of the main categories, namely, Opcode, Barriers Per Instruction, Global Memory Address Entropy, Branch Entropy (Linear Average).
Each category has also been segmented by colour: blue results represent *compute* metrics, green represent metrics that indicate *parallelism*, yellow represents *memory* metrics and purple bars represent *control* metrics.
Median results are presented for each metric -- while there is no variation between invocations of AIWC, certain kernels are iterated multiple times and over differing domains/data sets.
Each of the 4 sub-figures shows all kernels over the 4 different problem sizes.

For almost all benchmarks the global memory address entropy increases with problem size, whereas the other metrics do not increase.
Notably, memory entropy is low for \texttt{lud\_diagonal}, reflecting memory access with constant strides of diagonal matrix elements, and \texttt{cl\_fdt53Kernel}, again reflecting regular strides generated by downsampling in the discrete wavelet transform.
Note, we do not present **medium** and **large** problem sizes for some kernels due to various issues including: a lack of input datasets, failure of AIWC in tracing large numbers of memory and branch operations for entropy calculations.
These issues will be addressed in future work.

Looking at branch entropy, \texttt{bfs\_kernel2} stands out as having by far the greatest entropy.
This kernel is dominated by a single branch instruction based on a flag value which is entirely unpredictable, and could be expected to perform poorly on a SIMT architecture such as a GPU.

Barriers per instruction is quite low for most kernels, with the exception of \texttt{needle\_opencl\_shared\_1} and \texttt{needle\_opencl\_shared\_2} from the Needleman-Wunsch DNA sequence alignment dynamic programming benchmark.
These kernels each have 0.04 barriers per instruction (i.e. one barrier per 25 instructions), as they follow a highly-synchronized wavefront pattern through the matrix representing matching pairs.
The performance of this kernel on a particular architecture could be expected to be highly dependent on the cost of synchronization.

```{r, build_stacked_plots, include=FALSE, cache=TRUE, echo=FALSE}
source('../analysis_tools/stacked_features.R')
```

```{r eval=TRUE, dev='pdf', cache=TRUE, echo=FALSE, include=FALSE, draw_stacked_plots, fig.cap = "\\label{fig:stacked_plots} Selected AIWC metrics from each category over all kernels and 4 problem sizes.", fig.width=8.3, fig.height=11.7}

legend_generic <- get_legend(stacked_tiny + theme(legend.title=element_text(face="bold",size=10),
                                                  legend.text=element_text(size=10),
                                                  legend.position="bottom",
                                                  legend.justification="right",
                                                  legend.direction="horizontal"))
plots <- align_plots(stacked_tiny  ,
                     stacked_small ,
                     stacked_medium,
                     stacked_large)
                     #align="hv",axis="tblr")

xoff <- .22 # relative x position of label, within one plot
yoff <- .98 # relative y position of label, within one plot

x <- plot_grid(plot_grid(plots[[1]],plots[[2]],ncol=2,align="h")+draw_plot_label(label=c("Tiny",  "Small"),
                                                                                 x=(xoff+0:1)/2,
                                                                                 y=rep(1-(1-yoff),2),
                                                                                 hjust=.5, vjust=.5,
                                                                                 size=15),
               plot_grid(plots[[3]],plots[[4]],ncol=2,align="h")+draw_plot_label(label=c("       Medium",  "Large"),
                                                                                 x=(xoff+0:1)/2,
                                                                                 y=rep(1-(1-yoff),2),
                                                                                 hjust=.5, vjust=.5,
                                                                                 size=15),
               legend_generic,
               rel_heights=c(1,1,.05),
               nrow=3)
print(x)
```

\begin{figure*}
\centering
\iftoggle{ACM-BUILD}{
%acm
\includegraphics[width=2\columnwidth]{./figure/draw_stacked_plots-1.pdf}
}{}
\iftoggle{IEEE-BUILD}{
%ieee
\includegraphics[width=1.98\columnwidth]{./figure/draw_stacked_plots-1.pdf}
}{}
\iftoggle{LNCS-BUILD}{
%llncs
\includegraphics[width=0.95\textwidth,height=0.95\textheight,keepaspectratio]{./figure/draw_stacked_plots-1.pdf}
}{}
\caption{Selected AIWC metrics from each category over all kernels and 4 problem sizes.}
\label{fig:stacked_plots} 
\end{figure*}

```{r eval=TRUE, dev='pdf', echo=FALSE, include=FALSE, cache=TRUE, draw_lud_diagonal_internal_all_kiviat, fig.cap = "\\label{fig:lud_diagonal_internal_kiviat} feature-space of the \\texttt{diagonal} and \\texttt{internal} kernels in the lud application on all problem sizes.", fig.width=8,fig.height=16}
print(plot_grid(z2,z3,labels=c('A) LUD Diagonal','B) LUD Internal'),ncol=1))#labels = "AUTO"
```

```{r eval=TRUE, dev='pdf', cache=TRUE, echo=FALSE, include=FALSE, draw_lud_diagonal_perimeter_lmae_all_kiviat, fig.cap = "\\label{fig:lud_perimeter_kiviat_and_lmae} feature-space of the \\texttt{perimeter} kernel in the lud application on all problem sizes and Local Memory Address Entropy over the tiny problem size.", fig.width=8,fig.height=16}
print(plot_grid(z1,df_plot,labels=c('       A) LUD Perimeter','B) LUD Perimeter LMAE on tiny'),ncol=1))#labels = "AUTO"
```

\begin{figure*}
    \centering
    \newcommand{\plotwidth}{0.66\textwidth}
    \includegraphics[width=\plotwidth]{./figure/draw_lud_diagonal_internal_all_kiviat-1.pdf}
    \caption{A) and B) show the AIWC features of the \texttt{diagonal} and \texttt{internal} kernels of the LUD application over all problem sizes.}
    \label{fig:kiviat}
\end{figure*}

\begin{figure*}
    \centering
    \newcommand{\plotwidth}{0.66\textwidth}
    \includegraphics[width=\plotwidth]{./figure/draw_lud_diagonal_perimeter_lmae_all_kiviat-1.pdf}
    \caption{A) shows the AIWC features of the \texttt{perimeter} kernel of the LUD application over all problem sizes. B) shows the corresponding Local Memory Address Entropy for the \texttt{perimeter} kernel over the tiny problem size.}
    \label{fig:kiviat2}
\end{figure*}

##Detailed Analysis of LU Decomposition Benchmark

We now proceed with a more detailed investigation of one of the benchmarks, **lud**, which performs decomposition of a matrix into upper and lower triangular matrices.
Following Shao and Brooks [@shao2013isa], we present the AIWC metrics for a kernel as a Kiviat or radar diagram, for each of the problem sizes.
Unlike Shao and Brooks, we do not perform any dimensionality reduction but choose to present all collected metrics.
The ordering of the individual spokes is not chosen to reflect any statistical relationship between the metrics, however, they have been grouped into four main categories: green spokes represent metrics that indicate *parallelism*, blue spokes represent *compute* metrics, beige spokes represent *memory* metrics and purple spokes represent *control* metrics.
For clarity of visualization, we do not present the raw AIWC metrics but instead, normalize or invert the metrics to produce a scale from 0 to 1.
The parallelism metrics presented are the inverse values of the metrics collected by AIWC, i.e. **granularity** =  $1 / \textbf{work-items}$ ; **barriers per instruction** $= 1 / \textbf{mean ITB}$ ; **instructions per operand** $= 1 / \sum \textbf{SIMD widths}$.

Additionally, a common problem in parallel applications is load imbalance -- or the overhead introduced by unequal work distribution among threads.
A simple measure to quantify imbalance can be achieved using a subset of the existing AIWC metrics and is included as a further derived parallelism metric by computing **load imbalance** = **max IPT** $-$ **min IPT**.

All other values are normalized according to the maximum value measured across all kernels examined -- and on all problem sizes.
This presentation allows a quick value judgement between kernels, as values closer to the centre (0) generally have lower hardware requirements, for example, smaller entropy scores indicate more regular memory access or branch patterns, requiring less cache or branch predictor hardware; smaller granularity indicates higher exploitable parallelism; smaller barriers per instruction indicates less synchronization; and so on.

The **lud** benchmark application comprises three major kernels, **diagonal**, **internal** and **perimeter**, corresponding to updates on different parts of the matrix.
The AIWC metrics for each of these kernels are presented -- superimposed over all problem sizes -- in [Figure @fig:kiviat] A) B) and [Figure @fig:kiviat2] A) respectively.
Comparing the kernels, it is apparent that the diagonal and perimeter kernels have a large number of branch instructions with high branch entropy, whereas the internal kernel has few branch instructions and low entropy.
This is borne out through inspection of the OpenCL source code: the internal kernel is a single loop with fixed bounds, whereas diagonal and perimeter kernels contain doubly-nested loops over triangular bounds and branches which depend on thread id.
Comparing between problem sizes (moving across the page), the large problem size shows higher values than the tiny problem size for all of the memory metrics, with little change in any of the values.

The visual representation provided from the Kiviat diagrams allows the characteristics of OpenCL kernels to be readily assessed and compared.

Finally, we examine the linear memory access entropy (LMAE) presented in the Kiviat diagrams in greater detail.
[Figure @fig:kiviat2] B) presents a sample of the linear memory access entropy, in this instance of the LUD Perimeter kernel collected over the tiny problem size.
The kernel is launched 4 separate times during a run of the tiny problem size, this is application specific and in this instance, each successive invocation operates on a smaller data set per iteration.
Note there is a steady decrease in starting entropy, and each successive invocation of the LU Decomposition Perimeter kernel the lowers the starting entropy.
However, the descent in entropy -- which corresponds to more bits being skipped, or bigger the strides or the more localized the memory access -- shows that the memory access patterns are the same regardless of actual problem size.
In general, for cache-sensitive workloads -- such as LU-Decomposition -- a steeper descent between increasing LMAE distances indicates more localized memory accesses, and this corresponds to better cache utilisation when these applications are run on physical OpenCL devices.
It is unsurprising that applications with a smaller working memory footprint would exhibit more cache reuse with highly predictable memory access patterns.

<!-- OpenCL Performance Prediction using Architecture-Independent Features HPCS-DRSN -->
Recently, AIWC has been used for predictive modelling on a set of 15 compute devices including CPUs, GPUs and MIC.
The AIWC metrics generated from the full set of Extended OpenDwarfs kernels were used as input variables in a regression model to predict kernel execution time on each device. [@johnston18predicting]
The model predictions differed from the measured experimental results by an average of 1.1%, which corresponds to actual execution time mispredictions of 8 $\mu$s to 1 second according to problem size.
From the accuracy of these predictions, we can conclude that while our choice of AIWC metrics is not necessarily optimal, they are sufficient to characterize the behaviour of OpenCL kernel codes and identify the optimal execution device for a particular kernel.

#Conclusions and Future Work

We have presented the Architecture-Independent Workload Characterization tool (AIWC), which supports the collection of architecture-independent features of OpenCL application kernels.
It is the first workload characterization tool to support multi-threaded or parallel workloads.
These features can be used to predict the most suitable device for a particular kernel, or to determine the limiting factors for performance on a particular device, allowing OpenCL developers to try alternative implementations of a program for the available accelerators -- for instance, by reorganizing branches, eliminating intermediate variables et cetera.
In addition, the architecture independent characteristics of a scientific workload will inform designers and integrators of HPC systems, who must ensure that compute architectures are suitable for the intended workloads.

Caparrós Cabezas and Stanley-Marbell [@CaparrosCabezas:2011:PDM:1989493.1989506] examine the Berkeley dwarf taxonomy by measuring instruction-level parallelism (ILP), thread-level parallelism (TLP), and data movement.
They propose a sophisticated metric to assess ILP by examining the data dependency graph of the instruction stream.
Similarly, TLP was measured by analysing the block dependency graph.
While we propose alternative metrics to evaluate ILP and TLP -- using the max, mean and standard deviation statistics of SIMD and barrier metrics respectively -- a quantitative evaluation of the dwarf taxonomy using these metrics is left as future work.
<!-- killing dwarfs -->
We expect that the AIWC metrics will generate a comprehensive feature-space representation which will permit cluster analysis and comparison with the dwarf taxonomy.

We believe AIWC will also be useful in guiding device-specific optimization by providing feedback on how particular optimizations change performance-critical characteristics.
To identify which AIWC characteristics are the best indicators of opportunities for optimization, we are currently looking at how individual characteristics change for a particular code through the application of best-practice optimizations for CPUs and GPUs (as recommended in vendor optimization guides).

A major limitation of running large applications under AIWC is the high memory footprint.
Memory access entropy scores require a full recorded trace of every memory access during a kernel's execution.
However, a graceful degradation in performance is preferable to an abrupt crash in AIWC if virtual memory is exhausted.
For this reason, work is currently being undertaken for an optional build of AIWC with low memory usage by writing these traces to disk.
<!--If Oclgrind is compiled with levelDB -- a lightweight open-source database library for persistence -- the memory-light version of AIWC is used.-->

<!--\todo[inline]{
However the memory light work is still under-development but is very close to being finished. It can be completed in time for the camera-ready.
\todo[inline]{*Josh* had an idea that AIWC could be used to show cache sensitive vs cache oblivious workloads -- this could be mentioned in future work}

\todo[inline]{several typos and missing commas -- run grammarly}
\todo[inline]{discuss the considerations of how these metrics were chosen differ from Shao, what do we add to the theory about instrumenting parallel workloads?}
\todo[inline]{include a jupyter artefact}

\todo[inline]{Open review questions still to address:}
\todo[inline]{*Josh* Where should we address reviewer 1s question: How would you generalize the normalization of the metrics with the max over all measured applications? What would this be for a larger (or unknown) set of applications?}
\todo[inline]{motivate why these metrics and not others are relevant}
\todo[inline]{why the selected configuration parameters are chosen and not others.}
\todo[inline]{Especially given that the measured metrics show a significant change with respect to problem size and likely also with the parallelism involved (number of processing elements), what insights can be gained from their analysis?}
\todo[inline]{Is any beahvior simulated generalizable or must each separate configuration be analyzed on its own?}
\todo[inline]{A further concern is whether this methodology only works on small kernels or on real applications as well.}
\todo[inline]{Would the increased complexity from a real HPC application preclude, inhibit, or otherwise affect the use of AIWC? A case study with a real application would help me alleviate these concerns.}
\todo[inline]{It is not clear to this reviewer how such claims can be made about kernels based primarily on branching metrics, as most processors, including GPUs, can perform well even when code contains branching (this may have been different in the past). In this reviewer's experience, rarely is branching the determining factor as to whether an application is suitable for an accelerator, so I feel the paper needs to attempt to convince the audience otherwise.}
\todo[inline]{This paper could be greatly improved if there was more extensive work on use cases for the tool, for instance discussing the metrics of larger or more diverse kernels, or those notoriously challenging for manual analysis. This would also need to be backed up with more quantitative and objective analysis of the metrics.}
\todo[inline]{This approach is of course a suggestion and there might be other ways to present the work, but as it stands it appears to lack novel contribution beyond the implementation. Implementing a tool to collect metrics is useful; however, proving the tool can provide insights for modern HPC applications would make solid research suitable for the conference audience.}
-->

#References

### 2020-01-14 paper/paper.tex
paper addition

\documentclass[review=false, sigchi]{acmart}

\usepackage{listings}
\usepackage{color}
\usepackage{pgfplots}
\usepackage{subfig}

\raggedbottom
\acmConference[IWOCL'20]{International Workshop on OpenCL}{April 27--29, 2020}{Munich, Germany}
%\editor{John B. Appleseed}
\title[Architecture Independent Memory Access Pattern Analysis]{Characterising Optimisations to Memory Access Patterns using Architecture Independent Program Features}

\author{Aditya Chilukuri}
\email{aditya.chilukuri@anu.edu.au}
\affiliation{%
	\institution{Australian National University}}

\author{Josh Milthorpe}
\email{josh.milthorpe@anu.edu.au}
\affiliation{%
	\institution{Australian National University}}

\author{Beau Johnston}
\email{beau.johnston@anu.edu.au}
\affiliation{%
	\institution{Australian National University}}

%\acmVolume{10} %TODO
%\acmNumber{10} %TODO
%\acmArticle{10} %TODO
\acmYear{2020}
\acmMonth{6}
\acmPrice{15.00}

\keywords{Architecture Independent Analysis, Heterogenous Computing, Workload Characterisation, Memory Access Patterns} %TODO

\definecolor{dkgreen}{rgb}{0,0.6,0}
\definecolor{gray}{rgb}{0.5,0.5,0.5}
\definecolor{mauve}{rgb}{0.58,0,0.82}

\lstset{frame=tb,
	language=C,
	aboveskip=3mm,
	belowskip=3mm,
	showstringspaces=false,
	columns=flexible,
	basicstyle={\small\ttfamily},
	numbers=none,
	numberstyle=\tiny\color{gray},
	keywordstyle=\color{blue},
	commentstyle=\color{dkgreen},
	stringstyle=\color{mauve},
	breaklines=true,
	breakatwhitespace=true,
	tabsize=3,
	numbers=left
}

\begin{document}

	\begin{abstract}
	This paper addresses the challenge of profiling a code base to help programmers effectively optimise their codes for high performance.
	Current methods of analysing program behaviour have largely relied on performance counters on various architectures to provide architecture dependent metrics of code behaviour. 
	However, if workload characteristics are tied to architectural features specific to a particular system, these analysis frameworks may not generalise to alternative or future systems.
	Considering the trend towards heterogeneous computing, we work towards a framework to guide developers in optimising codes for any given hardware target using the architecture independent workload characterisation (AIWC) tool -- specifically in profiling memory access patterns of code. 
	Metrics must be easily interpreted and conceptually linked to patterns in hardware-specific memory access optimisations.
	To this end we propose, implement and analyse the addition of two new metrics to AIWC and present a new approach to profile spatial locality intrinsic to parallel codes.
	The improvements suggested to AIWC will help HPC developers better understand memory access patterns of complex codes, a crucial step in guiding optimisation of codes for arbitrary hardware targets.
	\end{abstract}
	\maketitle

	\section{Introduction}

	High-performance computing (HPC) systems are increasingly incorporating a multitude of heterogeneous compute elements. Single nodes on supercomputers now comprise of traditional CPUs and multiple accelerators such as GPUs, Field Programmable Gate Arrays (FPGAs), and many-integrated-core devices (MICs). The development of high bandwidth interconnects between compute devices supports tight integration between multiple devices of different types on the same compute node.

	The OpenCL programming language is designed to support modern HPC software engineers in writing code that executes on multiple hardware targets. 
	%It is possible for an engineer to write highly thread-divergent logical analysis code aimed at traditional CPUs; parallelisable data crunching kernels aimed at GPUs; and code to satisfy specialised computational requirements aimed at FPGAs -- all in a single code base using OpenCL. 
	The language provides support for individual \emph{kernels}, portions of parallelisable codes that make up a HPC code base, to be executed independently on diverse hardware targets such as traditional CPUs, GPUs and FPGAs \cite{khronosopenclworkinggroup2011opencl}. This gives HPC software developers greater flexibility by allowing codes aimed at a range of hardware targets to be written in a single programming language environment.

	Application codes differ in resource requirements, control structure and available parallelism. Similarly, compute devices differ in number and capabilities of execution units, processing model, and available resources. These heterogeneous computing environments present opportunities for software engineers and HPC integrators to design highly optimised systems with multiple kernels executing on hardware targets best suited for the diverse computational tasks performed by these kernels \cite{spafford2010maestro}.	However this opportunity raises the twin challenges of deciding on the right hardware target for complex kernels; and optimising codes to run on these diverse architectures. Our work is directed at helping developers tackle the latter of these two problems.

	Modern computer architectures have suffered from the exponential growth in computational speeds being unmet by proportional increases in memory speeds\cite{hennessycomparch}. Memory accesses pose larger bottlenecks to performance as applications' demand for main memory scales with the arithmetic capability of computer systems. To mitigate this latency, modern CPU designs have employed a wide range of cache technologies to reduce main memory accesses using the principle of \emph{spatial locality}: the idea that programs repeatedly access memory addresses in close locality. On the other hand, GPUs rely on hardware multithreading to hide memory latency, and their architectures favour ALU capability over sophisticated logic to manage a cache hierarchy and out-of-order execution. As a result, the performance of kernels on CPUs and GPUs alike is strongly dependent on memory access patterns intrinsic to the code. % We discuss the memory model design and the manufacturer recommended optimisations for NVIDIA GPUs in later sections..

	%We demonstrate the effectiveness of an architecture-independent analysis of codes in determining ... Our work guides optimisation efforts for OpenCL codes running on increasingly diverse hardware by examining the current state of the art in architecture-independent metrics for memory access patterns.

	%What we did briefly
	Our aim in this paper is to develop a framework to guide HPC software engineers in hardware dependent code optimisation -- specifically by guiding the improvement of codes' memory access patterns. We provide examples of manufacturer recommended code optimisations to improve memory access patterns on the target architecture, and examine these using AIWC \cite{beauaiwc}, a state of the art architecture-independent code analysis tool for parallel workloads. Our work highlights the benefits and challenges arising from an architecture independent analysis of memory based program characteristics. We suggest the addition of two new metrics to AIWC to better characterise these memory-based optimisations. We measure the presented codes using the new metrics and discuss the metrics' effectiveness in capturing the essence of the optimisations performed. With these improvements, we show that AIWC is better suited to be used by HPC software engineers in guiding optimisation of their codes.

	The structure of this paper is as follows. In section \ref{motivating example}, we use the example of GPU matrix multiplication code to showcase the specific vendor recommended optimisations our work aims to measure. In section \ref{related work} we discuss how AIWC and its precursors profiled memory access patterns in an architecture independent fashion, and relevant architecture dependent approaches to memory access profiling. Then in section \ref{method} we provide a methodology to critically examine AIWC, the current state of the art architecture independent workload characterisation tool. In section \ref{results} we measure the accuracy of AIWC in profiling the impact of performance optimisations to various OpenCL codes. In section \ref{discussion} we suggest new methods to accurately capture code patterns correlating to hardware specific memory optimisations. Finally, in section \ref{future work} we discuss further avenues to extend our work and conclude.

	\section{Motivating Example} \label{motivating example}

	We aim to capture the essence of vendor recommended optimisations for target architectures in our metrics. We demonstrate the optimisation strategies being analysed by presenting a simple unoptimised matrix multiplication kernel for input square matrices of order $N$, and then improve the memory access patterns of this kernel by incrementally performing optimisations recommended in the CUDA Optimisation Handbook \cite{cudaoptimisation} for NVIDIA GPUs. These incrementally optimised OpenCL codes are then analysed using AIWC to determine the accuracy in profiling favourable memory access patterns of the current framework and the proposed extensions to AIWC.

	\subsection{Simple unoptimised matrix multiplication}

	The unoptimised matrix multiplication kernel presented in Appendix \ref{simpleMultiply} is used as a baseline to validate the performance improvements measured from each following optimisation. Each thread of the kernel computes the $(globalRow, globalCol)$ element of matrix $C$ by reading the appropriate row and column from matrices $A$ and $B$ respectively.

	\subsection{Using Shared Memory to Coalesce Global Memory Access to Matrix A}

	We first notice that the number of global memory accesses to matrices A and B increases at an order $O(N^3)$ with respect to matrix size $N$. Global memory is typically located off-chip and accesses induce large delays. NVIDIA GPUs coalesce global memory loads and stores issued within thread-groups into as few DRAM transactions as possible. Multiple global memory loads and stores are coalesced into a single transaction when certain device specific conditions are met. On most NVIDIA GPUs, data accesses are coalesced when multiple requests are made for memory locations from the same cache-line in global memory \cite{cudamanual}. Appendix \ref{coalescedAMultiply} contains the \texttt{coalescedAMultiply} kernel, which coalesces accesses to matrix $A$ by storing \emph{tiles}, or square blocks, of $A$'s values into shared memory.

	% 9.2.1.2: Misaligned access pattern

	\subsection{Using Shared Memory to store Tiles of Matrix B}

	The code is further optimised by improving how the previous \texttt{coalescedAMultiply} kernel handles accesses to Matrix $B$. While each thread in coalescedAMultiply accesses only a single element of matrix $A$ from global memory, each thread accesses a full column of matrix $B$. The repeated reads to elements of matrix $B$ across multiple threads can be alleviated by reading tiles from matrix $B$ into shared memory. Appendix \ref{coalescedABMultiply} contains the \texttt{coalescedAB} kernel which performs this optimisation.

	\subsection{Optimising handling of shared memory}

	NVIDIA shared memory is divided into multiple banks -- stored in independent memory modules -- to allow parallel accesses to shared memory. Bank conflicts occur when shared memory in the same bank is accessed concurrently. The code in \texttt{coalescedAB} kernel is further optimised by implicitly transposing tiles of Matrix $A$ while loading from global memory. This improves memory bank utilisation during reads to shared memory tiles are performed during the necessary computation. The \texttt{coalescedABT} kernel demonstrates this optimisation by modifying lines 20,21 of \texttt{coalescedAB} to:

	\begin{lstlisting}[firstnumber=20]
		ASub[localCol][localRow] = A[tiledRow];
		BSub[localRow][localCol] = B[tiledCol];
	\end{lstlisting}

	\subsection{Alignment of Memory Allocation}

	Memory access alignment is important to best utilise all parts of GPU memory architecture. Global memory buffer alignment can allow threads to access blocks of global memory aligned to the nearest cache line. This enables coalescing of memory accesses. If buffers are misaligned, parallel memory requests may cross over cache lines, and may double the number of slow global memory accesses needed as demonstrated in figures \ref{fig: Simple memory alignment} and \ref{fig: Simple memory misalignment}.

	\begin{figure}[h]
		\includegraphics[width=\linewidth]{aligned_access_coalescing.pdf}
		\caption{All threads access memory aligned to nearest cache line in parallel \cite{cudaoptimisation}}
		\label{fig: Simple memory alignment}
	\end{figure}

	\begin{figure}[h]
		\includegraphics[width=\linewidth]{misaligned_access_coalescing.pdf}
		\caption{Unaligned sequential memory addresses fit in two cache lines \cite{cudaoptimisation}}
		\label{fig: Simple memory misalignment}
	\end{figure}

	A similar principle applies when using shared memory. Bank conflicts may be reduced by aligning allocations of shared memory buffers. The \texttt{alignedABT} kernel improves the alignment of shared memory tiles in \texttt{coalescedABT}. An arbitrary large alignment value of 4096 is chosen as it is larger than the size of modern hardwares' cache lines and attributed to the local array declarations in the existing code as shown below.

	\begin{lstlisting}
		__local float aTile[TILE_DIM][TILE_DIM] __attribute__ ((aligned(4096)));
		__local float bTile[TILE_DIM][TILE_DIM] __attribute__ ((aligned(4096)));
	\end{lstlisting}

	All the examples above are an incomplete listing of optimisation strategies developers can perform to code targeted at NVIDIA GPUs. We analyse the correlations between the collected AIWC metrics for each of these kernels and our conceptual understanding of GPU memory optimisations. This allows us to examine how effectively our methodology uncovers underlying bottlenecks and guides optimisation efforts.

	\flushbottom
	\section{Related Work} \label{related work}

	Hoste and Eeckout \cite{hoste2007microarchitecture} show that while conventional architecture dependent characteristics are useful in locating performance bottlenecks, they can hide the underlying, inherent program behaviour causing performance bottlenecks. A conceptual understanding of the performance characteristics of complex codes is necessary for the programmer to effectively optimise these codes. Architecture dependent characteristics typically include instructions per cycle (IPC), cache and branch prediction miss rates, page faults and DRAM bus data transfer rates. These are typically collected using hardware performance counters available on most target architectures. These performance counters do not serve to guide optimisation beyond highlighting potential bottlenecks \cite{hoste2007microarchitecture, ganesan2008performance}. Further, in many cases, architecture dependent characteristics usually cannot be directly correlated to specific code patterns. For example, the causes of high cache miss rates in the execution of a program are too complex and widely variant on microarchitecture specific features such as cache size, prefetch behaviour and cache placement policies. In summary, we argue that a HPC developer tasked with optimising code for a given hardware target would benefit from architecture-independent metrics of the code that can be conceptually correlated to hardware specific optimisation patterns. Such architecture-independent metrics will effectively guide the HPC developer's workflow in both finding and fixing bottlenecks in code performance.

	The Architecture Independent Workload Characterisation (AIWC) tool \cite{beauaiwc} relies on a set of instruction set architecture (ISA)-independent features determined by Shao and Brooks \cite{shao2013isa}. AIWC measures memory based metrics to characterise the target code's memory behaviours. AIWC firstly collects aggregate simple counts such as the (i) total memory footprint, the total number of unique addresses accessed; and (ii) 90\% memory footprint, the number of unique addresses covering 90\% of memory accesses. A small ratio of 90\% memory footprint to total memory footprint correlates to programs accessing the same small subset of memory addresses repeatedly -- which can be linked to increased performance in a cached memory hierarchy.

	AIWC also records the global memory address entropy (GMAE), a positive real number corresponding to the randomness of the memory access distribution of a program. To measure locality of memory accesses, AIWC collects the local memory address entropy (LMAE) of memory addresses accessed after dropping $n$ least significant bits of all memory addresses accessed by the program. To calculate this, AIWC collects a frequency distribution of all non-register memory accesses by all threads in the target kernel. Using the collected frequency distribution, AIWC calculates 10 separate local memory address entropy (LMAE) values according to increasing number of least significant bits (LSB) skipped using
	%in the following fashion. The information entropy of the collection of memory addresses accessed, after dropping $n$ LSBs and accounting for the number of memory accesses performed to each address is counted as the $n$-bits skipped LMAE. Given the frequency distribution of memory addresses accessed by the program, 
	the explicit formula for the $n$-bits skipped LMAE:

	\begin{equation}
	LMAE_{n-bits} = \sum_{a \in A_n}p_a \log_2(p_a^{-1})
	\end{equation}

	\begin{itemize}
		\item $A_n$ is the set of all addresses accessed after skipping $n$ LSBs of each address.
		\item $p_a := \frac{\#access_a}{\#access_{total}}$ is the probability (calculated as relative frequency) at which each memory address is accessed.
	\end{itemize}

	LMAE measures the locality of memory accesses performed over the full execution of a program. A steeper drop in entropy with increasing number of bits may correlate to more localised memory accesses over the program's execution.

	Limitations of the memory footprint and local memory address entropy values as memory access pattern metrics are further discussed by this work. %TODO: REMOVE?

	\subsection{Advancements in Memory Access Pattern Characterisation}

	We provide a brief summary of past work in architecture independent program characterisation and Memory Access Pattern analysis:

	\begin{itemize}
		\item The Microarchitecture Independent Workload Characterisation Tool (MICA) \cite{hoste2007microarchitecture} was the first microarchitecture independent workload characterisation tool for single threaded programs. 
		\item The Workload ISA-Independent Characterization for Applications (WIICA) \cite{shao2013isa} extends MICA to present a framework to analyse single threaded programs independent of the ISA.
		\item CuMAPz: CuMAPz \cite{kim2011cumapz} is a CUDA memory access profiling tool and the first to present metrics for CUDA codes which can be generalised to any given NVIDIA GPU.
	\end{itemize}

	\subsection{MICA}

	The MICA framework was developed by Hoste and Eeckout \cite{hoste2007microarchitecture} in response to their findings on the limitations of conventional microarchitecture dependent characteristics. Hoste and Eeckout discovered that performance counter based approaches to profiling codes often struggled in finding underlying program features that map to improved or worsened usage of performance critical hardware features of the target architecture. The MICA framework is a holistic characterisation tool, and thus collects features including instruction mix, instruction-level parallelism, register traffic, data stream strides and branch predictability.

	Of these metrics, data stream strides are of particular interest in memory access pattern profiling. MICA's stride length metric measures the distance between consecutive memory accesses in a single threaded application. For CPU architectures running single threaded applications, this metric correlates to the spatial locality of memory accesses -- a measure of how closely bunched are memory access in nearby times. This is directly correlated to cache reuse rates, critical to code performance on CPUs \cite{inteloptimisation}. %Intel optimisation manual.

	The MICA approach was tailored for single-threaded applications as the metrics collected rely heavily on Pin instrumentation \cite{luk2005pin}. As such, MICA is unsuited to analysing conventional supercomputing workloads with heavy use of parallelism.

	\subsection{CuMAPz}

	Kim and Shrivastava \cite{kim2011cumapz} present CuMAPz, a CUDA memory access pattern analysis tool to guide NVIDIA GPU optimisations. CuMAPz focuses on the problem of improving CUDA application performance using NVIDIA memory-based optimisations. CuMAPz analyses CUDA codes structurally and simulates code execution on the memory hierarchy of specific NVIDIA GPU models.

	During the simulated run of the target code, CuMAPz records all memory accesses in various buffers to simulate the global and shared memories on NVIDIA GPUs. Using this detailed simulation data, CuMAPz can estimate the performance critical (i) shared memory data reuse profit, (ii) profit from coalesced access (iii) memory channel skew cost and (iv) bank conflict cost characteristics of the target code.

	The simulation environment described in CuMAPz and the attached analysis framework is highly specific to CUDA enabled GPUs. Replicating the CuMAPz framework for all target architectures is challenging. However, CuMAPz is an interesting simulator from the standpoint of this paper's work since it adheres to the core parts of NVIDIA GPUs' memory models in its analysis, while allowing the user to specify their GPU model specific hardware information.

	Our approach is an architecture independent analysis of memory access patterns that provides metrics correlating to similar performance critical memory access optimisations as CuMAPz. We aim to further the state of the art by providing a framework to guide developers in optimising OpenCL codes for any given target architecture. To the best our knowledge, none of the previous works present a set of performance metrics that accurately characterise memory access patterns of parallel applications independent of the target architecture.

	\section{Methodology} \label{method}

	To provide memory access pattern metrics that represent vendor recommended optimisations on GPUs, we improved AIWC by adding a new metric: relative local memory usage. This measures the proportion of all memory accesses from the symbolic execution of the kernel that occurred to memory allocated as \texttt{\_\_local}. On GPUs, this memory address space is mapped to fast on-chip shared memory.

	Relative local memory usage is an example of a metric that is useful to measure performance critical access patterns on some architectures such as GPUs, and not others, such as CPUs. CPUs do not typically have a notion of user controlled on-chip memory shared between hardware threads such as NVIDIA GPUs' shared memory. This is a natural consequence of programming for a heterogeneous system. Specific code patterns may translate to performance improvements only on certain hardware.

	In the original AIWC tool \cite{beau_johnston_2017_1134175}, global and local memory address entropy (MAE) were calculated using physical addresses of memory used by the Oclgrind simulator back-end of AIWC. This caused irregularities in entropy calculations across multiple runs of the same simulation. We improved the calculation of memory address entropy by using \textit{virtual addresses} to calculate MAE values using an abstract ideal address space on which all memory accesses by the kernel occur. This allows AIWC to accurately abstract over the hardware and ISA-specific differences in memory layouts across the diverse hardware targets.

	\section{Results} \label{results}

	\begin{figure*}[h!]
		\centering
		\subfloat{
			\centering
			\begin{tikzpicture}
			\begin{axis}[symbolic x coords={simple,coalA,coalAB,coalABT,alignABT},
			xtick={simple,coalA,coalAB,coalABT,alignABT},
			xticklabel style={text height=2ex},
			ymin=0,
			ylabel={Execution time ($\mu s$)}, xlabel={Kernel},
			ylabel style={at={(axis description cs:0.1,.5)},anchor=south},
			legend entries={$N=80\ \ \ \ $,$N=256\ \ $},
			legend style={at={(0.6,0.45)},anchor=west}
			]

			\addplot[mark=*] coordinates {
				(simple,27.79)
				(coalA,16.36)
				(coalAB,17.02)
				(coalABT,16.34)
				(alignABT,30.21)
			};
			\addplot[mark=star] coordinates {
				(simple,319.71)
				(coalA,70.12)
				(coalAB,73.60)
				(coalABT,65.27)
				(alignABT,61.50)
			};	
			\end{axis}
			\end{tikzpicture} 
		}
		\subfloat{
			\begin{tikzpicture}
			\begin{axis}[symbolic x coords={simple,coalA,coalAB,coalABT,alignABT},
			xtick={simple,coalA,coalAB,coalABT,alignABT},
			xticklabel style={text height=2ex},
			ymin=0,
			ylabel={Execution time ($\mu s$)}, xlabel={Kernel},
			ylabel style={at={(axis description cs:0.1,.5)},anchor=south},
			legend entries={$N=1408$,$N=2048$},
			legend style={at={(0.6,0.45)},anchor=west}
			]

			\addplot[mark=triangle*] coordinates {
				(simple,46901.35)
				(coalA,8948.96)
				(coalAB,8894.74)
				(coalABT,7756.75)
				(alignABT,6951.7773435)
			};
			\addplot[mark=square] coordinates {
				(simple,130601.41)
				(coalA,28722.51)
				(coalAB,27169.72)
				(coalABT,23284.03)
				(alignABT,21363.90)
			};	
			\end{axis}
			\end{tikzpicture} 
		}
		\caption{NVIDIA Tesla P100 Runtime performance of section \ref{motivating example} codes averaged over 100 runs}
		\label{fig: matmul runtime}
	\end{figure*}

	Figure \ref{fig: matmul runtime} shows that across varying problem sizes, the vendor recommended optimisations to the matrix multiplication code lead to increased performance. Note that a matrix size of 80 is small -- overheads attached to optimisation techniques may be causing the observed decrease in performance as some optimisations are performed.

	\begin{table*}[h!]
	\centering
	\begin{tabular}[t]{lrrrrr}
		\toprule
		& simple & coalescedA & coalescedAB & coalescedABT & alignedABT \\ \midrule 
		Total memory footprint        & 196608 & 196608     & 196608      & 196608       & 196608     \\ %\hline
		90\% Memory Footprint         & 118196 & 56176      & 489         & 489          & 489        \\ %\hline
		Global MAE                    & 17.02  & 13.18      & 9.78        & 9.78         & 9.78       \\ %\hline
		LMAE \#bits=3                 & 16.02  & 12.18      & 8.78        & 8.78         & 8.78       \\ %\hline
		LMAE \#bits=10                & 9.02   & 5.18       & 1.78        & 1.78         & 1.78       \\ %\hline
		Relative Local Memory Usage   & 0      & 49.9083    & 94.28       & 94.28        & 94.28      \\ \bottomrule
	\end{tabular}
	\caption{Selection of AIWC\cite{beauaiwc} metrics for 256 $\times$ 256 matrix multiplication}
	\label{matmul aiwc table}
	\end{table*}

	Table \ref{matmul aiwc table} summarises AIWC memory metrics collected for each of the matrix multiply kernels in section \ref{motivating example}. Note that only two LMAE values are shown for brevity. We now analyse the effectiveness of AIWC metrics in profiling the NVIDIA recommended memory optimisations applied to the matrix multiplication kernel.

	\textit{Relative local memory usage}: As reliance on NVIDIA GPUs' shared memory increases in each kernel from \texttt{simpleMultiply} to \texttt{coalescedAB}, we find that the ratio of OpenCL local memory usage increases as expected.

	\textit{Global and local MAE}: Entropy measurements decrease from \texttt{simpleMultiply} to \texttt{coalescedAB}. This is explained by the optimisations performed incrementally reducing accesses to matrix $A$ and $B$, and instead performing the bulk of computations on smaller tiles in on-chip shared memory. We observe an almost completely uniform distribution of memory accesses in \texttt{simpleMultiply}, where the program makes $N$ loads to each element of $A$ and $B$ with $N$ the dimension of the matrices. The frequency distribution of memory addresses accessed become increasingly non-uniform as we perform less accesses to matrices $A$ and $B$ and offset this by increasing the frequency of accesses to smaller, more localised shared memory buffers. The non-uniformity of relative frequency with which each memory cell is accessed by the program translates directly to decreases in local and global memory entropy values as reliance on small shared memory buffers increases from \texttt{simpleMultiply} to \texttt{coalescedAB}.

	\textit{Memory Footprint}: Similar to trends in global and local MAE, we find that the ratio of 90\% memory footprint to total memory footprint decreases from $60.12\%$ for \texttt{simpleMultiply} to $0.25\%$ (\texttt{coalescedAB}). Increased utilisation of shared memory in the optimised kernels means that the shared memory buffers make up a greater proportion of total memory accesses in the program. As the shared memory buffers are small and reused within a workgroup, the memory footprint of the shared memory buffers is also smaller.

	AIWC's metrics strongly reward optimisations that tend to localise memory accesses. Shared memory buffers are typically smaller than global memory arrays when programming for GPUs due to hardware limitations on sizes of shared memory buffers \cite{cudamanual}. However, the metrics currently measured by AIWC do not have a direct causal relation to code patterns that optimise memory accesses on GPUs. The proposed relative local memory usage metric is the first to correspond to a recommended optimisation strategy of using fast on-chip shared memory. Further, we find that all current AIWC metrics do not measure any sizeable difference between \texttt{coalescedAB}, \texttt{coalescedABT} and \texttt{alignedABT} codes. We address the root cause of this -- a lack of temporal information about memory accesses, in the next section.%TODO READ THIS SECTION

	\section{Discussion} \label{discussion}

	AIWC metrics are inherently lossy. Given the set of metrics calculated for an unknown kernel, we can not reconstruct the kernel based on these metrics. Specifically to memory access pattern analysis, the calculation of memory address entropy of a kernel relies only on the frequency distribution of memory accesses to all addresses accessed by the kernel. We observe that calculation of memory address entropy does not make use of temporal information. The order of sequential memory accesses performed by each thread, as well as addresses requested in parallel by multiple threads in an OpenCL thread group, are both vital in accurately characterising parallel codes.

	We propose a new architecture independent metric to measure memory access patterns of parallel programs. The proposed metric is inspired by CuMAPz' direct approaches to measuring optimisation specific characteristics of CUDA codes.

	AIWC currently collects a list of all memory accessed by each thread of execution. In the OpenCL programming model, threads within thread groups execute the same code and are allowed to cooperate on \textit{local memory} addresses. We can group together threads by their work groups to align histories of memory accesses of threads in a group to analyse memory accesses performed by thread groups at each logical time step in the symbolic execution of the code. We argue that analysing the accesses occuring at the same logical time step is an accurate, conceptually sound method of predicting the extent of memory access coalescing and bank conflicts we are likely to find upon executing the code on a GPU.

	The precise definition of the proposed metric is provided in the following section.

	\subsection{A Parallel Spatial Locality Metric}

		\begin{figure*}[h!]
		\centering
		\begin{tikzpicture}
		\begin{axis}[symbolic x coords={0,1,2,3,4,5,6,7,8,9,10},
		width=0.7\textwidth,
		height=0.5\textwidth,
		xtick={0,1,2,3,4,5,6,7,8,9,10},
		xticklabel style={text height=2ex},
		ymin=0,
		ylabel={$n$-bits dropped Parallel Spatial Locality}, xlabel={\# of bits skipped},
		ylabel style={at={(axis description cs:0.05,.5)},anchor=south},
		legend entries={\texttt{simpleMultiply},\texttt{coalescedAMultiply},
			\texttt{coalescedAB}, \texttt{coalescedABT}, \texttt{alignedABT}},
		legend style={at={(0.55,0.75)},anchor=west}
		]

		\addplot[mark=*] coordinates {
			(0, 4.01)
			(1,3.51)
			(2,3.01)
			(3,2.50)
			(4,2.00)
			(5,2.00)
			(6,2.00)
			(7,2.00)
			(8,2.00)
			(9,1.5)
			(10,1.00)
		};

		\addplot[mark=star] coordinates {
			(0,4.24)
			(1,3.71)
			(2,3.18)
			(3,2.65)
			(4,2.12)
			(5,1.62)
			(6,1.12)
			(7,0.62)
			(8,0.12)
			(9,0.09)
			(10,0.06)
		};

		\addplot[mark=triangle*] coordinates {
			(0,4.45)
			(1,3.89)
			(2,3.33)
			(3,2.78)
			(4,2.23)
			(5,1.73)
			(6,1.23)
			(7,0.73)
			(8,0.23)
			(9,0.17)
			(10,0.11)
		};

		\addplot[mark=square] coordinates {
			(0,4.45)
			(1,3.45)
			(2,2.45)
			(3,1.45)
			(4,0.45)
			(5,0.40)
			(6,0.34)
			(7,0.28)
			(8,0.23)
			(9,0.17)
			(10,0.11)
		};

		\addplot[mark=o] coordinates {
			(0,4.45)
			(1,3.45)
			(2,2.45)
			(3,1.45)
			(4,0.45)
			(5,0.40)
			(6,0.34)
			(7,0.28)
			(8,0.23)
			(9,0.17)
			(10,0.11)
		};
		\end{axis}
		\end{tikzpicture}
		\caption{Calculated parallel spatial locality metric for matrix multiply kernels for 256 $\times$ 256 matrix multiplication}
		\label{fig: matmul parallel spatial locality}
	\end{figure*}

	There are three steps involved in generating an AIWC metric: \textit{recording}, \textit{calculating} and \textit{summarising} data collected from the symbolic execution of the kernel under inspection.

	\textit{Record}: we first record memory accesses performed by each thread in an OpenCL thread group as described above to achieve a global ordering of all memory accesses performed in a thread group. This ordering is collected in the form of logical timestamps ($t_0 .. t_{last}$) at which memory accesses occur.

	\textit{Calculate}: for each timestamp $t = t_0 .. t_{last}$, calculate the n-bits dropped entropy of memory addresses accessed by all threads in a work group within the timestamp $t$. Here $n$ can range between 1 and 10 as was the case for $LMAE$.

	\textit{Summarise}: average the collected entropy values across all the timestamps to calculate the parallel spatial locality metrics for one thread group. We then average the n-bits dropped entropy summaries across thread groups to obtain the \textit{n-bits dropped parallel spatial locality metrics} for the kernel's execution.

%	\begin{equation} \label{entropyaverage}
%		\log_2(\frac{\sum_{t = t_0}^{t_{last}} 2^{E_{n,t}}}{\#{timestamps}})
%	\end{equation}

%	With $E_{n,t}$ the n-bits dropped entropy at timestamp $t$.

	The proposed metric is a parallel computing analogue for MICA's data stride metric that measures the distance between consecutive data accesses in a single threaded environment. In parallel programs, to accurately measure spatial locality of accesses, we must consider memory accesses performed by multiple threads in a close temporal scope. The proposed metric calculates the locality of accesses in each time step of the program's execution and steeper drop-offs in $n$-bits dropped parallel spatial locality scores will be observed in programs that access nearby memory addresses within the same timestamp. Such programs will perform better on GPUs, as they will make better use of both global memory access coalescing and shared memory bank structures.

	In figure \ref{fig: matmul parallel spatial locality} we see theoretical calculations for the proposed parallel spatial locality metric for each of the matrix multiplication kernels from section \ref{motivating example}. In particular we observe that the \texttt{coalescedABT} and \texttt{alignedABT} kernels have the steepest drop-offs in entropy as the number of bits skipped is increased, which correlates to better locality of parallel memory accesses. It is expected for these kernels to exhibit better parallel spatial locality than \texttt{simpleMultiply}, as \texttt{coalescedABT} and \texttt{alignedABT} make use of shared memory, where accesses tend to be localised simply due to the small size of shared on-chip memory typically available on GPUs. Further we find that the proposed metric successfully distinguishes between the \texttt{coalescedAB} and \texttt{coalescedABT} kernels. It accurately depicts a steeper drop-off for the more optimised \texttt{coalescedABT} kernel, where a larger proportion of parallel memory accesses make better use of the memory bank structure of GPU shared memory than all previous kernels. This is a significant improvement over the state-of-the-art AIWC metrics in characterising how codes localise shared memory accesses to better use the hardware provided.

	\section{Results}

	We present the results of running the AIWC tool on the Extended OpenDwarfs benchmark suite \cite{opendwarfs2017base,extendedopendwarfs}.

	\subsection{N-body methods: GEM}

	\subsection{Dense Linear Algebra: LUD}

	Lower-upper decomposition (LUD) in OpenDwarfs is a program to decompose an input square matrix as the product of one lower and one upper diagonal matrix. LUD is a Memory access patterns in a dense linear algebra workload such as LUD are typically highly regular and deterministic for each OpenCL work item, based on the the matrix dimension and offset parameters.

	The OpenDwarfs implementation \cite{opendwarfs2017base} splits the LUD computation into three kernels, ...  %TODO: Divergence in the kernels. Make a diagram/code simplification for LUD.

	\subsection{Structured Grids: SRAD}

	Speckle Reducing Anisotropic Diffusion (SRAD) attempts to remove locally correlated noise from images by following a repeated, structured grid update computation pattern. Conditional statements in the code cause thread divergence, potentially within a work-group, to handle boundary conditions. These boundaries constitute a small portion of the executed work-items, especially on larger data-sets and so the .

	\subsection{Graph Traversal: BFS}

	Graph traversal algorithms require pointer chasing operations to traverse nodes of a graph to perform calculations. The Breadth First Search

	\section{Conclusions and Future Work} \label{future work}

	To the best of our knowledge this work is the first to critically examine architecture independent methods to measure memory access patterns and guide hardware-specific optimisation. We implemented and justified the addition of relative local memory usage to AIWC in our steps to characterise memory-based GPU code optimisations. We have also proposed the methodology to implement a new parallel spatial locality metric to AIWC. Upon implementation, our work would be the first to measure memory access patterns of parallel codes in an architecture-independent, easily interpreted manner.
	%TODO UPDATE IN FUTURE

	Our proposed parallel spatial locality metric may also correlate to some memory-based optimisation strategies for CPUs. Future work would apply the approach followed by this paper to CPU and FPGA architectures to evaluate the viability of an architecture independent characterisation of memory access patterns in codes targeted to these architectures.

	Future work will provide a concrete implementation of this new parallel spatial locality metric in the AIWC framework. It is then feasible to test these new metrics across a larger set of GPU targeted OpenCL codes. This will allow a critical examination of the effectiveness of both proposed metrics in accurately profiling memory access patterns of these diverse codes.

	\bibliographystyle{plain}
	\bibliography{bib/bibliography.bib}

	\begin{appendix}

		\section{\texttt{simpleMultiply} kernel} \label{simpleMultiply}

		\begin{lstlisting}
		__kernel void simpleMultiply(__global float *A,
				__global float *B, 
				__global float *C, 
				int N)
		{
			const int globalRow = get_global_id(0); // Row ID of C (0..N-1)
			const int globalCol = get_global_id(1); // Col ID of C (0..N-1)
			// Compute a single element of C (loop over K)
			float acc = 0.0f;
			for (int k = 0; k < N; ++k) {
				acc += B[k * N + globalCol] * A[globalRow * N + k];
		`	}
			// Store the result
			C[globalRow * N + globalCol] = acc;
		}

		\end{lstlisting}

		\section{\texttt{coalescedAMultiply} kernel} \label{coalescedAMultiply}

		\begin{lstlisting}
		__kernel void coalescedAMultiply(
				const __global float* A,
				const __global float* B,
				__global float* C,
				const int N)
		{
			__local float aTile[TILE_DIM][TILE_DIM];

			const int localRow = get_local_id(0);
			const int localCol = get_local_id(1);

			const int globalRow = get_global_id(0);
			const int globalCol = get_global_id(1);

			__private float sum = 0.0f;

			const int numTiles = N / TILE_DIM;
			__private const int tiledRow = globalRow*N+ localCol;
			for (int i = 0; i < numTiles; i++) {
				aTile[localRow][localCol] = A[tiledRow+i*TILE_DIM];
				barrier(CLK_LOCAL_MEM_FENCE);
				for (int k = 0; k < TILE_DIM; k++) {
					sum += aTile[localRow][k] * B[(i*TILE_DIM+k)*N+globalCol];
				}
				barrier(CLK_LOCAL_MEM_FENCE);
			}
			C[globalRow*N+globalCol] = sum;
		}
		\end{lstlisting}

		\section{\texttt{coalescedAB} kernel} \label{coalescedABMultiply}

		\begin{lstlisting}
		__kernel void coalescedAB(
				const __global float* A,
				const __global float* B,
				__global float* C,
				const int N) {
			__local float ASub[TILE_DIM][TILE_DIM];
			__local float BSub[TILE_DIM][TILE_DIM];

			const int localRow = get_local_id(0);
			const int localCol = get_local_id(1);
			const int globalRow = get_global_id(0);
			const int globalCol = get_global_id(1);

			float acc = 0.0f;
			const int numTiles = N/TILE_DIM;

			for (int i = 0; i < numTiles; i++) {
				const int tiledRow = globalRow*N+i*TILE_DIM + localCol;
				const int tiledCol = globalCol + (TILE_DIM*i + localRow)*N;
				ASub[localRow][localCol] = A[tiledRow];
				BSub[localRow][localCol] = B[tiledCol];

				barrier(CLK_LOCAL_MEM_FENCE);

				for (int k=0; k<TILE_DIM; k++) {
					acc += ASub[localRow][k] * BSub[k][localCol];
				}

				barrier(CLK_LOCAL_MEM_FENCE);
			}
			C[globalRow*N + globalCol] = acc;
		}
		\end{lstlisting}
	\end{appendix}

\end{document}

### 2020-01-16 paper/paper.tex
pretty tables
don't need to show all significant digits for relative local memory use

@@
 }
+
+% tables (Josh) %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\usepackage{booktabs}
+\usepackage{relsize}
+\usepackage{multirow}
+\usepackage{array}
+\renewcommand{\arraystretch}{1.2}  % stretch space between table rows a bit
+
+\newcommand{\otoprule}{\midrule[\heavyrulewidth]}
+\definecolor{tableheadcolor}{rgb}{0.8,0.8,1.0}
+\definecolor{tablealtcolor}{rgb}{0.9,0.9,0.95}
+
+% bold table rows (use \rowstyle{\bffamily} on each bold row)
+\newcolumntype{B}{>{\global\let\currentrowstyle\relax}}
+\newcolumntype{^}{>{\currentrowstyle}}
+\newcommand{\rowstyle}[1]{\gdef\currentrowstyle{#1}#1\ignorespaces}
+\newcolumntype{C}{>{\bfseries}}
+
+% sans-serif tables
+\let\oldtabular\tabular
+\let\endoldtabular\endtabular
+\renewenvironment{tabular}{\sffamily\oldtabular}{\endoldtabular}
 
@@
 		LMAE \#bits=10                & 9.02   & 5.18       & 1.78        & 1.78         & 1.78       \\ %\hline
-		Relative Local Memory Usage   & 0      & 49.9083    & 94.28       & 94.28        & 94.28      \\ \bottomrule
+		Relative Local Memory Usage   & 0      & 49.91    & 94.28       & 94.28        & 94.28      \\ \bottomrule
 	\end{tabular}

### 2020-01-16 paper/paper.tex
Adding analysis of opendwarfs to paper

--- 2020-01-14 paper/paper.tex
@@
 
+	% As all kernels access memory on word (32-bit) boundaries, there is no effect in dropping upto two bits in the PSL metric...
+	\begin{figure*}[h!]
+		\centering
+		\begin{tikzpicture}
+		\begin{axis}[symbolic x coords={0,1,2,3,4,5,6,7,8,9,10},
+		width=0.7\textwidth,
+		height=0.5\textwidth,
+		xtick={0,1,2,3,4,5,6,7,8,9,10},
+		xticklabel style={text height=2ex},
+		ymin=0,
+		ylabel={$n$-bits dropped Parallel Spatial Locality}, xlabel={\# of bits skipped},
+		ylabel style={at={(axis description cs:0.05,.5)},anchor=south},
+		legend entries={\texttt{GEM}, \texttt{NW*}, \texttt{SRAD Kernel 1},\texttt{SRAD Kernel 2},
+			\texttt{CSR}, \texttt{BFS*}, \texttt{LUD Diagonal}, \texttt{LUD Internal}, \texttt{LUD Perimeter}},
+		legend style={at={(0.55,0.75)},anchor=west}
+		]
+
+		\addplot[mark=*] coordinates {
+			(0, 0.1139)
+			(1, 0.1139)
+			(2, 0.1139)
+			(3, 0.1139)
+			(4, 0.0989)
+			(5, 0.08402)
+			(6, 0.06914)
+			(7, 0.05431)
+			(8, 0.03938)
+			(9, 0.02442)
+			(10, 0.0124)
+		};
+
+		\addplot[mark=star] coordinates {
+			(0, 2.175)
+			(1, 2.175)
+			(2, 2.175)
+			(3, 2.175)
+			(4, 2.175)
+			(5, 2.175)
+			(6, 2.175)
+			(7, 2.175)
+			(8, 2.175)
+			(9, 2.175)
+			(10, 2.175)
+		};
+
+		\addplot[mark=*] coordinates {
+			(0, 7.232)
+			(1, 7.232)
+			(2, 7.232)
+			(3, 7.232)
+			(4, 6.351)
+			(5, 5.461)
+			(6, 4.570)
+			(7, 3.678)
+			(8, 3.648)
+			(9, 3.633)
+			(10, 3.625)
+		};
+
+		\addplot[mark=star] coordinates {
+			(0, 7.098)
+			(1, 7.098)
+			(2, 7.098)
+			(3, 7.098)
+			(4, 6.223)
+			(5, 5.344)
+			(6, 4.464)
+			(7, 3.583)
+			(8, 3.567)
+			(9, 3.559)
+			(10, 3.556)
+		};
+
+		\addplot[mark=triangle*] coordinates {
+			(0, 9.605)
+			(1, 9.605)
+			(2, 9.605)
+			(3, 9.605)
+			(4, 9.478)
+			(5, 9.293)
+			(6, 9.049)
+			(7, 8.761)
+			(8, 8.447)
+			(9, 8.12)
+			(10,7.233)
+		};
+
+		\addplot[mark=square] coordinates {
+			(0,4.406)
+			(1,4.406)
+			(2,4.406)
+			(3,4.406)
+			(4,4.361)
+			(5,4.273)
+			(6,4.135)
+			(7,3.941)
+			(8,3.689)
+			(9,3.382)
+			(10,3.029)
+		};
+
+		\addplot[mark=o] coordinates {
+			(0,1.548)
+			(1,1.548)
+			(2,1.548)
+			(3,1.548)
+			(4,1.305)
+			(5,1.092)
+			(6,0.9385)
+			(7,0.8796)
+			(8,0.8796)
+			(9,0.8796)
+			(10,0.8796)
+		};
+		\end{axis}
+		\end{tikzpicture}
+		\caption{Parallel spatial locality metric for selected OpenDwarfs benchmark kernels}
+		\label{fig: opendwarfs parallel spatial locality}
+	\end{figure*}
 	\subsection{N-body methods: GEM}
 
-	\subsection{Dense Linear Algebra: LUD}
+	% TODO
 
-	Lower-upper decomposition (LUD) in OpenDwarfs is a program to decompose an input square matrix as the product of one lower and one upper diagonal matrix.
+	\subsection{Dense Linear Algebra: Lower-upper decomposition (LUD)}
+
+	LUD in OpenDwarfs is a program to decompose an input $N\times N$ matrix as the product of one lower and one upper diagonal matrix.
 LUD is a Memory access patterns in a dense linear algebra workload such as LUD are typically highly regular and deterministic for each OpenCL work item, based on the the matrix dimension and offset parameters.
+The OpenDwarfs implementation of LUD \cite{opendwarfs2017base} splits the LUD computation into three kernels: LUD perimeter and diagonal kernels spawn work-items in a single work dimension while LUD internal performs work in both dimensions of the input matrix.
 
-	The OpenDwarfs implementation \cite{opendwarfs2017base} splits the LUD computation into three kernels, ...  %TODO: Divergence in the kernels. Make a diagram/code simplification for LUD.
+	LUD perimeter and diagonal partially benefit from memory access coalescing, for the lines of code where all the work-items access contiguous memory along a row of the input matrix.
+However, when threads parallelly access a column of the input matrix, multiple memory requests are made as addresses accessed are too distant to be coalesced into a single memory transaction on GPUs.
+This is reflected in the slow drop in parallel spatial locality for LUD perimeter and diagonal kernels.
 
-	\subsection{Structured Grids: SRAD}
+	LUD internal %TODO: 2Dimensional work -- better parallelism
 
-	Speckle Reducing Anisotropic Diffusion (SRAD) attempts to remove locally correlated noise from images by following a repeated, structured grid update computation pattern.
+	%The use of creates imbalanced workload and divergent control flow within work-items in each kernel iteration.
+While the presence of conditional statements cause some thread divergence in the LUD diagonal kernel, the use of barriers following conditional statements synchronises the threads.
+LUD diagonal and perimeter kernels , while LUD.
+
+	...  %TODO: Divergence in the kernels. Make a diagram/code simplification for LUD.
+
+	\subsection{Dynamic Programming: Needleman-Wunsch (NW)}
+
+	Needleman-Wunsch is a dynamic programming algorithm used to perform protein sequence alignment by identifying the similarity level between two strings of Amino acids.
+Elements in a matrix are computed, with each element dependent on its west, north and north-west neighbours.
+This dependency enforces a wavefront computation pattern, travelling along the main diagonal of the matrix.
+Each iteration of the kernels computes over an antidiagonal of the matrix, starting at its top left corner and finishing at its bottom right corner.
+
+	In particular, the wavefront pattern of computations causes each work-item to request distant memory addresses in a parallel fashion.
+Thus, parallel memory requests into the input matrix prohibit locality.
+We note the lack of any dropoff in parallel spatial locality as the number of bits dropped increases.
+This rightly indicates that memory addresses accessed at each logical timestamp are very distant.
+On GPUs this will translate to poor utilisation of both memory access coalescing and caching.
+This trend in the parallel spatial locality metric suggests that a possible improvement for GPU performance would be to load blocks of the input matrix into on-chip local memory to reduce the number of global memory requests.
+
+	\subsection{Structured Grids: Speckle Reducing Anisotropic Diffusion (SRAD)}
+
+	SRAD removes locally correlated noise from images by following a repeated grid update computation pattern.
 Conditional statements in the code cause thread divergence, potentially within a work-group, to handle boundary conditions.
-These boundaries constitute a small portion of the executed work-items, especially on larger data-sets and so the .
+These boundaries constitute a small portion of the executed work-items, especially on larger data-sets and so the effective thread divergence is minimal.
+Thus, memory access patterns in SRAD are statically determined and relatively localised.
+The observed rapid drop-off in parallel spatial locality of the SRAD kernels (figure \ref{fig: opendwarfs parallel spatial locality}) is due to the highly localised nature of parallel memory accesses.
+At each point in the kernels' memory access profile, each work-item $(i,j)$ in an OpenCL work-group accesses the $(i,j)^{th}$ elements of various matrices.
+The sequential memory access pattern is non-linear since different matrices are accessed consecutively by the kernel, prohibiting ideal caching.
+However, memory requests made parallelly are highly localised, allowing memory access coalescing on GPUs, greatly reducing the number of memory requests made on each line of the kernel code on GPUs.
+Thus while cache hit rates are typically low on this benchmark, particularly on GPUs \cite{krommydas2016opendwarfs}, GPUs can hide the latency of global memory accesses through coalescing.
 
@@
 
-	Graph traversal algorithms require pointer chasing operations to traverse nodes of a graph to perform calculations. The Breadth First Search
+	Graph traversal algorithms require pointer chasing operations to traverse nodes of a graph and perform calculations.
+The Breadth First Search implemented in OpenDwarfs calls OpenCL kernels to traverse nodes immediately connected to the list of nodes at each depth starting from the root node.
+BFS is characterised by an imbalanced workload per kernel launch depending on the degree of the nodes being operated on -- with only a proportion of launched nodes performing meaningful work.
+As such the AIWC features collected for each kernel invocation vary significantly.
+The parallel spatial locality metric collects the entropy of memory accesses at each time step, and workload imbalances are dealt with by averaging entropy scores across all the time steps in the execution of an OpenCL work-group.
+
+	The memory access patterns in BFS are irregular.
+Work-items fetch discontinuous memory locations attributed to any particular node in the graph, depending on the connectivity of the node they operate on.
+The program structure involves multiple levels of pointer-chasing and thus the precise memory address accessed are runtime dependent.
+This leads to poor parallel spatial locality.
+Thus, we see a slow dropoff in parallel spatial locality for BFS.
+
+	\subsection{Sparse Linear Algebra: CSR}
+
+	Compressed Sparse Row Matrix-Vector Multiplication (CSR) computes the sum of each of a matrix's row's elements after it is multiplied by a given vector.
+The matrix is stored in a compressed row storage sparse matrix format, which is very efficient for storage when the number of zero elements is far greater than the number of non-zero elements.
+Only non-zero elements of each row of a matrix are stored in $Ax$, along with separate arrays $Aj,\ Ap$ indicating the position of each non-zero element in the matrix.
+Figure \ref{?} depicts the sparse matrix format of storage for a $4\times4$ matrix.
+
+	In this particular implementation, a single OpenCL work-item is assigned to compute over each row of the input sparse matrix.
+Similar to BFS, memory access patterns are runtime-dependent, since the locations of matrix data read by each thread are dependent on the number and position of non-zero values in the sparse-matrix, decided by the values in $Aj$ and $Ap$.
+This pattern of indirect addressing is typical of applications in the Spare Linear Algebra Dwarf, which severely hinders locality of memory accesses performed.
+The parallel spatial locality metric reflects this trend.
+Figure \ref{fig: opendwarfs parallel spatial locality} shows the very gradual decline in parallel spatial locality as the number of bits dropped is increased.
+
+	\subsection{Finite State Machine: Temporal Data Mining (TDM)}
+
+	%TODO
 

### 2020-01-16 paper/paper.tex
propose new abstract

--- 2020-01-16 paper/paper.tex
@@
 	\begin{abstract}
+	proposed new abstract:
+
+	Context: AIWC gathers features of OpenCL programs which can be used to understand and predict program performance
+
+	Problem: metrics aren't easily interpreted and don't reflect important memory access patterns that affect efficiency across all architectures
+
+	Solution: new metrics
+
+	Results: differences in metrics reflect optimizations for matrix multiply; improved performance prediction by x\% for Extended OpenDwarfs benchmarks
+
+	Significance: as per last sentence\\
+
 	This paper addresses the challenge of profiling a code base to help programmers effectively optimise their codes for high performance.
@@
 
-	High-performance computing (HPC) systems are increasingly incorporating a multitude of heterogeneous compute elements.
-Single nodes on supercomputers now comprise of traditional CPUs and multiple accelerators such as GPUs, Field Programmable Gate Arrays (FPGAs), and many-integrated-core devices (MICs).
+	High-performance computing (HPC) systems are increasingly heterogeneous.
+	A single node on a modern supercomputers may combine a traditional CPUs with one or more accelerators such as GPUs, Field Programmable Gate Arrays (FPGAs), or many integrated core devices (MICs).
 	The development of high bandwidth interconnects between compute devices supports tight integration between multiple devices of different types on the same compute node.
@@
 Similarly, compute devices differ in number and capabilities of execution units, processing model, and available resources.
-These heterogeneous computing environments present opportunities for software engineers and HPC integrators to design highly optimised systems with multiple kernels executing on hardware targets best suited for the diverse computational tasks performed by these kernels \cite{spafford2010maestro}.	However this opportunity raises the twin challenges of deciding on the right hardware target for complex kernels; and optimising codes to run on these diverse architectures.
+These heterogeneous computing environments present opportunities for software engineers and HPC integrators to design highly optimised systems with multiple kernels executing on hardware targets best suited for the diverse computational tasks performed by these kernels \cite{spafford2010maestro}.
+	However this opportunity raises the twin challenges of determining the best hardware target for complex kernels and optimising codes to run on these diverse architectures.
 	Our work is directed at helping developers tackle the latter of these two problems.

### 2020-01-17 paper/paper.tex
add GEM analysis, working on abstract

@@
 	This paper addresses the challenge of profiling a code base to help programmers effectively optimise their codes for high performance.
+	The Architecture Independent Workload Characterisation Tool (AIWC) gathers features of OpenCL programs used to understand and predict program performance.
+
 	Current methods of analysing program behaviour have largely relied on performance counters on various architectures to provide architecture dependent metrics of code behaviour. 
@@
 	Considering the trend towards heterogeneous computing, we work towards a framework to guide developers in optimising codes for any given hardware target using the architecture independent workload characterisation (AIWC) tool -- specifically in profiling memory access patterns of code.
+
 	Metrics must be easily interpreted and conceptually linked to patterns in hardware-specific memory access optimisations.
@@
 		legend entries={\texttt{GEM}, \texttt{NW*}, \texttt{SRAD Kernel 1},\texttt{SRAD Kernel 2},
-			\texttt{CSR}, \texttt{BFS*}, \texttt{LUD Diagonal}, \texttt{LUD Internal}, \texttt{LUD Perimeter}},
+			\texttt{CSR}, \texttt{BFS*}, \texttt{LUD Diagonal}, \texttt{LUD Internal}, \texttt{LUD Perimeter}, \texttt{TDM}},
 		legend style={at={(0.55,0.75)},anchor=west}
@@
 	% TODO
-
+	The OpenDwarfs benchmark GEM computes the electrostatic potential of a biomolecule by calculating the sum of charges contributed by all
+	This is an embarrassingly parallel problem.
+Each OpenCL work-item operates on a single surface vertex, finding the electrostatic potential generated by looping through every atom in the biomolecule in a global order.
+	The computation pattern is highly regular and memory accesses are perfectly synchronised. Atom data is accessed consecutively, with all work-items simultaneously accessing each single atom's data.
+	The parallel spatial locality metric reflects this pattern of efficient memory utilisation (single loads from global memory servicing all OpenCL work-items).
+The recorded parallel spatial locality approaches the theoretical limit of 0 (0.0124 at 10 bits dropped).
+This indicates that almost all memory accesses made by the OpenCL kernel are perfectly synchronised between OpenCL threads.
+	%TODO: Clarify above sentence: I mean efficient memory fetches (one load from global memory services all threads) -- not reduced usage of memory.
 	\subsection{Dense Linear Algebra: Lower-upper decomposition (LUD)}

### 2020-01-18 paper/paper.tex
Updated abstract

@@
 
-	\begin{abstract}
+	\begin{comment}
 	proposed new abstract:
@@
 	Significance: as per last sentence\\
+	\end{comment}
 
-	This paper addresses the challenge of profiling a code base to help programmers effectively optimise their codes for high performance.
-	The Architecture Independent Workload Characterisation Tool (AIWC) gathers features of OpenCL programs used to understand and predict program performance.
+	\begin{abstract}
 
-	Current methods of analysing program behaviour have largely relied on performance counters on various architectures to provide architecture dependent metrics of code behaviour. 
-	However, if workload characteristics are tied to architectural features specific to a particular system, these analysis frameworks may not generalise to alternative or future systems.
-	Considering the trend towards heterogeneous computing, we work towards a framework to guide developers in optimising codes for any given hardware target using the architecture independent workload characterisation (AIWC) tool -- specifically in profiling memory access patterns of code.
-
-	Metrics must be easily interpreted and conceptually linked to patterns in hardware-specific memory access optimisations.
-	To this end we propose, implement and analyse the addition of two new metrics to AIWC and present a new approach to profile spatial locality intrinsic to parallel codes.
+	%This paper addresses the challenge of profiling a code base to help programmers effectively optimise their codes for high performance.
+	HPC developers are faced with the challenge of optimising OpenCL workloads for high performance on novel architectures.
+	The Architecture Independent Workload Characterisation (AIWC) tool is a plugin for the Oclgrind OpenCL simulator that gathers metrics of OpenCL programs used to understand and predict program performance on an arbitrary given hardware architecture.
+	However, AIWC metrics are not always easily interpreted and do not reflect some important memory access patterns affecting efficiency across architectures.
+	We propose the concept of parallel spatial locality -- the locality of memory accesses by OpenCL work-items (threads) executing simultaneously.
+	We implement the parallel spatial locality metric in the AIWC framework, and analyse gathered results on matrix multiply and the Extended OpenDwarfs OpenCL benchmarks.
+	The differences in the obtained parallel spatial locality metric across implementations of matrix multiply reflect the optimisations performed.
+	The new metric showcases the memory access patterns in the OpenDwarfs benchmarks that affect their performance on various architectures.
+	%Current methods of analysing program behaviour have largely relied on performance counters on various architectures to provide architecture dependent metrics of code behaviour. 
+	%However, if workload characteristics are tied to architectural features specific to a particular system, these analysis frameworks may not generalise to alternative or future systems.
+	%Considering the trend towards heterogeneous computing, we work towards a framework to guide developers in optimising codes for any given hardware target using the architecture independent workload characterisation (AIWC) tool -- specifically in profiling memory access patterns of code.
+	%Metrics must be easily interpreted and conceptually linked to patterns in hardware-specific memory access optimisations.
+	%To this end we propose, implement and analyse the addition of two new metrics to AIWC and present a new approach to profile spatial locality intrinsic to parallel codes.
 	The improvements suggested to AIWC will help HPC developers better understand memory access patterns of complex codes, a crucial step in guiding optimisation of codes for arbitrary hardware targets.
 	\end{abstract}
+
 	\maketitle
@@
 				(coalA,8948.96)
-				(coalAB,8894.74)
-				(coalABT,7756.75)
-				(alignABT,6951.7773435)
+				(coalAB,9237.5)
+				(coalABT,5307)
+				(alignABT,5272)
 			};
@@
 				(coalAB,27169.72)
-				(coalABT,23284.03)
-				(alignABT,21363.90)
+				(coalABT,16179.03)
+				(alignABT,16071.90)
 			};	
@@
 	Such programs will perform better on GPUs, as they will make better use of both global memory access coalescing and shared memory bank structures.
+	To a lesser extent, the proposed metric captures performance-critical memory access patterns on CPUs, as pulling a single cacheline from global memory services the threads running on all CPU cores.
+%TODO: Josh, Beau -- Please check this paragraph.
 
@@
 
+	The Extended OpenDwarfs benchmark suite is a set of OpenCL workload classified according to the computation patterns or Dwarfs. First presented in 2006, the Dwarfs categorise 
 	We present the results of running the AIWC tool on the Extended OpenDwarfs benchmark suite \cite{opendwarfs2017base,extendedopendwarfs}.
@@
 	% TODO
-	The OpenDwarfs benchmark GEM computes the electrostatic potential of a biomolecule by calculating the sum of charges contributed by all
+	The OpenDwarfs benchmark GEM computes the electrostatic potential of a biomolecule by calculating the sum of charges contributed by all atoms in the biomolecule at each specific surface vertex.
 	This is an embarrassingly parallel problem.
@@
 This indicates that almost all memory accesses made by the OpenCL kernel are perfectly synchronised between OpenCL threads.
-	%TODO: Clarify above sentence: I mean efficient memory fetches (one load from global memory services all threads) -- not reduced usage of memory.
+
 	\subsection{Dense Linear Algebra: Lower-upper decomposition (LUD)}

### 2020-01-19 paper/paper.tex
minor changes in wording for abstract and intro
Also changed 'ise' to 'ize' globally: see
https://theconversation.com/its-time-to-recognize-and-internalize-the-us-suffix-ize-19828

@@
 %\editor{John B. Appleseed}
-\title[Architecture Independent Memory Access Pattern Analysis]{Characterising Optimisations to Memory Access Patterns using Architecture Independent Program Features}
+\title[Architecture-Independent Memory Access Pattern Analysis]{Characterising Optimizations to Memory Access Patterns using Architecture-Independent Program Features}
 
@@
 
-\keywords{Architecture Independent Analysis, Heterogenous Computing, Workload Characterisation, Memory Access Patterns} %TODO
+\keywords{Architecture Independent Analysis, Heterogenous Computing, Workload Characterization, Memory Access Patterns} %TODO
 
@@
 
-	%This paper addresses the challenge of profiling a code base to help programmers effectively optimise their codes for high performance.
-	HPC developers are faced with the challenge of optimising OpenCL workloads for high performance on novel architectures.
-	The Architecture Independent Workload Characterisation (AIWC) tool is a plugin for the Oclgrind OpenCL simulator that gathers metrics of OpenCL programs used to understand and predict program performance on an arbitrary given hardware architecture.
+	%This paper addresses the challenge of profiling a code base to help programmers effectively optimize their codes for high performance.
+	HPC developers are faced with the challenge of optimizing OpenCL workloads for high performance on novel architectures.
+	The Architecture-Independent Workload Characterization (AIWC) tool is a plugin for the Oclgrind OpenCL simulator that gathers metrics of OpenCL programs used to understand and predict program performance on an arbitrary given hardware architecture.
 	However, AIWC metrics are not always easily interpreted and do not reflect some important memory access patterns affecting efficiency across architectures.
-	We propose the concept of parallel spatial locality -- the locality of memory accesses by OpenCL work-items (threads) executing simultaneously.
-	We implement the parallel spatial locality metric in the AIWC framework, and analyse gathered results on matrix multiply and the Extended OpenDwarfs OpenCL benchmarks.
-	The differences in the obtained parallel spatial locality metric across implementations of matrix multiply reflect the optimisations performed.
+	We propose the concept of parallel spatial locality: the locality of memory accesses by OpenCL work-items (threads) executing simultaneously.
+	We implement the parallel spatial locality metric in the AIWC framework and analyse gathered results on matrix multiply and the Extended OpenDwarfs OpenCL benchmarks.
+	The differences in the parallel spatial locality metric between implementations of matrix multiply reflect the optimizations performed.
 	The new metric showcases the memory access patterns in the OpenDwarfs benchmarks that affect their performance on various architectures.
 	%Current methods of analysing program behaviour have largely relied on performance counters on various architectures to provide architecture dependent metrics of code behaviour. 
-	%However, if workload characteristics are tied to architectural features specific to a particular system, these analysis frameworks may not generalise to alternative or future systems.
-	%Considering the trend towards heterogeneous computing, we work towards a framework to guide developers in optimising codes for any given hardware target using the architecture independent workload characterisation (AIWC) tool -- specifically in profiling memory access patterns of code.
-	%Metrics must be easily interpreted and conceptually linked to patterns in hardware-specific memory access optimisations.
+	%However, if workload characteristics are tied to architectural features specific to a particular system, these analysis frameworks may not generalize to alternative or future systems.
+	%Considering the trend towards heterogeneous computing, we work towards a framework to guide developers in optimizing codes for any given hardware target using the architecture independent workload characterization (AIWC) tool -- specifically in profiling memory access patterns of code.
+	%Metrics must be easily interpreted and conceptually linked to patterns in hardware-specific memory access optimizations.
 	%To this end we propose, implement and analyse the addition of two new metrics to AIWC and present a new approach to profile spatial locality intrinsic to parallel codes.
-	The improvements suggested to AIWC will help HPC developers better understand memory access patterns of complex codes, a crucial step in guiding optimisation of codes for arbitrary hardware targets.
+	The improvements suggested to AIWC will help HPC developers better understand memory access patterns of complex codes, a crucial step in guiding optimization of codes for arbitrary hardware targets.
 	\end{abstract}
@@
 	High-performance computing (HPC) systems are increasingly heterogeneous.
-	A single node on a modern supercomputers may combine a traditional CPUs with one or more accelerators such as GPUs, Field Programmable Gate Arrays (FPGAs), or many integrated core devices (MICs).
-	The development of high bandwidth interconnects between compute devices supports tight integration between multiple devices of different types on the same compute node.
+	A single node on a modern supercomputer may combine traditional CPUs with one or more accelerators such as GPUs, Field Programmable Gate Arrays (FPGAs), or many integrated core devices (MICs).
+	High bandwidth interconnects support tight integration between multiple devices of different types on a single compute node.
 
-	The OpenCL programming language is designed to support modern HPC software engineers in writing code that executes on multiple hardware targets. 
-	%It is possible for an engineer to write highly thread-divergent logical analysis code aimed at traditional CPUs; parallelisable data crunching kernels aimed at GPUs; and code to satisfy specialised computational requirements aimed at FPGAs -- all in a single code base using OpenCL.
-	The language provides support for individual \emph{kernels}, portions of parallelisable codes that make up a HPC code base, to be executed independently on diverse hardware targets such as traditional CPUs, GPUs and FPGAs \cite{khronosopenclworkinggroup2011opencl}.
+	The OpenCL programming language is designed to support portability for heterogeneous systems by allowing a single code to execute on multiple hardware targets. 
+	%It is possible for an engineer to write highly thread-divergent logical analysis code aimed at traditional CPUs; parallelisable data crunching kernels aimed at GPUs; and code to satisfy specialized computational requirements aimed at FPGAs -- all in a single code base using OpenCL.
+	The language provides support for individual \emph{kernels}, portions of parallelizable codes that make up a HPC code base, to be executed independently on diverse hardware targets such as traditional CPUs, GPUs and FPGAs \cite{khronosopenclworkinggroup2011opencl}.
 This gives HPC software developers greater flexibility by allowing codes aimed at a range of hardware targets to be written in a single programming language environment.
@@
 Similarly, compute devices differ in number and capabilities of execution units, processing model, and available resources.
-These heterogeneous computing environments present opportunities for software engineers and HPC integrators to design highly optimised systems with multiple kernels executing on hardware targets best suited for the diverse computational tasks performed by these kernels \cite{spafford2010maestro}.
-	However this opportunity raises the twin challenges of determining the best hardware target for complex kernels and optimising codes to run on these diverse architectures.
-	Our work is directed at helping developers tackle the latter of these two problems.
+	These heterogeneous computing environments present opportunities for software engineers and HPC integrators to design highly optimized systems with multiple kernels executing on hardware targets best suited for the diverse computational tasks performed by each kernel~\cite{spafford2010maestro}.
+	However, this opportunity also presents a challenge in optimizing code to run on diverse architectures.
+	The work reported here aims to help HPC developers understand and optimize memory access patterns to improve performance on heterogeneous systems.
 
-	Modern computer architectures have suffered from the exponential growth in computational speeds being unmet by proportional increases in memory speeds\cite{hennessycomparch}.
-Memory accesses pose larger bottlenecks to performance as applications' demand for main memory scales with the arithmetic capability of computer systems.
-To mitigate this latency, modern CPU designs have employed a wide range of cache technologies to reduce main memory accesses using the principle of \emph{spatial locality}: the idea that programs repeatedly access memory addresses in close locality.
+	In the evolution of computer architectures over the last few decades, the exponential growth in computational capability has not been matched by proportional increases in memory speeds\cite{hennessycomparch}.
+	Memory accesses pose larger bottlenecks to performance as application demand for main memory scales with the arithmetic capability of computer systems.
+	To mitigate this latency, modern CPU designs have employed a wide range of cache technologies to reduce main memory accesses using the principle of \emph{spatial locality}: the observation that data that a program accesses close together in time tend to also be close together in memory.
 	On the other hand, GPUs rely on hardware multithreading to hide memory latency, and their architectures favour ALU capability over sophisticated logic to manage a cache hierarchy and out-of-order execution.
 As a result, the performance of kernels on CPUs and GPUs alike is strongly dependent on memory access patterns intrinsic to the code.
-% We discuss the memory model design and the manufacturer recommended optimisations for NVIDIA GPUs in later sections..
+% We discuss the memory model design and the manufacturer recommended optimizations for NVIDIA GPUs in later sections..
 
 	%We demonstrate the effectiveness of an architecture-independent analysis of codes in determining ...
-Our work guides optimisation efforts for OpenCL codes running on increasingly diverse hardware by examining the current state of the art in architecture-independent metrics for memory access patterns.
+Our work guides optimization efforts for OpenCL codes running on increasingly diverse hardware by examining the current state of the art in architecture-independent metrics for memory access patterns.
 
 	%What we did briefly
-	Our aim in this paper is to develop a framework to guide HPC software engineers in hardware dependent code optimisation -- specifically by guiding the improvement of codes' memory access patterns.
-We provide examples of manufacturer recommended code optimisations to improve memory access patterns on the target architecture, and examine these using AIWC \cite{beauaiwc}, a state of the art architecture-independent code analysis tool for parallel workloads.
-Our work highlights the benefits and challenges arising from an architecture independent analysis of memory based program characteristics.
-We suggest the addition of two new metrics to AIWC to better characterise these memory-based optimisations.
-We measure the presented codes using the new metrics and discuss the metrics' effectiveness in capturing the essence of the optimisations performed.
-With these improvements, we show that AIWC is better suited to be used by HPC software engineers in guiding optimisation of their codes.
+	Our aim in this paper is to develop a framework to guide HPC software engineers in hardware-dependent code optimization -- specifically by guiding the improvement of memory access patterns.
+	We provide examples of manufacturer-recommended code optimizations to improve memory access patterns on the target architecture, and examine these using the architecture-independent workload characterization (AIWC)~\cite{beauaiwc} plugin for Oclgrind.
+	Our work highlights the benefits and challenges arising from an architecture-independent analysis of memory-based program characteristics.
+	We propose two new metrics to AIWC to better characterize these memory-based optimizations.
+	We measure the presented codes using the new metrics and demonstrate the metrics' effectiveness in capturing the essence of the optimizations performed.
 
 	The structure of this paper is as follows.
-In section \ref{motivating example}, we use the example of GPU matrix multiplication code to showcase the specific vendor recommended optimisations our work aims to measure.
-In section \ref{related work} we discuss how AIWC and its precursors profiled memory access patterns in an architecture independent fashion, and relevant architecture dependent approaches to memory access profiling.
-Then in section \ref{method} we provide a methodology to critically examine AIWC, the current state of the art architecture independent workload characterisation tool.
-In section \ref{results} we measure the accuracy of AIWC in profiling the impact of performance optimisations to various OpenCL codes.
-In section \ref{discussion} we suggest new methods to accurately capture code patterns correlating to hardware specific memory optimisations.
-Finally, in section \ref{future work} we discuss further avenues to extend our work and conclude.
+	In section~\ref{motivating example} we use the example of matrix multiplication on a GPU to showcase the specific vendor-recommended optimizations our work aims to measure.
+	In section~\ref{related work} we discuss how AIWC and its precursors profiled memory access patterns in an architecture-independent fashion, and relevant architecture dependent approaches to memory access profiling.
+	In section \ref{method} we present a methodology to critically examine the architecture-independent workload characterization tool, which we use in section~\ref{results} to measure the accuracy of AIWC in profiling the impact of performance optimizations to various OpenCL codes.
+	In section~\ref{discussion} we suggest new methods to accurately capture code patterns correlating to hardware-specific memory optimizations.
+	Finally, in section~\ref{future work} we discuss further avenues to extend our work and conclude.
 
@@
 
-	We aim to capture the essence of vendor recommended optimisations for target architectures in our metrics.
-We demonstrate the optimisation strategies being analysed by presenting a simple unoptimised matrix multiplication kernel for input square matrices of order $N$, and then improve the memory access patterns of this kernel by incrementally performing optimisations recommended in the CUDA Optimisation Handbook \cite{cudaoptimisation} for NVIDIA GPUs.
-These incrementally optimised OpenCL codes are then analysed using AIWC to determine the accuracy in profiling favourable memory access patterns of the current framework and the proposed extensions to AIWC.
+	We aim to capture the essence of vendor recommended optimizations for target architectures in our metrics.
+	We demonstrate the optimization strategies being analysed by presenting a simple unoptimized matrix multiplication kernel for input square matrices of order $N$, and then improve the memory access patterns of this kernel by incrementally performing optimizations recommended in the CUDA Optimization Handbook \cite{cudaoptimisation} for NVIDIA GPUs.
+	These incrementally optimized OpenCL codes are then analysed using AIWC to determine the accuracy in profiling favourable memory access patterns of the current framework and the proposed extensions to AIWC.
 
-	\subsection{Simple unoptimised matrix multiplication}
+	\subsection{Simple unoptimized matrix multiplication}
 
-	The unoptimised matrix multiplication kernel presented in Appendix \ref{simpleMultiply} is used as a baseline to validate the performance improvements measured from each following optimisation.
+	The unoptimized matrix multiplication kernel presented in Appendix \ref{simpleMultiply} is used as a baseline to validate the performance improvements measured from each following optimization.
 Each thread of the kernel computes the $(globalRow, globalCol)$ element of matrix $C$ by reading the appropriate row and column from matrices $A$ and $B$ respectively.
@@
 
-	The code is further optimised by improving how the previous \texttt{coalescedAMultiply} kernel handles accesses to Matrix $B$.
+	The code is further optimized by improving how the previous \texttt{coalescedAMultiply} kernel handles accesses to Matrix $B$.
 While each thread in coalescedAMultiply accesses only a single element of matrix $A$ from global memory, each thread accesses a full column of matrix $B$.
 The repeated reads to elements of matrix $B$ across multiple threads can be alleviated by reading tiles from matrix $B$ into shared memory.
-Appendix \ref{coalescedABMultiply} contains the \texttt{coalescedAB} kernel which performs this optimisation.
+Appendix \ref{coalescedABMultiply} contains the \texttt{coalescedAB} kernel which performs this optimization.
 
@@
 Bank conflicts occur when shared memory in the same bank is accessed concurrently.
-The code in \texttt{coalescedAB} kernel is further optimised by implicitly transposing tiles of Matrix $A$ while loading from global memory.
-This improves memory bank utilisation during reads to shared memory tiles are performed during the necessary computation.
-The \texttt{coalescedABT} kernel demonstrates this optimisation by modifying lines 20,21 of \texttt{coalescedAB} to:
+The code in \texttt{coalescedAB} kernel is further optimized by implicitly transposing tiles of Matrix $A$ while loading from global memory.
+This improves memory bank utilization during reads to shared memory tiles are performed during the necessary computation.
+The \texttt{coalescedABT} kernel demonstrates this optimization by modifying lines 20,21 of \texttt{coalescedAB} to:
 
@@
 
-	Memory access alignment is important to best utilise all parts of GPU memory architecture.
+	Memory access alignment is important to best utilize all parts of GPU memory architecture.
 Global memory buffer alignment can allow threads to access blocks of global memory aligned to the nearest cache line.
@@
 
-	All the examples above are an incomplete listing of optimisation strategies developers can perform to code targeted at NVIDIA GPUs.
-We analyse the correlations between the collected AIWC metrics for each of these kernels and our conceptual understanding of GPU memory optimisations.
-This allows us to examine how effectively our methodology uncovers underlying bottlenecks and guides optimisation efforts.
+	All the examples above are an incomplete listing of optimization strategies developers can perform to code targeted at NVIDIA GPUs.
+We analyse the correlations between the collected AIWC metrics for each of these kernels and our conceptual understanding of GPU memory optimizations.
+This allows us to examine how effectively our methodology uncovers underlying bottlenecks and guides optimization efforts.
 
@@
 	Hoste and Eeckout \cite{hoste2007microarchitecture} show that while conventional architecture dependent characteristics are useful in locating performance bottlenecks, they can hide the underlying, inherent program behaviour causing performance bottlenecks.
-A conceptual understanding of the performance characteristics of complex codes is necessary for the programmer to effectively optimise these codes.
+A conceptual understanding of the performance characteristics of complex codes is necessary for the programmer to effectively optimize these codes.
 Architecture dependent characteristics typically include instructions per cycle (IPC), cache and branch prediction miss rates, page faults and DRAM bus data transfer rates.
 These are typically collected using hardware performance counters available on most target architectures.
-These performance counters do not serve to guide optimisation beyond highlighting potential bottlenecks \cite{hoste2007microarchitecture, ganesan2008performance}.
+These performance counters do not serve to guide optimization beyond highlighting potential bottlenecks \cite{hoste2007microarchitecture, ganesan2008performance}.
 Further, in many cases, architecture dependent characteristics usually cannot be directly correlated to specific code patterns.
 For example, the causes of high cache miss rates in the execution of a program are too complex and widely variant on microarchitecture specific features such as cache size, prefetch behaviour and cache placement policies.
-In summary, we argue that a HPC developer tasked with optimising code for a given hardware target would benefit from architecture-independent metrics of the code that can be conceptually correlated to hardware specific optimisation patterns.
+In summary, we argue that a HPC developer tasked with optimizing code for a given hardware target would benefit from architecture-independent metrics of the code that can be conceptually correlated to hardware specific optimization patterns.
 Such architecture-independent metrics will effectively guide the HPC developer's workflow in both finding and fixing bottlenecks in code performance.
 
-	The Architecture Independent Workload Characterisation (AIWC) tool \cite{beauaiwc} relies on a set of instruction set architecture (ISA)-independent features determined by Shao and Brooks \cite{shao2013isa}.
-AIWC measures memory based metrics to characterise the target code's memory behaviours.
+	The Architecture-Independent Workload Characterization (AIWC) tool \cite{beauaiwc} relies on a set of instruction set architecture (ISA)-independent features determined by Shao and Brooks \cite{shao2013isa}.
+AIWC measures memory based metrics to characterize the target code's memory behaviours.
 AIWC firstly collects aggregate simple counts such as the (i) total memory footprint, the total number of unique addresses accessed; and (ii) 90\% memory footprint, the number of unique addresses covering 90\% of memory accesses.
@@
 	LMAE measures the locality of memory accesses performed over the full execution of a program.
-A steeper drop in entropy with increasing number of bits may correlate to more localised memory accesses over the program's execution.
+A steeper drop in entropy with increasing number of bits may correlate to more localized memory accesses over the program's execution.
 
@@
 
-	\subsection{Advancements in Memory Access Pattern Characterisation}
+	\subsection{Advancements in Memory Access Pattern Characterization}
 
-	We provide a brief summary of past work in architecture independent program characterisation and Memory Access Pattern analysis:
+	We provide a brief summary of past work in architecture independent program characterization and Memory Access Pattern analysis:
 
 	\begin{itemize}
-		\item The Microarchitecture Independent Workload Characterisation Tool (MICA) \cite{hoste2007microarchitecture} was the first microarchitecture independent workload characterisation tool for single threaded programs.
+		\item The Microarchitecture Independent Workload Characterization Tool (MICA) \cite{hoste2007microarchitecture} was the first microarchitecture independent workload characterization tool for single threaded programs.
 		\item The Workload ISA-Independent Characterization for Applications (WIICA) \cite{shao2013isa} extends MICA to present a framework to analyse single threaded programs independent of the ISA.
-		\item CuMAPz: CuMAPz \cite{kim2011cumapz} is a CUDA memory access profiling tool and the first to present metrics for CUDA codes which can be generalised to any given NVIDIA GPU.
+		\item CuMAPz: CuMAPz \cite{kim2011cumapz} is a CUDA memory access profiling tool and the first to present metrics for CUDA codes which can be generalized to any given NVIDIA GPU.
 	\end{itemize}
@@
 Hoste and Eeckout discovered that performance counter based approaches to profiling codes often struggled in finding underlying program features that map to improved or worsened usage of performance critical hardware features of the target architecture.
-The MICA framework is a holistic characterisation tool, and thus collects features including instruction mix, instruction-level parallelism, register traffic, data stream strides and branch predictability.
+The MICA framework is a holistic characterization tool, and thus collects features including instruction mix, instruction-level parallelism, register traffic, data stream strides and branch predictability.
 
@@
 This is directly correlated to cache reuse rates, critical to code performance on CPUs \cite{inteloptimisation}.
-%Intel optimisation manual.
+%Intel optimization manual.
 
@@
 
-	Kim and Shrivastava \cite{kim2011cumapz} present CuMAPz, a CUDA memory access pattern analysis tool to guide NVIDIA GPU optimisations.
-CuMAPz focuses on the problem of improving CUDA application performance using NVIDIA memory-based optimisations.
+	Kim and Shrivastava \cite{kim2011cumapz} present CuMAPz, a CUDA memory access pattern analysis tool to guide NVIDIA GPU optimizations.
+CuMAPz focuses on the problem of improving CUDA application performance using NVIDIA memory-based optimizations.
 CuMAPz analyses CUDA codes structurally and simulates code execution on the memory hierarchy of specific NVIDIA GPU models.
@@
 
-	Our approach is an architecture independent analysis of memory access patterns that provides metrics correlating to similar performance critical memory access optimisations as CuMAPz.
-We aim to further the state of the art by providing a framework to guide developers in optimising OpenCL codes for any given target architecture.
-To the best our knowledge, none of the previous works present a set of performance metrics that accurately characterise memory access patterns of parallel applications independent of the target architecture.
+	Our approach is an architecture independent analysis of memory access patterns that provides metrics correlating to similar performance critical memory access optimizations as CuMAPz.
+We aim to further the state of the art by providing a framework to guide developers in optimizing OpenCL codes for any given target architecture.
+To the best our knowledge, none of the previous works present a set of performance metrics that accurately characterize memory access patterns of parallel applications independent of the target architecture.
 
@@
 
-	To provide memory access pattern metrics that represent vendor recommended optimisations on GPUs, we improved AIWC by adding a new metric: relative local memory usage.
+	To provide memory access pattern metrics that represent vendor recommended optimizations on GPUs, we improved AIWC by adding a new metric: relative local memory usage.
 This measures the proportion of all memory accesses from the symbolic execution of the kernel that occurred to memory allocated as \texttt{\_\_local}.
@@
 
-	Figure \ref{fig: matmul runtime} shows that across varying problem sizes, the vendor recommended optimisations to the matrix multiplication code lead to increased performance.
-Note that a matrix size of 80 is small -- overheads attached to optimisation techniques may be causing the observed decrease in performance as some optimisations are performed.
+	Figure \ref{fig: matmul runtime} shows that across varying problem sizes, the vendor recommended optimizations to the matrix multiplication code lead to increased performance.
+Note that a matrix size of 80 is small -- overheads attached to optimization techniques may be causing the observed decrease in performance as some optimizations are performed.
 
@@
 
-	Table \ref{matmul aiwc table} summarises AIWC memory metrics collected for each of the matrix multiply kernels in section \ref{motivating example}.
+	Table \ref{matmul aiwc table} summarizes AIWC memory metrics collected for each of the matrix multiply kernels in section \ref{motivating example}.
 Note that only two LMAE values are shown for brevity.
-We now analyse the effectiveness of AIWC metrics in profiling the NVIDIA recommended memory optimisations applied to the matrix multiplication kernel.
+We now analyse the effectiveness of AIWC metrics in profiling the NVIDIA recommended memory optimizations applied to the matrix multiplication kernel.
 
@@
 	\textit{Global and local MAE}: Entropy measurements decrease from \texttt{simpleMultiply} to \texttt{coalescedAB}.
-This is explained by the optimisations performed incrementally reducing accesses to matrix $A$ and $B$, and instead performing the bulk of computations on smaller tiles in on-chip shared memory.
+This is explained by the optimizations performed incrementally reducing accesses to matrix $A$ and $B$, and instead performing the bulk of computations on smaller tiles in on-chip shared memory.
 We observe an almost completely uniform distribution of memory accesses in \texttt{simpleMultiply}, where the program makes $N$ loads to each element of $A$ and $B$ with $N$ the dimension of the matrices.
-The frequency distribution of memory addresses accessed become increasingly non-uniform as we perform less accesses to matrices $A$ and $B$ and offset this by increasing the frequency of accesses to smaller, more localised shared memory buffers.
+The frequency distribution of memory addresses accessed become increasingly non-uniform as we perform less accesses to matrices $A$ and $B$ and offset this by increasing the frequency of accesses to smaller, more localized shared memory buffers.
 The non-uniformity of relative frequency with which each memory cell is accessed by the program translates directly to decreases in local and global memory entropy values as reliance on small shared memory buffers increases from \texttt{simpleMultiply} to \texttt{coalescedAB}.
@@
 	\textit{Memory Footprint}: Similar to trends in global and local MAE, we find that the ratio of 90\% memory footprint to total memory footprint decreases from $60.12\%$ for \texttt{simpleMultiply} to $0.25\%$ (\texttt{coalescedAB}).
-Increased utilisation of shared memory in the optimised kernels means that the shared memory buffers make up a greater proportion of total memory accesses in the program.
+Increased utilization of shared memory in the optimized kernels means that the shared memory buffers make up a greater proportion of total memory accesses in the program.
 As the shared memory buffers are small and reused within a workgroup, the memory footprint of the shared memory buffers is also smaller.
 
-	AIWC's metrics strongly reward optimisations that tend to localise memory accesses.
+	AIWC's metrics strongly reward optimizations that tend to localize memory accesses.
 Shared memory buffers are typically smaller than global memory arrays when programming for GPUs due to hardware limitations on sizes of shared memory buffers \cite{cudamanual}.
-However, the metrics currently measured by AIWC do not have a direct causal relation to code patterns that optimise memory accesses on GPUs.
-The proposed relative local memory usage metric is the first to correspond to a recommended optimisation strategy of using fast on-chip shared memory.
+However, the metrics currently measured by AIWC do not have a direct causal relation to code patterns that optimize memory accesses on GPUs.
+The proposed relative local memory usage metric is the first to correspond to a recommended optimization strategy of using fast on-chip shared memory.
 Further, we find that all current AIWC metrics do not measure any sizeable difference between \texttt{coalescedAB}, \texttt{coalescedABT} and \texttt{alignedABT} codes.
@@
 	We propose a new architecture independent metric to measure memory access patterns of parallel programs.
-The proposed metric is inspired by CuMAPz' direct approaches to measuring optimisation specific characteristics of CUDA codes.
+The proposed metric is inspired by CuMAPz' direct approaches to measuring optimization specific characteristics of CUDA codes.
 
@@
 
-	\textit{Summarise}: average the collected entropy values across all the timestamps to calculate the parallel spatial locality metrics for one thread group.
+	\textit{Summarize}: average the collected entropy values across all the timestamps to calculate the parallel spatial locality metrics for one thread group.
 We then average the n-bits dropped entropy summaries across thread groups to obtain the \textit{n-bits dropped parallel spatial locality metrics} for the kernel's execution.
@@
 In particular we observe that the \texttt{coalescedABT} and \texttt{alignedABT} kernels have the steepest drop-offs in entropy as the number of bits skipped is increased, which correlates to better locality of parallel memory accesses.
-It is expected for these kernels to exhibit better parallel spatial locality than \texttt{simpleMultiply}, as \texttt{coalescedABT} and \texttt{alignedABT} make use of shared memory, where accesses tend to be localised simply due to the small size of shared on-chip memory typically available on GPUs.
+It is expected for these kernels to exhibit better parallel spatial locality than \texttt{simpleMultiply}, as \texttt{coalescedABT} and \texttt{alignedABT} make use of shared memory, where accesses tend to be localized simply due to the small size of shared on-chip memory typically available on GPUs.
 Further we find that the proposed metric successfully distinguishes between the \texttt{coalescedAB} and \texttt{coalescedABT} kernels.
-It accurately depicts a steeper drop-off for the more optimised \texttt{coalescedABT} kernel, where a larger proportion of parallel memory accesses make better use of the memory bank structure of GPU shared memory than all previous kernels.
-This is a significant improvement over the state-of-the-art AIWC metrics in characterising how codes localise shared memory accesses to better use the hardware provided.
+It accurately depicts a steeper drop-off for the more optimized \texttt{coalescedABT} kernel, where a larger proportion of parallel memory accesses make better use of the memory bank structure of GPU shared memory than all previous kernels.
+This is a significant improvement over the state-of-the-art AIWC metrics in characterising how codes localize shared memory accesses to better use the hardware provided.
 
@@
 
-	The Extended OpenDwarfs benchmark suite is a set of OpenCL workload classified according to the computation patterns or Dwarfs. First presented in 2006, the Dwarfs categorise 
+	The Extended OpenDwarfs benchmark suite is a set of OpenCL workload classified according to the computation patterns or Dwarfs. First presented in 2006, the Dwarfs categorize 
 	We present the results of running the AIWC tool on the Extended OpenDwarfs benchmark suite \cite{opendwarfs2017base,extendedopendwarfs}.
@@
 Each OpenCL work-item operates on a single surface vertex, finding the electrostatic potential generated by looping through every atom in the biomolecule in a global order.
-	The computation pattern is highly regular and memory accesses are perfectly synchronised. Atom data is accessed consecutively, with all work-items simultaneously accessing each single atom's data.
-	The parallel spatial locality metric reflects this pattern of efficient memory utilisation (single loads from global memory servicing all OpenCL work-items).
+	The computation pattern is highly regular and memory accesses are perfectly synchronized. Atom data is accessed consecutively, with all work-items simultaneously accessing each single atom's data.
+	The parallel spatial locality metric reflects this pattern of efficient memory utilization (single loads from global memory servicing all OpenCL work-items).
 The recorded parallel spatial locality approaches the theoretical limit of 0 (0.0124 at 10 bits dropped).
-This indicates that almost all memory accesses made by the OpenCL kernel are perfectly synchronised between OpenCL threads.
+This indicates that almost all memory accesses made by the OpenCL kernel are perfectly synchronized between OpenCL threads.
 
@@
 	%The use of creates imbalanced workload and divergent control flow within work-items in each kernel iteration.
-While the presence of conditional statements cause some thread divergence in the LUD diagonal kernel, the use of barriers following conditional statements synchronises the threads.
+While the presence of conditional statements cause some thread divergence in the LUD diagonal kernel, the use of barriers following conditional statements synchronizes the threads.
 LUD diagonal and perimeter kernels , while LUD.
@@
 This rightly indicates that memory addresses accessed at each logical timestamp are very distant.
-On GPUs this will translate to poor utilisation of both memory access coalescing and caching.
+On GPUs this will translate to poor utilization of both memory access coalescing and caching.
 This trend in the parallel spatial locality metric suggests that a possible improvement for GPU performance would be to load blocks of the input matrix into on-chip local memory to reduce the number of global memory requests.
@@
 These boundaries constitute a small portion of the executed work-items, especially on larger data-sets and so the effective thread divergence is minimal.
-Thus, memory access patterns in SRAD are statically determined and relatively localised.
-The observed rapid drop-off in parallel spatial locality of the SRAD kernels (figure \ref{fig: opendwarfs parallel spatial locality}) is due to the highly localised nature of parallel memory accesses.
+Thus, memory access patterns in SRAD are statically determined and relatively localized.
+The observed rapid drop-off in parallel spatial locality of the SRAD kernels (figure \ref{fig: opendwarfs parallel spatial locality}) is due to the highly localized nature of parallel memory accesses.
 At each point in the kernels' memory access profile, each work-item $(i,j)$ in an OpenCL work-group accesses the $(i,j)^{th}$ elements of various matrices.
 The sequential memory access pattern is non-linear since different matrices are accessed consecutively by the kernel, prohibiting ideal caching.
-However, memory requests made parallelly are highly localised, allowing memory access coalescing on GPUs, greatly reducing the number of memory requests made on each line of the kernel code on GPUs.
+However, memory requests made parallelly are highly localized, allowing memory access coalescing on GPUs, greatly reducing the number of memory requests made on each line of the kernel code on GPUs.
 Thus while cache hit rates are typically low on this benchmark, particularly on GPUs \cite{krommydas2016opendwarfs}, GPUs can hide the latency of global memory accesses through coalescing.
@@
 The Breadth First Search implemented in OpenDwarfs calls OpenCL kernels to traverse nodes immediately connected to the list of nodes at each depth starting from the root node.
-BFS is characterised by an imbalanced workload per kernel launch depending on the degree of the nodes being operated on -- with only a proportion of launched nodes performing meaningful work.
+BFS is characterized by an imbalanced workload per kernel launch depending on the degree of the nodes being operated on -- with only a proportion of launched nodes performing meaningful work.
 As such the AIWC features collected for each kernel invocation vary significantly.
@@
 
-	To the best of our knowledge this work is the first to critically examine architecture independent methods to measure memory access patterns and guide hardware-specific optimisation.
-We implemented and justified the addition of relative local memory usage to AIWC in our steps to characterise memory-based GPU code optimisations.
+	To the best of our knowledge this work is the first to critically examine architecture independent methods to measure memory access patterns and guide hardware-specific optimization.
+We implemented and justified the addition of relative local memory usage to AIWC in our steps to characterize memory-based GPU code optimizations.
 We have also proposed the methodology to implement a new parallel spatial locality metric to AIWC.
@@
 
-	Our proposed parallel spatial locality metric may also correlate to some memory-based optimisation strategies for CPUs.
-Future work would apply the approach followed by this paper to CPU and FPGA architectures to evaluate the viability of an architecture independent characterisation of memory access patterns in codes targeted to these architectures.
+	Our proposed parallel spatial locality metric may also correlate to some memory-based optimization strategies for CPUs.
+Future work would apply the approach followed by this paper to CPU and FPGA architectures to evaluate the viability of an architecture independent characterization of memory access patterns in codes targeted to these architectures.
 

### 2020-01-19 paper/paper.tex
Finished results, improving overall flow of ideas in the paper.

--- 2020-01-18 paper/paper.tex
@@
 	HPC developers are faced with the challenge of optimising OpenCL workloads for high performance on novel architectures.
-	The Architecture Independent Workload Characterisation (AIWC) tool is a plugin for the Oclgrind OpenCL simulator that gathers metrics of OpenCL programs used to understand and predict program performance on an arbitrary given hardware architecture.
+	The Architecture Independent Workload Characterisation (AIWC) tool is a plugin for the Oclgrind OpenCL simulator that gathers metrics of OpenCL programs that can be used to understand and predict program performance on an arbitrary given hardware architecture.
 	However, AIWC metrics are not always easily interpreted and do not reflect some important memory access patterns affecting efficiency across architectures.
-	We propose the concept of parallel spatial locality -- the locality of memory accesses by OpenCL work-items (threads) executing simultaneously.
+	We propose the concept of parallel spatial locality -- the closeness of memory accesses simultaneously issued by OpenCL work-items (threads).
 	We implement the parallel spatial locality metric in the AIWC framework, and analyse gathered results on matrix multiply and the Extended OpenDwarfs OpenCL benchmarks.
 	The differences in the obtained parallel spatial locality metric across implementations of matrix multiply reflect the optimisations performed.
-	The new metric showcases the memory access patterns in the OpenDwarfs benchmarks that affect their performance on various architectures.
+	The new metric can be used to distinguish between the OpenDwarfs benchmarks based on the memory access patters affecting their performance on various architectures.
 	%Current methods of analysing program behaviour have largely relied on performance counters on various architectures to provide architecture dependent metrics of code behaviour. 
@@
 	%It is possible for an engineer to write highly thread-divergent logical analysis code aimed at traditional CPUs; parallelisable data crunching kernels aimed at GPUs; and code to satisfy specialised computational requirements aimed at FPGAs -- all in a single code base using OpenCL.
-	The language provides support for individual \emph{kernels}, portions of parallelisable codes that make up a HPC code base, to be executed independently on diverse hardware targets such as traditional CPUs, GPUs and FPGAs \cite{khronosopenclworkinggroup2011opencl}.
+	%The language provides support for individual \emph{kernels}, portions of parallelisable codes that make up a HPC code base, to be executed independently on diverse hardware targets such as traditional CPUs, GPUs and FPGAs \cite{khronosopenclworkinggroup2011opencl}.
 	This gives HPC software developers greater flexibility by allowing codes aimed at a range of hardware targets to be written in a single programming language environment.
@@
 	Our aim in this paper is to develop a framework to guide HPC software engineers in hardware dependent code optimisation -- specifically by guiding the improvement of codes' memory access patterns.
-We provide examples of manufacturer recommended code optimisations to improve memory access patterns on the target architecture, and examine these using AIWC \cite{beauaiwc}, a state of the art architecture-independent code analysis tool for parallel workloads.
-Our work highlights the benefits and challenges arising from an architecture independent analysis of memory based program characteristics.
+%TODO FIXME: remove this or reword it.
+Personally don't think it sounds good
+	We provide examples of manufacturer recommended code optimisations to improve memory access patterns on the target architecture, and examine these using AIWC \cite{beauaiwc}, a state of the art architecture-independent code analysis tool for OpenCL workloads.
 	We suggest the addition of two new metrics to AIWC to better characterise these memory-based optimisations.
 	We measure the presented codes using the new metrics and discuss the metrics' effectiveness in capturing the essence of the optimisations performed.
+	Further, we validate the effectiveness of the added metrics in finding favourable memory access patterns in the Extended OpenDwarfs benchmarks.
 	With these improvements, we show that AIWC is better suited to be used by HPC software engineers in guiding optimisation of their codes.
@@
 Then in section \ref{method} we provide a methodology to critically examine AIWC, the current state of the art architecture independent workload characterisation tool.
-In section \ref{results} we measure the accuracy of AIWC in profiling the impact of performance optimisations to various OpenCL codes.
-In section \ref{discussion} we suggest new methods to accurately capture code patterns correlating to hardware specific memory optimisations.
+In section \ref{preliminary results} we measure the accuracy of AIWC in profiling the impact of performance optimisations to various OpenCL codes.
+In section \ref{discussion} we propose and implement our parallel spatial locality metric into the AIWC framework.
+In section \ref{results} we present the collected metric for selected OpenDwarfs benchmarks to validate our methodology in this paper.
 Finally, in section \ref{future work} we discuss further avenues to extend our work and conclude.
+	%TODO: Change the above paragraph as paper structure evolves
 
@@
 
-	\section{Results} \label{results}
+	\section{Preliminary Results} \label{preliminary results}
 
@@
 
-			\addplot[mark=*] coordinates {
-				(simple,27.79)
-				(coalA,16.36)
-				(coalAB,17.02)
-				(coalABT,16.34)
-				(alignABT,30.21)
+			\addplot coordinates {%[mark=*] coordinates {
+				(simple,31)
+				(coalA,19)
+				(coalAB,20)
+				(coalABT,18)
+				(alignABT,18)
 			};
-			\addplot[mark=star] coordinates {
-				(simple,319.71)
-				(coalA,70.12)
-				(coalAB,73.60)
-				(coalABT,65.27)
-				(alignABT,61.50)
+			\addplot coordinates {%[mark=star] coordinates {
+				(simple,325)
+				(coalA,75)
+				(coalAB,81)
+				(coalABT,55)
+				(alignABT,55)
 			};	
@@
 
-			\addplot[mark=triangle*] coordinates {
-				(simple,46901.35)
-				(coalA,8948.96)
+			\addplot [mark=*, color=brown] coordinates {%[mark=triangle*] coordinates {
+				(simple,46920)
+				(coalA,8783)
 				(coalAB,9237.5)
@@
 			};
-			\addplot[mark=square] coordinates {
-				(simple,130601.41)
-				(coalA,28722.51)
-				(coalAB,27169.72)
-				(coalABT,16179.03)
-				(alignABT,16071.90)
+			\addplot [mark=star] coordinates {%[mark=square] coordinates {
+				(simple,130682.0)
+				(coalA,28602)
+				(coalAB,27939)
+				(coalABT,16179)
+				(alignABT,16071)
 			};	
@@
 	Figure \ref{fig: matmul runtime} shows that across varying problem sizes, the vendor recommended optimisations to the matrix multiplication code lead to increased performance. 
-Note that a matrix size of 80 is small -- overheads attached to optimisation techniques may be causing the observed decrease in performance as some optimisations are performed.
+	%Note that a matrix size of 80 is small -- overheads attached to optimisation techniques may be causing the observed decrease in performance as some optimisations are performed.
 
@@
 
-	\section{Discussion} \label{discussion}
+	\section{Discussion } \label{discussion}
 
@@
 
-		\addplot[mark=*] coordinates {
-			(0, 4.01)
-			(1,3.51)
-			(2,3.01)
-			(3,2.50)
-			(4,2.00)
-			(5,2.00)
-			(6,2.00)
-			(7,2.00)
-			(8,2.00)
-			(9,1.5)
-			(10,1.00)
+		\addplot coordinates {%[mark=*] coordinates {
+			(0, 3.987)
+			(1, 3.987)
+			(2, 3.987)
+			(3, 3.49)
+			(4, 2.992)
+			(5, 2.494)
+			(6, 1.996)
+			(7, 1.996)
+			(8, 1.996)
+			(9, 1.996)
+			(10, 1.996)
 		};
 
-		\addplot[mark=star] coordinates {
-			(0,4.24)
-			(1,3.71)
-			(2,3.18)
-			(3,2.65)
-			(4,2.12)
-			(5,1.62)
-			(6,1.12)
-			(7,0.62)
-			(8,0.12)
-			(9,0.09)
-			(10,0.06)
+		\addplot coordinates {%[mark=star] coordinates {
+			(0,3.952)
+			(1,3.952)
+			(2,3.952)
+			(3,3.459)
+			(4,2.965)
+			(5,2.472)
+			(6,1.979)
+			(7,1.506)
+			(8,1.033)
+			(9,0.5598)
+			(10,0.08691)
 		};
 
-		\addplot[mark=triangle*] coordinates {
-			(0,4.45)
-			(1,3.89)
-			(2,3.33)
-			(3,2.78)
-			(4,2.23)
-			(5,1.73)
-			(6,1.23)
-			(7,0.73)
-			(8,0.23)
-			(9,0.17)
-			(10,0.11)
+		\addplot coordinates {%[mark=triangle*] coordinates {
+			(0,4.145)
+			(1,4.145)
+			(2,4.145)
+			(3,3.628)
+			(4,3.11)
+			(5,2.598)
+			(6,2.075)
+			(7,1.603)
+			(8,1.13)
+			(9,0.6579)
+			(10,0.1854)
 		};
 
-		\addplot[mark=square] coordinates {
-			(0,4.45)
-			(1,3.45)
-			(2,2.45)
-			(3,1.45)
-			(4,0.45)
-			(5,0.40)
-			(6,0.34)
-			(7,0.28)
-			(8,0.23)
-			(9,0.17)
-			(10,0.11)
+		\addplot [mark=*, color=red] coordinates {%[mark=square] coordinates {
+			(0,4.145)
+			(1,4.145)
+			(2,4.145)
+			(3,3.2)
+			(4,2.254)
+			(5,1.308)
+			(6,0.3621)
+			(7,0.318)
+			(8,0.2738)
+			(9,0.2296)
+			(10,0.1854)
 		};
 
-		\addplot[mark=o] coordinates {
-			(0,4.45)
-			(1,3.45)
-			(2,2.45)
-			(3,1.45)
-			(4,0.45)
-			(5,0.40)
-			(6,0.34)
-			(7,0.28)
-			(8,0.23)
-			(9,0.17)
-			(10,0.11)
+		\addplot coordinates {%[mark=o] coordinates {
+			(0,4.145)
+			(1,4.145)
+			(2,4.145)
+			(3,3.2)
+			(4,2.254)
+			(5,1.308)
+			(6,0.3621)
+			(7,0.318)
+			(8,0.2738)
+			(9,0.2296)	
+			(10,0.1854)
 		};
@@
 
-	\textit{Summarise}: average the collected entropy values across all the timestamps to calculate the parallel spatial locality metrics for one thread group.
+	\textit{Summarise}: average the collected entropy values across all the timestamps to calculate the \emph{parallel spatial locality} metrics for one thread group.
 We then average the n-bits dropped entropy summaries across thread groups to obtain the \textit{n-bits dropped parallel spatial locality metrics} for the kernel's execution.
@@
 
-	In figure \ref{fig: matmul parallel spatial locality} we see theoretical calculations for the proposed parallel spatial locality metric for each of the matrix multiplication kernels from section \ref{motivating example}.
+	\section{Results} \label{results}
+
+	In figure \ref{fig: matmul parallel spatial locality} we see the proposed parallel spatial locality metric obtained using the AIWC tool for each of the matrix multiplication kernels from section \ref{motivating example}.
 In particular we observe that the \texttt{coalescedABT} and \texttt{alignedABT} kernels have the steepest drop-offs in entropy as the number of bits skipped is increased, which correlates to better locality of parallel memory accesses.
@@
 
-	\section{Results}
+	The Extended OpenDwarfs benchmark suite is a set of OpenCL workload classified according to the computation patterns or Dwarfs. 
+	First presented in 2006, the 13 Berkeley Dwarfs are common computational and communication patterns that aim to capture the landscape of parallel computing workloads. 
+	The Extended OpenDwarfs benchmark suite presents a set of highly diverse workloads that each fit the programming paradigm of one of the 13 Berkeley Dwarfs.
+	The workloads presented are not optimised for any specific architecture -- hence optimisations using OpenCL local memory (which translates to CUDA shared memory) are not performed. 
+	Many of these benchmarks can be run with up to four \emph{problem sizes} based on the sizes of caches found in modern CPU memory heirarchies \cite{johnston18opendwarfs}.
+	We present the results of running the AIWC tool on selected benchmarks from the Extended OpenDwarfs benchmark suite \cite{opendwarfs2020head,johnston18opendwarfs} with a problem size setting of \emph{medium}, except for the GEM benchmark, which is run at a setting of \emph{tiny}.
+	For benchmarks such as BFS with multiple kernel invocations per run, the AIWC parallel spatial locality metrics of the invocation with the highest executed LLVM-SPIR instructions is presented for analysis.
+Note that the presented benchmarks use 32 bit numbers (OpenCL \texttt{int} and \texttt{float} types), and so dropping up to two bits of the memory addresses accessed will not change the data entropy, since any addresses accessed are at least 4 bytes apart.
 
-	The Extended OpenDwarfs benchmark suite is a set of OpenCL workload classified according to the computation patterns or Dwarfs. First presented in 2006, the Dwarfs categorise 
-	We present the results of running the AIWC tool on the Extended OpenDwarfs benchmark suite \cite{opendwarfs2017base,extendedopendwarfs}.
+	%TODO: Add some info about EOD benchmarks here?
 
@@
 		ylabel={$n$-bits dropped Parallel Spatial Locality}, xlabel={\# of bits skipped},
+		legend columns=4,
 		ylabel style={at={(axis description cs:0.05,.5)},anchor=south},
-		legend entries={\texttt{GEM}, \texttt{NW*}, \texttt{SRAD Kernel 1},\texttt{SRAD Kernel 2},
-			\texttt{CSR}, \texttt{BFS*}, \texttt{LUD Diagonal}, \texttt{LUD Internal}, \texttt{LUD Perimeter}, \texttt{TDM}},
-		legend style={at={(0.55,0.75)},anchor=west}
+		legend entries={\texttt{GEM}, \texttt{NW}, \texttt{SRAD Kernel 1},\texttt{SRAD Kernel 2},
+			\texttt{CSR}, \texttt{BFS Kernel 1}, \texttt{BFS Kernel 2}, \texttt{LUD Diagonal}, \texttt{LUD Perimeter}, \texttt{LUD Internal}, \texttt{TDM}},
+		legend style={at={(0.5,-0.15)},anchor=north}
 		]
 
-		\addplot[mark=*] coordinates {
+		\addplot coordinates {%[mark=*] coordinates {
 			(0, 0.1139)
@@
 			(2, 0.1139)
-			(3, 0.1139)
-			(4, 0.0989)
-			(5, 0.08402)
-			(6, 0.06914)
-			(7, 0.05431)
-			(8, 0.03938)
-			(9, 0.02442)
-			(10, 0.0124)
+			(3, 0.0989)
+			(4, 0.08402)
+			(5, 0.06914)
+			(6, 0.05431)
+			(7, 0.03938)
+			(8, 0.02442)
+			(9, 0.0124)
+			(10, 0.00038)
 		};
 
-		\addplot[mark=star] coordinates {
+		\addplot coordinates {%[mark=star] coordinates {
 			(0, 2.175)
@@
 
-		\addplot[mark=*] coordinates {
-			(0, 7.232)
-			(1, 7.232)
-			(2, 7.232)
-			(3, 7.232)
-			(4, 6.351)
-			(5, 5.461)
-			(6, 4.570)
-			(7, 3.678)
-			(8, 3.648)
-			(9, 3.633)
-			(10, 3.625)
+		\addplot coordinates {%[mark=*] coordinates {
+			(0, 7.176)
+			(1, 7.176)
+			(2, 7.176)
+			(3, 6.301)
+			(4, 5.418)
+			(5, 4.533)
+			(6, 3.648)
+			(7, 3.619)
+			(8, 3.605)
+			(9, 3.598)
+			(10, 3.594)
 		};
 
-		\addplot[mark=star] coordinates {
-			(0, 7.098)
-			(1, 7.098)
-			(2, 7.098)
-			(3, 7.098)
-			(4, 6.223)
-			(5, 5.344)
-			(6, 4.464)
-			(7, 3.583)
-			(8, 3.567)
-			(9, 3.559)
-			(10, 3.556)
+		\addplot coordinates {%[mark=star] coordinates {
+			(0, 7.097)
+			(1, 7.097)
+			(2, 7.097)
+			(3, 6.223)
+			(4, 5.344)
+			(5, 4.464)
+			(6, 3.583)
+			(7, 3.567)
+			(8, 3.559)
+			(9, 3.554)
+			(10, 3.553)
 		};
 
-		\addplot[mark=triangle*] coordinates {
+		\addplot coordinates {%[mark=triangle*] coordinates {
 			(0, 9.605)
@@
 			(2, 9.605)
-			(3, 9.605)
-			(4, 9.478)
-			(5, 9.293)
-			(6, 9.049)
-			(7, 8.761)
-			(8, 8.447)
-			(9, 8.12)
-			(10,7.233)
+			(3, 9.478)
+			(4, 9.293)
+			(5, 9.049)
+			(6, 8.761)
+			(7, 8.447)
+			(8, 8.12)
+			(9,7.233)
+			(10,6.287)
 		};
 
-		\addplot[mark=square] coordinates {
-			(0,4.406)
-			(1,4.406)
-			(2,4.406)
-			(3,4.406)
-			(4,4.361)
-			(5,4.273)
-			(6,4.135)
-			(7,3.941)
-			(8,3.689)
-			(9,3.382)
-			(10,3.029)
+		\addplot coordinates { %[mark=square] coordinates {
+			(0,4.933)	
+			(1,4.933)
+			(2,4.933)
+			(3,4.89)
+			(4,4.795)
+			(5,4.652)
+			(6,4.456)
+			(7,4.215)
+			(8,3.93)
+			(9,3.608)
+			(10,3.253)
 		};
 
-		\addplot[mark=o] coordinates {
+		\addplot coordinates { %[mark=square] coordinates {
+			(0,4.691)
+			(1,4.691)
+			(2,4.691)
+			(3,4.276)
+			(4,3.760)
+			(5,3.171)
+			(6,2.552)
+			(7,1.922)
+			(8,1.286)
+			(9,0.648)
+			(10,0.00776)
+		};
+
+		\addplot coordinates { %[mark=o] coordinates {
 			(0,1.548)
@@
 			(2,1.548)
-			(3,1.548)
-			(4,1.305)
-			(5,1.092)
-			(6,0.9385)
+			(3,1.305)
+			(4,1.092)
+			(5,0.9385)
+			(6,0.8796)
 			(7,0.8796)
@@
 		};
+
+	    \addplot coordinates { %[mark=star] coordinates {
+	    	(0,3.803)
+	    	(1,3.803)
+	    	(2,3.803)
+	    	(3,3.473)
+	    	(4,3.144)
+	    	(5,2.814)
+	    	(6,2.485)
+	    	(7,2.485)
+	    	(8,2.485)
+	    	(9,2.485)
+	    	(10,2.485)
+    	};
+
+   		\addplot coordinates { %[mark=square*] coordinates {
+   			(0,4.101)
+   			(1,4.101)
+   			(2,4.101)
+   			(3,3.589)
+   			(4,3.077)
+   			(5,2.565)
+   			(6,2.053)
+   			(7,2.053)
+   			(8,2.053)
+   			(9,2.053)
+   			(10,2.053)
+   		};
+
 		\end{axis}
@@
 	\end{figure*}
+
 	\subsection{N-body methods: GEM}
@@
 
+	Performance results \cite{dwarfsonaccelerators, krommydas2016opendwarfs} show that GPUs perform significantly better than CPUs for GEM, as memory unit stalls are at low levels for both CPUs and GPUs due to the highly efficient memory utilisation of this benchmark.
+	As memory operations do not present a major slowdown, increased FPU compute capability of GPUs leads to far better performance on GPUs than CPUs \cite{krommydas2016opendwarfs}.
+
 	\subsection{Dense Linear Algebra: Lower-upper decomposition (LUD)}
@@
 	LUD in OpenDwarfs is a program to decompose an input $N\times N$ matrix as the product of one lower and one upper diagonal matrix. 
-LUD is a Memory access patterns in a dense linear algebra workload such as LUD are typically highly regular and deterministic for each OpenCL work item, based on the the matrix dimension and offset parameters.
-The OpenDwarfs implementation of LUD \cite{opendwarfs2017base} splits the LUD computation into three kernels: LUD perimeter and diagonal kernels spawn work-items in a single work dimension while LUD internal performs work in both dimensions of the input matrix.
+	Memory access patterns in a dense linear algebra workload such as LUD are typically highly regular and deterministic for each OpenCL work item, based on the matrix dimension and offset parameters.
+The OpenDwarfs implementation of LUD \cite{opendwarfs2020head} splits the LUD computation into three kernels: \texttt{LUD perimeter} and \texttt{LUD diagonal} kernels spawn work-items in a single work dimension; while \texttt{LUD internal} performs work in both dimensions of the input matrix.
 
-	LUD perimeter and diagonal partially benefit from memory access coalescing, for the lines of code where all the work-items access contiguous memory along a row of the input matrix.
+	The LUD kernels partially benefit from memory access coalescing on GPUs, for the lines of code where all the work-items access contiguous memory along a row of the input matrix. 
 	However, when threads parallelly access a column of the input matrix, multiple memory requests are made as addresses accessed are too distant to be coalesced into a single memory transaction on GPUs.
-This is reflected in the slow drop in parallel spatial locality for LUD perimeter and diagonal kernels.
-
-	LUD internal %TODO: 2Dimensional work -- better parallelism
+	This is reflected in the drop in parallel spatial locality for the LUD kernels to approximately half the value at 0 bits skipped.
+Taking \texttt{LUD internal} as an example, the 0-bits skipped parallel spatial locality metric is 4.101, while the 10-bits skipped parallel spatial locality metric is 2.053.
+	The swift decline of the metric to 2.053 from 2 to 6-bits skipped parallel spatial locality indicates that approximately half the parallel memory accesses in \texttt{LUD internal} are highly localised -- where work-items simultaneously access contiguous memory along a matrix row.
+	Similar to the presented matrix multiply codes, LUD can be optimised for GPUs by effective utilisation of shared memory \cite{opendwarfs2020head}.
 
@@
 LUD diagonal and perimeter kernels , while LUD.
-
-	...  %TODO: Divergence in the kernels. Make a diagram/code simplification for LUD.
+	%TODO: Divergence in the kernels. Make a diagram/code simplification for LUD.
 
@@
 This dependency enforces a wavefront computation pattern, travelling along the main diagonal of the matrix.
-Each iteration of the kernels computes over an antidiagonal of the matrix, starting at its top left corner and finishing at its bottom right corner.
+Each iteration of the kernels computes over an antidiagonal of the matrix, starting at its top left corner, and finishing at its bottom right corner.
 
@@
 This rightly indicates that memory addresses accessed at each logical timestamp are very distant.
-On GPUs this will translate to poor utilisation of both memory access coalescing and caching.
-This trend in the parallel spatial locality metric suggests that a possible improvement for GPU performance would be to load blocks of the input matrix into on-chip local memory to reduce the number of global memory requests.
+On GPUs this will translate to poor utilisation of both memory access coalescing and caching \cite{krommydas2016opendwarfs}.
+This trend in the parallel spatial locality metric suggests that a possible improvement for GPU performance would be to load blocks of the input matrix into on-chip local memory to reduce the number of global memory requests -- this is typically performed in GPU optimised implementations of the Needleman-Wunsch algorithm \cite{opendwarfs2020head}.
 
@@
 
-	SRAD removes locally correlated noise from images by following a repeated grid update computation pattern.
+	SRAD removes locally correlated noise from images by following a repeated grid update computation pattern on image pixel grids. 
 	Conditional statements in the code cause thread divergence, potentially within a work-group, to handle boundary conditions. 
 	These boundaries constitute a small portion of the executed work-items, especially on larger data-sets and so the effective thread divergence is minimal.
-Thus, memory access patterns in SRAD are statically determined and relatively localised.
-The observed rapid drop-off in parallel spatial locality of the SRAD kernels (figure \ref{fig: opendwarfs parallel spatial locality}) is due to the highly localised nature of parallel memory accesses.
-At each point in the kernels' memory access profile, each work-item $(i,j)$ in an OpenCL work-group accesses the $(i,j)^{th}$ elements of various matrices.
-The sequential memory access pattern is non-linear since different matrices are accessed consecutively by the kernel, prohibiting ideal caching.
-However, memory requests made parallelly are highly localised, allowing memory access coalescing on GPUs, greatly reducing the number of memory requests made on each line of the kernel code on GPUs.
-Thus while cache hit rates are typically low on this benchmark, particularly on GPUs \cite{krommydas2016opendwarfs}, GPUs can hide the latency of global memory accesses through coalescing.
+	Memory access patterns in SRAD are statically determined and relatively localised. 
+	Similar to the LUD and \texttt{simpleMultiply} kernels, both SRAD kernels observe a rapid drop-off in parallel spatial locality (figure \ref{fig: opendwarfs parallel spatial locality}) as the number of bits skipped is increased, with the metric stabilising at approximately half its original value when 0 bits are skipped.
+
+	At each point in the kernels' memory access profile, each work-item with global ID $(i,j)$ in an OpenCL work-group accesses the $(i,j)^{th}$ elements of various matrices.
+The sequential memory access pattern is non-linear since different matrices are accessed consecutively by the kernel, prohibiting ideal caching \cite{krommydas2016opendwarfs}.
+However, memory requests made simultaneously by a work-group always fall within a rectangular block of one of the matrices.
+	This allows memory access coalescing on GPUs for OpenCL work-items accessing contiguous matrix data, greatly reducing the number of memory requests made on each line of the kernel code on GPUs. 
+	However, work-items accessing data along a column of the matrix do not observe memory access coalescing.
+	Thus we observe that while cache hit rates are typically low on this benchmark, particularly on GPUs \cite{krommydas2016opendwarfs}, GPUs can hide the latency of global memory accesses through memory access coalescing to some extent.
 
@@
 	Graph traversal algorithms require pointer chasing operations to traverse nodes of a graph and perform calculations. 
-The Breadth First Search implemented in OpenDwarfs calls OpenCL kernels to traverse nodes immediately connected to the list of nodes at each depth starting from the root node.
-BFS is characterised by an imbalanced workload per kernel launch depending on the degree of the nodes being operated on -- with only a proportion of launched nodes performing meaningful work.
-As such the AIWC features collected for each kernel invocation vary significantly.
-The parallel spatial locality metric collects the entropy of memory accesses at each time step, and workload imbalances are dealt with by averaging entropy scores across all the time steps in the execution of an OpenCL work-group.
+	The Breadth First Search implemented in OpenDwarfs calls two OpenCL kernels to traverse nodes immediately connected to the list of nodes at each depth starting from the root node. 
+	BFS is characterised by an imbalanced workload per kernel launch depending on the degree of the nodes being operated on, with only a proportion of launched nodes performing meaningful work. 
+	As such, the AIWC features collected for each kernel invocation vary significantly. 
+	The parallel spatial locality metric collects the entropy of memory accesses at each time step, and workload imbalances across work-items are dealt with by averaging entropy scores across all the time steps in the execution of an OpenCL work-group.
 
-	The memory access patterns in BFS are irregular.
+	Of the two BFS kernels, \texttt{BFS kernel 1} performs the graph traversal necessary to generate a list of neighbours at each node. The memory access patterns in \texttt{BFS kernel 1} are irregular. 
 	Work-items fetch discontinuous memory locations attributed to any particular node in the graph, depending on the connectivity of the node they operate on. 
 	The program structure involves multiple levels of pointer-chasing and thus the precise memory address accessed are runtime dependent. 
-This leads to poor parallel spatial locality.
-Thus, we see a slow dropoff in parallel spatial locality for BFS.
+	This leads to poor parallel spatial locality. Thus, we see a slow dropoff in parallel spatial locality for \texttt{BFS kernel 1}.
 
@@
 The matrix is stored in a compressed row storage sparse matrix format, which is very efficient for storage when the number of zero elements is far greater than the number of non-zero elements.
+Three inputs are provided to the CSR kernel.
 Only non-zero elements of each row of a matrix are stored in $Ax$, along with separate arrays $Aj,\ Ap$ indicating the position of each non-zero element in the matrix.
-Figure \ref{?} depicts the sparse matrix format of storage for a $4\times4$ matrix.
 
 	In this particular implementation, a single OpenCL work-item is assigned to compute over each row of the input sparse matrix. 
-Similar to BFS, memory access patterns are runtime-dependent, since the locations of matrix data read by each thread are dependent on the number and position of non-zero values in the sparse-matrix, decided by the values in $Aj$ and $Ap$.
+	The locations of matrix data read by each work-item are dependent on the number and position of non-zero values in the sparse-matrix, which are decided by the values in $Aj$ and $Ap$. 
+	Similar to BFS, this means memory access patterns are runtime-dependent due to indirect addressing (pointer chasing).
 	This pattern of indirect addressing is typical of applications in the Spare Linear Algebra Dwarf, which severely hinders locality of memory accesses performed.
@@
 
-	\subsection{Finite State Machine: Temporal Data Mining (TDM)}
-
-	%TODO
-
 	\section{Conclusions and Future Work} \label{future work}
 
-	To the best of our knowledge this work is the first to critically examine architecture independent methods to measure memory access patterns and guide hardware-specific optimisation.
+	To the best of our knowledge this work is the first to critically examine AIWC's measurement of memory access patterns to guide hardware-specific optimisation. 
 	We implemented and justified the addition of relative local memory usage to AIWC in our steps to characterise memory-based GPU code optimisations.
-We have also proposed the methodology to implement a new parallel spatial locality metric to AIWC.
-Upon implementation, our work would be the first to measure memory access patterns of parallel codes in an architecture-independent, easily interpreted manner.
+	We proposed the concept of parallel spatial locality to capture the idea of closeness of memory accesses made by parallel OpenCL workloads.
+	We then implemented a new parallel spatial locality metric to AIWC, presented results of running AIWC on matrix multiply kernels and selected OpenDwarfs benchmarks and validated our methodology. 
+	Our work is the first to measure memory access patterns of parallel codes in an architecture-independent, easily interpreted manner.
+%Josh, Beau: is this too bold a claim?
+If unnecessary, please remove.
 	%TODO UPDATE IN FUTURE
@@
 
-	Future work will provide a concrete implementation of this new parallel spatial locality metric in the AIWC framework.
-It is then feasible to test these new metrics across a larger set of GPU targeted OpenCL codes.
-This will allow a critical examination of the effectiveness of both proposed metrics in accurately profiling memory access patterns of these diverse codes.
+	Our work focused on measuring memory access patterns in OpenCL codes intends to guide optimisation efforts.
+Another potential use case for AIWC is providing architecture-independent performance predictions for OpenCL kernels by generating machine learning models based on AIWC metrics \cite{beauperformanceprediction}.
+Future work would modify the presented metrics and develop new memory metrics with the specific intent of being fed to machine learning models to predict kernel performance on arbitrary and novel architectures.
+%FIXME: I want to say that we have been focusing on _guiding optimisation_.
+We can do more work in making the AIWC metrics better for _performance prediction_.
+Is this coming across okay?
 

### 2020-01-19 paper/paper.tex
minor grammar / wording changes in Section 2

--- 2020-01-19 paper/paper.tex
@@
 
-	We aim to capture the essence of vendor recommended optimizations for target architectures in our metrics.
-	We demonstrate the optimization strategies being analysed by presenting a simple unoptimized matrix multiplication kernel for input square matrices of order $N$, and then improve the memory access patterns of this kernel by incrementally performing optimizations recommended in the CUDA Optimization Handbook \cite{cudaoptimisation} for NVIDIA GPUs.
+	We aim to capture the essence of vendor-recommended optimizations for target architectures in our metrics.
+	We demonstrate the optimization strategies using an OpenCL kernel which multiplies square matrices of order $N$.
+	We start with a simple unoptimized kernel, and then improve the memory access patterns of this kernel by incrementally performing the optimizations recommended in the CUDA Optimization Handbook \cite{cudaoptimisation} for NVIDIA GPUs.
 	These incrementally optimized OpenCL codes are then analysed using AIWC to determine the accuracy in profiling favourable memory access patterns of the current framework and the proposed extensions to AIWC.
 
-	\subsection{Simple unoptimized matrix multiplication}
+	\subsection{Simple Unoptimized Matrix Multiplication}
 
-	The unoptimized matrix multiplication kernel presented in Appendix \ref{simpleMultiply} is used as a baseline to validate the performance improvements measured from each following optimization.
-Each thread of the kernel computes the $(globalRow, globalCol)$ element of matrix $C$ by reading the appropriate row and column from matrices $A$ and $B$ respectively.
+	The unoptimized matrix multiplication kernel presented in Appendix~\ref{simpleMultiply} is used as a baseline to validate the performance improvements measured from each following optimization.
+	Each thread of the kernel updates the $(globalRow, globalCol)$ element of matrix $C$ by computing the dot product of the corresponding row of $A$ and column of $B$.
 
@@
 
-	We first notice that the number of global memory accesses to matrices A and B increases at an order $O(N^3)$ with respect to matrix size $N$.
+	We first notice that the number of global memory accesses to matrices A and B increases as $O(N^3)$ with respect to matrix size $N$.
 	Global memory is typically located off-chip and accesses induce large delays.
 	NVIDIA GPUs coalesce global memory loads and stores issued within thread-groups into as few DRAM transactions as possible.
-Multiple global memory loads and stores are coalesced into a single transaction when certain device specific conditions are met.
-On most NVIDIA GPUs, data accesses are coalesced when multiple requests are made for memory locations from the same cache-line in global memory \cite{cudamanual}.
+	Multiple global memory loads and stores are coalesced into a single transaction when certain device-specific conditions are met. 
+	On most NVIDIA GPUs, data accesses are coalesced when multiple requests are made for memory locations from the same cache line in global memory \cite{cudamanual}.
 Appendix \ref{coalescedAMultiply} contains the \texttt{coalescedAMultiply} kernel, which coalesces accesses to matrix $A$ by storing \emph{tiles}, or square blocks, of $A$'s values into shared memory.
@@
 
-	The code is further optimized by improving how the previous \texttt{coalescedAMultiply} kernel handles accesses to Matrix $B$.
-While each thread in coalescedAMultiply accesses only a single element of matrix $A$ from global memory, each thread accesses a full column of matrix $B$.
-The repeated reads to elements of matrix $B$ across multiple threads can be alleviated by reading tiles from matrix $B$ into shared memory.
-Appendix \ref{coalescedABMultiply} contains the \texttt{coalescedAB} kernel which performs this optimization.
+	The code is further optimized by improving how the previous \texttt{coalescedAMultiply} kernel reads from Matrix $B$.
+	While each thread in coalescedAMultiply reads only a single element of matrix $A$ from global memory, each thread reads a full column of matrix $B$.
+	The repeated reads of elements of matrix $B$ across multiple threads can be alleviated by reading tiles from matrix $B$ into shared memory.
+	Appendix~\ref{coalescedABMultiply} contains the \texttt{coalescedAB} kernel which performs this optimization.
 
@@
 The \texttt{alignedABT} kernel improves the alignment of shared memory tiles in \texttt{coalescedABT}.
-An arbitrary large alignment value of 4096 is chosen as it is larger than the size of modern hardwares' cache lines and attributed to the local array declarations in the existing code as shown below.
+	An arbitrary large alignment value of 4096 is chosen as it is larger than the cache line size of modern hardware and attributed to the local array declarations in the existing code as shown below.
 
@@
 
-	All the examples above are an incomplete listing of optimization strategies developers can perform to code targeted at NVIDIA GPUs.
-We analyse the correlations between the collected AIWC metrics for each of these kernels and our conceptual understanding of GPU memory optimizations.
-This allows us to examine how effectively our methodology uncovers underlying bottlenecks and guides optimization efforts.
+	All the examples above are an incomplete listing of optimization strategies developers can apply to code targeted at NVIDIA GPUs.
+	In section~\ref{results}, we will compare the collected AIWC metrics with the expected effect of each optimization, to examine how effectively our methodology uncovers underlying bottlenecks and guides optimization efforts.
 

### 2020-01-19 paper/paper.tex

--- 2020-01-19 paper/paper.tex
@@
 %\editor{John B. Appleseed}
-\title[Architecture-Independent Memory Access Pattern Analysis]{Characterising Optimizations to Memory Access Patterns using Architecture-Independent Program Features}
+\title[Architecture-Independent Memory Access Pattern Analysis]{Characterizing Optimizations to Memory Access Patterns using Architecture-Independent Program Features}
 
@@
 	HPC developers are faced with the challenge of optimizing OpenCL workloads for high performance on novel architectures.
-	The Architecture-Independent Workload Characterization (AIWC) tool is a plugin for the Oclgrind OpenCL simulator that gathers metrics of OpenCL programs used to understand and predict program performance on an arbitrary given hardware architecture.
+	The Architecture Independent Workload Characterisation (AIWC) tool is a plugin for the Oclgrind OpenCL simulator that gathers metrics of OpenCL programs that can be used to understand and predict program performance on an arbitrary given hardware architecture.
 	However, AIWC metrics are not always easily interpreted and do not reflect some important memory access patterns affecting efficiency across architectures.
-	We propose the concept of parallel spatial locality: the locality of memory accesses by OpenCL work-items (threads) executing simultaneously.
-	We implement the parallel spatial locality metric in the AIWC framework and analyse gathered results on matrix multiply and the Extended OpenDwarfs OpenCL benchmarks.
-	The differences in the parallel spatial locality metric between implementations of matrix multiply reflect the optimizations performed.
-	The new metric showcases the memory access patterns in the OpenDwarfs benchmarks that affect their performance on various architectures.
+	We propose the concept of parallel spatial locality -- the closeness of memory accesses simultaneously issued by OpenCL work-items (threads).
+	We implement the parallel spatial locality metric in the AIWC framework, and analyse gathered results on matrix multiply and the Extended OpenDwarfs OpenCL benchmarks.
+	The differences in the obtained parallel spatial locality metric across implementations of matrix multiply reflect the optimizations performed.
+	The new metric can be used to distinguish between the OpenDwarfs benchmarks based on the memory access patters affecting their performance on various architectures.
 	%Current methods of analysing program behaviour have largely relied on performance counters on various architectures to provide architecture dependent metrics of code behaviour. 
@@
 
-	The OpenCL programming language is designed to support portability for heterogeneous systems by allowing a single code to execute on multiple hardware targets. 
+	The OpenCL programming language is designed to support modern HPC software engineers in writing code that executes on multiple hardware targets. 
 	%It is possible for an engineer to write highly thread-divergent logical analysis code aimed at traditional CPUs; parallelisable data crunching kernels aimed at GPUs; and code to satisfy specialized computational requirements aimed at FPGAs -- all in a single code base using OpenCL.
-	The language provides support for individual \emph{kernels}, portions of parallelizable codes that make up a HPC code base, to be executed independently on diverse hardware targets such as traditional CPUs, GPUs and FPGAs \cite{khronosopenclworkinggroup2011opencl}.
+	%The language provides support for individual \emph{kernels}, portions of parallelisable codes that make up a HPC code base, to be executed independently on diverse hardware targets such as traditional CPUs, GPUs and FPGAs \cite{khronosopenclworkinggroup2011opencl}.
 	This gives HPC software developers greater flexibility by allowing codes aimed at a range of hardware targets to be written in a single programming language environment.
@@
 
-	\subsection{Optimising handling of shared memory}
+	\subsection{Optimizing handling of shared memory}
 
@@
 	Figure \ref{fig: matmul runtime} shows that across varying problem sizes, the vendor recommended optimizations to the matrix multiplication code lead to increased performance. 
-Note that a matrix size of 80 is small -- overheads attached to optimization techniques may be causing the observed decrease in performance as some optimizations are performed.
+	%Note that a matrix size of 80 is small -- overheads attached to optimization techniques may be causing the observed decrease in performance as some optimizations are performed.
 
@@
 We observe that calculation of memory address entropy does not make use of temporal information.
-The order of sequential memory accesses performed by each thread, as well as addresses requested in parallel by multiple threads in an OpenCL thread group, are both vital in accurately characterising parallel codes.
+The order of sequential memory accesses performed by each thread, as well as addresses requested in parallel by multiple threads in an OpenCL thread group, are both vital in accurately characterizing parallel codes.
 
@@
 
-	There are three steps involved in generating an AIWC metric: \textit{recording}, \textit{calculating} and \textit{summarising} data collected from the symbolic execution of the kernel under inspection.
+	There are three steps involved in generating an AIWC metric: \textit{recording}, \textit{calculating} and \textit{summarizing} data collected from the symbolic execution of the kernel under inspection.
 
@@
 It accurately depicts a steeper drop-off for the more optimized \texttt{coalescedABT} kernel, where a larger proportion of parallel memory accesses make better use of the memory bank structure of GPU shared memory than all previous kernels.
-This is a significant improvement over the state-of-the-art AIWC metrics in characterising how codes localize shared memory accesses to better use the hardware provided.
+This is a significant improvement over the state-of-the-art AIWC metrics in characterizing how codes localize shared memory accesses to better use the hardware provided.
 
-	\section{Results}
+	The Extended OpenDwarfs benchmark suite is a set of OpenCL workload classified according to the computation patterns or Dwarfs. 
+	First presented in 2006, the 13 Berkeley Dwarfs are common computational and communication patterns that aim to capture the landscape of parallel computing workloads. 
+	The Extended OpenDwarfs benchmark suite presents a set of highly diverse workloads that each fit the programming paradigm of one of the 13 Berkeley Dwarfs.
+	The workloads presented are not optimized for any specific architecture -- hence optimizations using OpenCL local memory (which translates to CUDA shared memory) are not performed. 
+	Many of these benchmarks can be run with up to four \emph{problem sizes} based on the sizes of caches found in modern CPU memory heirarchies \cite{johnston18opendwarfs}.
+	We present the results of running the AIWC tool on selected benchmarks from the Extended OpenDwarfs benchmark suite \cite{opendwarfs2020head,johnston18opendwarfs} with a problem size setting of \emph{medium}, except for the GEM benchmark, which is run at a setting of \emph{tiny}.
+	For benchmarks such as BFS with multiple kernel invocations per run, the AIWC parallel spatial locality metrics of the invocation with the highest executed LLVM-SPIR instructions is presented for analysis.
+Note that the presented benchmarks use 32 bit numbers (OpenCL \texttt{int} and \texttt{float} types), and so dropping up to two bits of the memory addresses accessed will not change the data entropy, since any addresses accessed are at least 4 bytes apart.
 
-	The Extended OpenDwarfs benchmark suite is a set of OpenCL workload classified according to the computation patterns or Dwarfs. First presented in 2006, the Dwarfs categorize 
-	We present the results of running the AIWC tool on the Extended OpenDwarfs benchmark suite \cite{opendwarfs2017base,extendedopendwarfs}.
+	%TODO: Add some info about EOD benchmarks here?
 
@@
 
+	Performance results \cite{johnston18opendwarfs, krommydas2016opendwarfs} show that GPUs perform significantly better than CPUs for GEM, as memory unit stalls are at low levels for both CPUs and GPUs due to the highly efficient memory utilization of this benchmark.
+	As memory operations do not present a major slowdown, increased FPU compute capability of GPUs leads to far better performance on GPUs than CPUs \cite{krommydas2016opendwarfs}.
+
 	\subsection{Dense Linear Algebra: Lower-upper decomposition (LUD)}
@@
 	LUD in OpenDwarfs is a program to decompose an input $N\times N$ matrix as the product of one lower and one upper diagonal matrix. 
-LUD is a Memory access patterns in a dense linear algebra workload such as LUD are typically highly regular and deterministic for each OpenCL work item, based on the the matrix dimension and offset parameters.
-The OpenDwarfs implementation of LUD \cite{opendwarfs2017base} splits the LUD computation into three kernels: LUD perimeter and diagonal kernels spawn work-items in a single work dimension while LUD internal performs work in both dimensions of the input matrix.
+	Memory access patterns in a dense linear algebra workload such as LUD are typically highly regular and deterministic for each OpenCL work item, based on the matrix dimension and offset parameters.
+The OpenDwarfs implementation of LUD \cite{opendwarfs2020head} splits the LUD computation into three kernels: \texttt{LUD perimeter} and \texttt{LUD diagonal} kernels spawn work-items in a single work dimension; while \texttt{LUD internal} performs work in both dimensions of the input matrix.
 
-	LUD perimeter and diagonal partially benefit from memory access coalescing, for the lines of code where all the work-items access contiguous memory along a row of the input matrix.
+	The LUD kernels partially benefit from memory access coalescing on GPUs, for the lines of code where all the work-items access contiguous memory along a row of the input matrix. 
 	However, when threads parallelly access a column of the input matrix, multiple memory requests are made as addresses accessed are too distant to be coalesced into a single memory transaction on GPUs.
-This is reflected in the slow drop in parallel spatial locality for LUD perimeter and diagonal kernels.
-
-	LUD internal %TODO: 2Dimensional work -- better parallelism
+	This is reflected in the drop in parallel spatial locality for the LUD kernels to approximately half the value at 0 bits skipped.
+Taking \texttt{LUD internal} as an example, the 0-bits skipped parallel spatial locality metric is 4.101, while the 10-bits skipped parallel spatial locality metric is 2.053.
+	The swift decline of the metric to 2.053 from 2 to 6-bits skipped parallel spatial locality indicates that approximately half the parallel memory accesses in \texttt{LUD internal} are highly localized -- where work-items simultaneously access contiguous memory along a matrix row.
+	Similar to the presented matrix multiply codes, LUD can be optimized for GPUs by effective utilization of shared memory \cite{opendwarfs2020head}.
 
@@
 This rightly indicates that memory addresses accessed at each logical timestamp are very distant.
-On GPUs this will translate to poor utilization of both memory access coalescing and caching.
-This trend in the parallel spatial locality metric suggests that a possible improvement for GPU performance would be to load blocks of the input matrix into on-chip local memory to reduce the number of global memory requests.
+On GPUs this will translate to poor utilization of both memory access coalescing and caching \cite{krommydas2016opendwarfs}.
+This trend in the parallel spatial locality metric suggests that a possible improvement for GPU performance would be to load blocks of the input matrix into on-chip local memory to reduce the number of global memory requests -- this is typically performed in GPU optimized implementations of the Needleman-Wunsch algorithm \cite{opendwarfs2020head}.
 
@@
 
-	SRAD removes locally correlated noise from images by following a repeated grid update computation pattern.
+	SRAD removes locally correlated noise from images by following a repeated grid update computation pattern on image pixel grids. 
 	Conditional statements in the code cause thread divergence, potentially within a work-group, to handle boundary conditions. 
 	These boundaries constitute a small portion of the executed work-items, especially on larger data-sets and so the effective thread divergence is minimal.
-Thus, memory access patterns in SRAD are statically determined and relatively localized.
-The observed rapid drop-off in parallel spatial locality of the SRAD kernels (figure \ref{fig: opendwarfs parallel spatial locality}) is due to the highly localized nature of parallel memory accesses.
-At each point in the kernels' memory access profile, each work-item $(i,j)$ in an OpenCL work-group accesses the $(i,j)^{th}$ elements of various matrices.
-The sequential memory access pattern is non-linear since different matrices are accessed consecutively by the kernel, prohibiting ideal caching.
-However, memory requests made parallelly are highly localized, allowing memory access coalescing on GPUs, greatly reducing the number of memory requests made on each line of the kernel code on GPUs.
-Thus while cache hit rates are typically low on this benchmark, particularly on GPUs \cite{krommydas2016opendwarfs}, GPUs can hide the latency of global memory accesses through coalescing.
+	Memory access patterns in SRAD are statically determined and relatively localized. 
+	Similar to the LUD and \texttt{simpleMultiply} kernels, both SRAD kernels observe a rapid drop-off in parallel spatial locality (figure \ref{fig: opendwarfs parallel spatial locality}) as the number of bits skipped is increased, with the metric stabilizing at approximately half its original value when 0 bits are skipped.
+
+	At each point in the kernels' memory access profile, each work-item with global ID $(i,j)$ in an OpenCL work-group accesses the $(i,j)^{th}$ elements of various matrices.
+The sequential memory access pattern is non-linear since different matrices are accessed consecutively by the kernel, prohibiting ideal caching \cite{krommydas2016opendwarfs}.
+However, memory requests made simultaneously by a work-group always fall within a rectangular block of one of the matrices.
+	This allows memory access coalescing on GPUs for OpenCL work-items accessing contiguous matrix data, greatly reducing the number of memory requests made on each line of the kernel code on GPUs. 
+	However, work-items accessing data along a column of the matrix do not observe memory access coalescing.
+	Thus we observe that while cache hit rates are typically low on this benchmark, particularly on GPUs \cite{krommydas2016opendwarfs}, GPUs can hide the latency of global memory accesses through memory access coalescing to some extent.
 
@@
 	Graph traversal algorithms require pointer chasing operations to traverse nodes of a graph and perform calculations. 
-The Breadth First Search implemented in OpenDwarfs calls OpenCL kernels to traverse nodes immediately connected to the list of nodes at each depth starting from the root node.
-BFS is characterized by an imbalanced workload per kernel launch depending on the degree of the nodes being operated on -- with only a proportion of launched nodes performing meaningful work.
-As such the AIWC features collected for each kernel invocation vary significantly.
-The parallel spatial locality metric collects the entropy of memory accesses at each time step, and workload imbalances are dealt with by averaging entropy scores across all the time steps in the execution of an OpenCL work-group.
+	The Breadth First Search implemented in OpenDwarfs calls two OpenCL kernels to traverse nodes immediately connected to the list of nodes at each depth starting from the root node. 
+	BFS is characterized by an imbalanced workload per kernel launch depending on the degree of the nodes being operated on, with only a proportion of launched nodes performing meaningful work. 
+	As such, the AIWC features collected for each kernel invocation vary significantly. 
+	The parallel spatial locality metric collects the entropy of memory accesses at each time step, and workload imbalances across work-items are dealt with by averaging entropy scores across all the time steps in the execution of an OpenCL work-group.
 
-	The memory access patterns in BFS are irregular.
+	Of the two BFS kernels, \texttt{BFS kernel 1} performs the graph traversal necessary to generate a list of neighbours at each node. The memory access patterns in \texttt{BFS kernel 1} are irregular. 
 	Work-items fetch discontinuous memory locations attributed to any particular node in the graph, depending on the connectivity of the node they operate on. 
 	The program structure involves multiple levels of pointer-chasing and thus the precise memory address accessed are runtime dependent. 
-This leads to poor parallel spatial locality.
-Thus, we see a slow dropoff in parallel spatial locality for BFS.
+	This leads to poor parallel spatial locality. Thus, we see a slow dropoff in parallel spatial locality for \texttt{BFS kernel 1}.
 
@@
 
-	\subsection{Finite State Machine: Temporal Data Mining (TDM)}
-
-	%TODO
-
 	\section{Conclusions and Future Work} \label{future work}
 
-	To the best of our knowledge this work is the first to critically examine architecture independent methods to measure memory access patterns and guide hardware-specific optimization.
+	To the best of our knowledge this work is the first to critically examine AIWC's measurement of memory access patterns to guide hardware-specific optimization. 
 	We implemented and justified the addition of relative local memory usage to AIWC in our steps to characterize memory-based GPU code optimizations.
-We have also proposed the methodology to implement a new parallel spatial locality metric to AIWC.
-Upon implementation, our work would be the first to measure memory access patterns of parallel codes in an architecture-independent, easily interpreted manner.
+	We proposed the concept of parallel spatial locality to capture the idea of closeness of memory accesses made by parallel OpenCL workloads.
+	We then implemented a new parallel spatial locality metric to AIWC, presented results of running AIWC on matrix multiply kernels and selected OpenDwarfs benchmarks and validated our methodology. 
+	Our work is the first to measure memory access patterns of parallel codes in an architecture-independent, easily interpreted manner.
+%Josh, Beau: is this too bold a claim?
+If unnecessary, please remove.
 	%TODO UPDATE IN FUTURE
@@
 
-	Future work will provide a concrete implementation of this new parallel spatial locality metric in the AIWC framework.
-It is then feasible to test these new metrics across a larger set of GPU targeted OpenCL codes.
-This will allow a critical examination of the effectiveness of both proposed metrics in accurately profiling memory access patterns of these diverse codes.
+	Our work focused on measuring memory access patterns in OpenCL codes intends to guide optimization efforts.
+Another potential use case for AIWC is providing architecture-independent performance predictions for OpenCL kernels by generating machine learning models based on AIWC metrics \cite{beauperformanceprediction}.
+Future work would modify the presented metrics and develop new memory metrics with the specific intent of being fed to machine learning models to predict kernel performance on arbitrary and novel architectures.
+%FIXME: I want to say that we have been focusing on _guiding optimization_.
+We can do more work in making the AIWC metrics better for _performance prediction_.
+Is this coming across okay?
 

### 2020-01-19 paper/paper.tex
Finalising paper. Please review and make final edits

@@
 	The differences in the obtained parallel spatial locality metric across implementations of matrix multiply reflect the optimizations performed.
-	The new metric can be used to distinguish between the OpenDwarfs benchmarks based on the memory access patters affecting their performance on various architectures.
+	The new metric can be used to distinguish between the OpenDwarfs benchmarks based on the memory access patterns affecting their performance on various architectures.
 	%Current methods of analysing program behaviour have largely relied on performance counters on various architectures to provide architecture dependent metrics of code behaviour. 
@@
 	In section~\ref{related work} we discuss how AIWC and its precursors profiled memory access patterns in an architecture-independent fashion, and relevant architecture dependent approaches to memory access profiling.
-	In section \ref{method} we present a methodology to critically examine the architecture-independent workload characterization tool, which we use in section~\ref{results} to measure the accuracy of AIWC in profiling the impact of performance optimizations to various OpenCL codes.
-	In section~\ref{discussion} we suggest new methods to accurately capture code patterns correlating to hardware-specific memory optimizations.
+	In section \ref{method} we present a methodology to critically examine the architecture-independent workload characterization tool, which we use in section~\ref{preliminary results} to measure the accuracy of AIWC in profiling the impact of performance optimizations to various OpenCL codes.
+	In section~\ref{discussion} we discuss these preliminary results, and propose and implement our parallel spatial locality metric into the AIWC framework.
+	In section~\ref{results} we present the collected metric for the matrix multiplication kernels and selected OpenDwarfs benchmarks to validate our methodology in this paper.
 	Finally, in section~\ref{future work} we discuss further avenues to extend our work and conclude.
@@
 	On most NVIDIA GPUs, data accesses are coalesced when multiple requests are made for memory locations from the same cache line in global memory \cite{cudamanual}.
-Appendix \ref{coalescedAMultiply} contains the \texttt{coalescedAMultiply} kernel, which coalesces accesses to matrix $A$ by storing \emph{tiles}, or square blocks, of $A$'s values into shared memory.
+Appendix \ref{coalescedAMultiply} contains the \texttt{coalescedAMultiply} kernel, which coalesces accesses to matrix $A$ by storing \emph{tiles}, or square blocks, of $A$'s values into shared memory (OpenCL \texttt{\_\_local} memory).
 
@@
 The code in \texttt{coalescedAB} kernel is further optimized by implicitly transposing tiles of Matrix $A$ while loading from global memory.
-This improves memory bank utilization during reads to shared memory tiles are performed during the necessary computation.
+This improves memory bank utilization during reads to shared memory.
 The \texttt{coalescedABT} kernel demonstrates this optimization by modifying lines 20,21 of \texttt{coalescedAB} to:
@@
 AIWC measures memory based metrics to characterize the target code's memory behaviours.
-AIWC firstly collects aggregate simple counts such as the (i) total memory footprint, the total number of unique addresses accessed; and (ii) 90\% memory footprint, the number of unique addresses covering 90\% of memory accesses.
+AIWC first collects aggregate simple counts such as the (i) total memory footprint, the total number of unique addresses accessed; and (ii) 90\% memory footprint, the number of unique addresses covering 90\% of memory accesses.
 A small ratio of 90\% memory footprint to total memory footprint correlates to programs accessing the same small subset of memory addresses repeatedly -- which can be linked to increased performance in a cached memory hierarchy.
@@
 
-	Our approach is an architecture independent analysis of memory access patterns that provides metrics correlating to similar performance critical memory access optimizations as CuMAPz.
+	Our approach is an architecture independent analysis of memory access patterns to provide metrics correlating to similar performance critical memory access optimizations as CuMAPz.
 We aim to further the state of the art by providing a framework to guide developers in optimizing OpenCL codes for any given target architecture.
@@
 
-	To provide memory access pattern metrics that represent vendor recommended optimizations on GPUs, we improved AIWC by adding a new metric: relative local memory usage.
+	To provide memory access pattern metrics that represent vendor recommended optimizations on GPUs, we improved AIWC by adding our first new metric: relative local memory usage.
 This measures the proportion of all memory accesses from the symbolic execution of the kernel that occurred to memory allocated as \texttt{\_\_local}.
-On GPUs, this memory address space is mapped to fast on-chip shared memory.
+On NVIDIA GPUs, this memory address space is mapped to fast on-chip shared memory.
 
@@
 	\textit{Global and local MAE}: Entropy measurements decrease from \texttt{simpleMultiply} to \texttt{coalescedAB}.
-This is explained by the optimizations performed incrementally reducing accesses to matrix $A$ and $B$, and instead performing the bulk of computations on smaller tiles in on-chip shared memory.
+This is explained by the optimizations performed incrementally reducing accesses to matrix $A$ and $B$, and instead performing the bulk of computations on smaller tiles in local memory.
 We observe an almost completely uniform distribution of memory accesses in \texttt{simpleMultiply}, where the program makes $N$ loads to each element of $A$ and $B$ with $N$ the dimension of the matrices.
-The frequency distribution of memory addresses accessed become increasingly non-uniform as we perform less accesses to matrices $A$ and $B$ and offset this by increasing the frequency of accesses to smaller, more localized shared memory buffers.
-The non-uniformity of relative frequency with which each memory cell is accessed by the program translates directly to decreases in local and global memory entropy values as reliance on small shared memory buffers increases from \texttt{simpleMultiply} to \texttt{coalescedAB}.
+The frequency distribution of memory addresses accessed become increasingly non-uniform as we perform less accesses to matrices $A$ and $B$ and offset this by increasing the frequency of accesses to smaller, more localized local memory buffers.
+The non-uniformity of relative frequency with which each memory cell is accessed by the program translates directly to decreases in local and global memory entropy values as reliance on small local memory buffers increases from \texttt{simpleMultiply} to \texttt{coalescedAB}.
 
 	\textit{Memory Footprint}: Similar to trends in global and local MAE, we find that the ratio of 90\% memory footprint to total memory footprint decreases from $60.12\%$ for \texttt{simpleMultiply} to $0.25\%$ (\texttt{coalescedAB}).
-Increased utilization of shared memory in the optimized kernels means that the shared memory buffers make up a greater proportion of total memory accesses in the program.
-As the shared memory buffers are small and reused within a workgroup, the memory footprint of the shared memory buffers is also smaller.
+Increased utilization of local memory in the optimized kernels means that the local memory buffers make up a greater proportion of total memory accesses in the program.
+As the local memory buffers are small and reused within a workgroup, the memory footprint of the local memory buffers is also smaller.
 
 	AIWC's metrics strongly reward optimizations that tend to localize memory accesses.
-Shared memory buffers are typically smaller than global memory arrays when programming for GPUs due to hardware limitations on sizes of shared memory buffers \cite{cudamanual}.
+Local memory buffers are typically smaller than global memory arrays when programming for GPUs due to hardware limitations on sizes of shared memory \cite{cudamanual}.
 However, the metrics currently measured by AIWC do not have a direct causal relation to code patterns that optimize memory accesses on GPUs.
@@
 		\end{tikzpicture}
-		\caption{Calculated parallel spatial locality metric for matrix multiply kernels for 256 $\times$ 256 matrix multiplication}
+		\caption{Parallel spatial locality metric obtained from AIWC for matrix multiply kernels for 256 $\times$ 256 matrix multiplication}
 		\label{fig: matmul parallel spatial locality}
@@
 t_{last}$, calculate the n-bits dropped entropy of memory addresses accessed by all threads in a work group within the timestamp $t$.
-Here $n$ can range between 1 and 10 as was the case for $LMAE$.
+Here $n$ can range between 0 and 10 as was the case for $LMAE$.
 
@@
 	In parallel programs, to accurately measure spatial locality of accesses, we must consider memory accesses performed by multiple threads in a close temporal scope. 
-	The proposed metric calculates the locality of accesses in each time step of the program's execution and steeper drop-offs in $n$-bits dropped parallel spatial locality scores will be observed in programs that access nearby memory addresses within the same timestamp.
+	The proposed metric calculates the locality of accesses in each time step of the program's execution and steeper drop-offs in $n$-bits dropped parallel spatial locality scores will be observed in programs that often access nearby memory addresses within the same timestamp.
 	Such programs will perform better on GPUs, as they will make better use of both global memory access coalescing and shared memory bank structures.
@@
 In particular we observe that the \texttt{coalescedABT} and \texttt{alignedABT} kernels have the steepest drop-offs in entropy as the number of bits skipped is increased, which correlates to better locality of parallel memory accesses.
-It is expected for these kernels to exhibit better parallel spatial locality than \texttt{simpleMultiply}, as \texttt{coalescedABT} and \texttt{alignedABT} make use of shared memory, where accesses tend to be localized simply due to the small size of shared on-chip memory typically available on GPUs.
+It is expected for these kernels to exhibit better parallel spatial locality than \texttt{simpleMultiply}, as \texttt{coalescedABT} and \texttt{alignedABT} make use of local memory, where accesses tend to be localized simply due to the small size of shared on-chip memory typically available on GPUs.
 Further we find that the proposed metric successfully distinguishes between the \texttt{coalescedAB} and \texttt{coalescedABT} kernels.
 It accurately depicts a steeper drop-off for the more optimized \texttt{coalescedABT} kernel, where a larger proportion of parallel memory accesses make better use of the memory bank structure of GPU shared memory than all previous kernels.
-This is a significant improvement over the state-of-the-art AIWC metrics in characterizing how codes localize shared memory accesses to better use the hardware provided.
+This is a significant improvement over the state-of-the-art AIWC metrics in characterizing how codes localize simultaneous memory accesses to better use the hardware provided.
 
@@
 	We present the results of running the AIWC tool on selected benchmarks from the Extended OpenDwarfs benchmark suite \cite{opendwarfs2020head,johnston18opendwarfs} with a problem size setting of \emph{medium}, except for the GEM benchmark, which is run at a setting of \emph{tiny}.
-	For benchmarks such as BFS with multiple kernel invocations per run, the AIWC parallel spatial locality metrics of the invocation with the highest executed LLVM-SPIR instructions is presented for analysis.
-Note that the presented benchmarks use 32 bit numbers (OpenCL \texttt{int} and \texttt{float} types), and so dropping up to two bits of the memory addresses accessed will not change the data entropy, since any addresses accessed are at least 4 bytes apart.
+	For benchmarks such as BFS with multiple kernel invocations per run, the AIWC parallel spatial locality metrics of the invocation with the highest executed LLVM-SPIR instructions are presented for analysis.
+Note that the presented benchmarks use 32 bit numbers (OpenCL \texttt{int} and \texttt{float} types), and so dropping up to 2 bits of the memory addresses accessed will not change the data entropy, since any addresses accessed are at least 4 bytes apart.
 
@@
 		legend entries={\texttt{GEM}, \texttt{NW}, \texttt{SRAD Kernel 1},\texttt{SRAD Kernel 2},
-			\texttt{CSR}, \texttt{BFS Kernel 1}, \texttt{BFS Kernel 2}, \texttt{LUD Diagonal}, \texttt{LUD Perimeter}, \texttt{LUD Internal}, \texttt{TDM}},
+			\texttt{CSR}, \texttt{BFS Kernel 1}, \texttt{BFS Kernel 2}, \texttt{LUD Diagonal}, \texttt{LUD Perimeter}, \texttt{LUD Internal}},
 		legend style={at={(0.5,-0.15)},anchor=north}
@@
 	Performance results \cite{johnston18opendwarfs, krommydas2016opendwarfs} show that GPUs perform significantly better than CPUs for GEM, as memory unit stalls are at low levels for both CPUs and GPUs due to the highly efficient memory utilization of this benchmark.
-	As memory operations do not present a major slowdown, increased FPU compute capability of GPUs leads to far better performance on GPUs than CPUs \cite{krommydas2016opendwarfs}.
+	As memory operations do not present a major slowdown, superior FPU compute capability of GPUs leads to far better performance on GPUs than CPUs \cite{krommydas2016opendwarfs}.
 
@@
 	The LUD kernels partially benefit from memory access coalescing on GPUs, for the lines of code where all the work-items access contiguous memory along a row of the input matrix. 
-	However, when threads parallelly access a column of the input matrix, multiple memory requests are made as addresses accessed are too distant to be coalesced into a single memory transaction on GPUs.
+	However, when threads simultaneously access a column of the input matrix, multiple memory requests are made as addresses accessed are too distant to be coalesced into a single memory transaction on GPUs.
 	This is reflected in the drop in parallel spatial locality for the LUD kernels to approximately half the value at 0 bits skipped.
@@
 This rightly indicates that memory addresses accessed at each logical timestamp are very distant.
-On GPUs this will translate to poor utilization of both memory access coalescing and caching \cite{krommydas2016opendwarfs}.
+On GPUs this translates to poor utilization of both memory access coalescing and caching \cite{krommydas2016opendwarfs}.
 This trend in the parallel spatial locality metric suggests that a possible improvement for GPU performance would be to load blocks of the input matrix into on-chip local memory to reduce the number of global memory requests -- this is typically performed in GPU optimized implementations of the Needleman-Wunsch algorithm \cite{opendwarfs2020head}.
@@
 
-	Of the two BFS kernels, \texttt{BFS kernel 1} performs the graph traversal necessary to generate a list of neighbours at each node. The memory access patterns in \texttt{BFS kernel 1} are irregular. 
+	Of the two BFS kernels, it is \texttt{BFS kernel 1} that performs the graph traversal necessary to generate a list of neighbours at each node.
+The memory access patterns in \texttt{BFS kernel 1} are irregular.
 	Work-items fetch discontinuous memory locations attributed to any particular node in the graph, depending on the connectivity of the node they operate on. 
@@
 Three inputs are provided to the CSR kernel.
-Only non-zero elements of each row of a matrix are stored in $Ax$, along with separate arrays $Aj,\ Ap$ indicating the position of each non-zero element in the matrix.
+The non-zero elements of a matrix are stored in row-major order in $Ax$, along with separate arrays $Aj,\ Ap$ indicating the position of each non-zero element in the matrix.
 
@@
 	The locations of matrix data read by each work-item are dependent on the number and position of non-zero values in the sparse-matrix, which are decided by the values in $Aj$ and $Ap$. 
-	Similar to BFS, this means memory access patterns are runtime-dependent due to indirect addressing (pointer chasing).
+	Similar to BFS, this means memory access patterns are runtime-dependent due to indirect addressing.
 	This pattern of indirect addressing is typical of applications in the Spare Linear Algebra Dwarf, which severely hinders locality of memory accesses performed.
-The parallel spatial locality metric reflects this trend.
-Figure \ref{fig: opendwarfs parallel spatial locality} shows the very gradual decline in parallel spatial locality as the number of bits dropped is increased.
+The collected parallel spatial locality metric reflects this trend.
+Figure \ref{fig: opendwarfs parallel spatial locality} shows the gradual decline in parallel spatial locality as the number of bits dropped is increased.
 
@@
 	To the best of our knowledge this work is the first to critically examine AIWC's measurement of memory access patterns to guide hardware-specific optimization. 
-	We implemented and justified the addition of relative local memory usage to AIWC in our steps to characterize memory-based GPU code optimizations.
-	We proposed the concept of parallel spatial locality to capture the idea of closeness of memory accesses made by parallel OpenCL workloads.
-	We then implemented a new parallel spatial locality metric to AIWC, presented results of running AIWC on matrix multiply kernels and selected OpenDwarfs benchmarks and validated our methodology. 
-	Our work is the first to measure memory access patterns of parallel codes in an architecture-independent, easily interpreted manner.
+	We implemented and justified the addition of relative local memory usage to AIWC to help characterize memory-based GPU code optimizations.
+	We proposed the concept of parallel spatial locality to capture the idea of closeness of memory accesses made by parallel OpenCL workloads, and implemented a new parallel spatial locality metric to AIWC.
+	We ran the modified AIWC tool on matrix multiply kernels and selected OpenDwarfs benchmarks, presenting results and analysis to validate our methodology. 
+	%Our work is the first to measure memory access patterns of parallel codes in an architecture-independent, easily interpreted manner.
 %Josh, Beau: is this too bold a claim?
@@
 	Our proposed parallel spatial locality metric may also correlate to some memory-based optimization strategies for CPUs. 
-Future work would apply the approach followed by this paper to CPU and FPGA architectures to evaluate the viability of an architecture independent characterization of memory access patterns in codes targeted to these architectures.
+	Future work would apply the approach followed by this paper to optimizations for CPU and FPGA architectures to critically evaluate the viability of AIWC in analyzing memory access patterns in codes targeted to these architectures.
 
-	Our work focused on measuring memory access patterns in OpenCL codes intends to guide optimization efforts.
+	Our work in this paper intended to guide optimization efforts.
 Another potential use case for AIWC is providing architecture-independent performance predictions for OpenCL kernels by generating machine learning models based on AIWC metrics \cite{beauperformanceprediction}.
@@
 We can do more work in making the AIWC metrics better for _performance prediction_.
-Is this coming across okay?
+Is this coming across okay in this paragraph?
 

### 2020-01-19 paper/paper.tex
Finalising paper. Please review and make final edits

@@
 	For benchmarks such as BFS with multiple kernel invocations per run, the AIWC parallel spatial locality metrics of the invocation with the highest executed LLVM-SPIR instructions are presented for analysis.
-Note that the presented benchmarks use 32 bit numbers (OpenCL \texttt{int} and \texttt{float} types), and so dropping up to 2 bits of the memory addresses accessed will not change the data entropy, since any addresses accessed are at least 4 bytes apart.
+Note that the presented benchmarks use 32 bit numbers (OpenCL \texttt{int} and \texttt{float} types), and so dropping up to 2 bits of the memory addresses accessed will not change the parallel spatial locality, since any addresses accessed are at least 4 bytes apart.
 
@@
 Taking \texttt{LUD internal} as an example, the 0-bits skipped parallel spatial locality metric is 4.101, while the 10-bits skipped parallel spatial locality metric is 2.053.
-	The swift decline of the metric to 2.053 from 2 to 6-bits skipped parallel spatial locality indicates that approximately half the parallel memory accesses in \texttt{LUD internal} are highly localized -- where work-items simultaneously access contiguous memory along a matrix row.
+	The swift decline of the metric to 2.053 from 2 to 6-bits skipped parallel spatial locality indicates that approximately half the parallel memory accesses in \texttt{LUD internal} are highly localized.
+This occurs when work-items simultaneously access contiguous memory along a matrix row.
 	Similar to the presented matrix multiply codes, LUD can be optimized for GPUs by effective utilization of shared memory \cite{opendwarfs2020head}.

### 2020-01-19 paper/paper.tex
further editing pass

@@
 
-	The code is further optimized by improving how the previous \texttt{coalescedAMultiply} kernel reads from Matrix $B$.
-	While each thread in coalescedAMultiply reads only a single element of matrix $A$ from global memory, each thread reads a full column of matrix $B$.
-	The repeated reads of elements of matrix $B$ across multiple threads can be alleviated by reading tiles from matrix $B$ into shared memory.
+	The code is further optimized by improving the locality of reads from Matrix $B$.
+	While in the previous kernel each thread reads only a single element of matrix $A$ from global memory, each thread reads a full column of matrix $B$.
+	The repeated reads of elements of matrix $B$ can be shared between threads in a work group by reading tiles from matrix $B$ into shared memory.
 	Appendix~\ref{coalescedABMultiply} contains the \texttt{coalescedAB} kernel which performs this optimization.
 
-	\subsection{Optimizing handling of shared memory}
+	\subsection{Optimizing Handling of Shared Memory}
 
-	NVIDIA shared memory is divided into multiple banks -- stored in independent memory modules -- to allow parallel accesses to shared memory.
+	NVIDIA shared memory is divided into multiple banks -- stored in independent memory modules -- to allow parallel memory access.
 	Bank conflicts occur when shared memory in the same bank is accessed concurrently.
@@
 Further, in many cases, architecture dependent characteristics usually cannot be directly correlated to specific code patterns.
-For example, the causes of high cache miss rates in the execution of a program are too complex and widely variant on microarchitecture specific features such as cache size, prefetch behaviour and cache placement policies.
-In summary, we argue that a HPC developer tasked with optimizing code for a given hardware target would benefit from architecture-independent metrics of the code that can be conceptually correlated to hardware specific optimization patterns.
+	For example, the causes of high cache miss rates in the execution of a program are complex and depend on microarchitecture specific features such as cache size, prefetch behaviour and cache placement policies.
+	We argue that a HPC developer tasked with optimizing code for a given hardware target would benefit from architecture-independent metrics of the code that can be conceptually correlated to hardware specific optimization patterns.
 Such architecture-independent metrics will effectively guide the HPC developer's workflow in both finding and fixing bottlenecks in code performance.
+	% TODO this argument isn't very clear
 
@@
 	The MICA approach was tailored for single-threaded applications as the metrics collected rely heavily on Pin instrumentation \cite{luk2005pin}.
-As such, MICA is unsuited to analysing conventional supercomputing workloads with heavy use of parallelism.
+As such, MICA is unsuited to analysing HPC workloads with heavy use of parallelism.
 
@@
 
-	\textit{Global and local MAE}: Entropy measurements decrease from \texttt{simpleMultiply} to \texttt{coalescedAB}.
-This is explained by the optimizations performed incrementally reducing accesses to matrix $A$ and $B$, and instead performing the bulk of computations on smaller tiles in local memory.
+	\textit{Global and local MAE}: Entropy measurements decrease from \texttt{simpleMultiply} to \texttt{coalescedAB}, as the optimizations reduce the number of reads from matrices $A$ and $B$ in global memory, replacing these with reads from smaller tiles in local memory.
 	We observe an almost completely uniform distribution of memory accesses in \texttt{simpleMultiply}, where the program makes $N$ loads to each element of $A$ and $B$ with $N$ the dimension of the matrices.
-The frequency distribution of memory addresses accessed become increasingly non-uniform as we perform less accesses to matrices $A$ and $B$ and offset this by increasing the frequency of accesses to smaller, more localized local memory buffers.
-The non-uniformity of relative frequency with which each memory cell is accessed by the program translates directly to decreases in local and global memory entropy values as reliance on small local memory buffers increases from \texttt{simpleMultiply} to \texttt{coalescedAB}.
+	The distribution of memory accesses becomes increasingly non-uniform as we perform fewer accesses to matrices $A$ and $B$ and more accesses to smaller local memory buffers, resulting in decreases in local and global memory entropy values from \texttt{simpleMultiply} to \texttt{coalescedAB}.
 
@@
 
-	\section{Discussion } \label{discussion}
+	\section{A Parallel Spatial Locality Metric} \label{discussion}
 
@@
 
-	We propose a new architecture independent metric to measure memory access patterns of parallel programs.
+	We propose a new architecture-independent metric to measure memory access patterns of parallel programs.
 	The proposed metric is inspired by CuMAPz' direct approaches to measuring optimization specific characteristics of CUDA codes.
@@
 	AIWC currently collects a list of all memory accessed by each thread of execution.
-In the OpenCL programming model, threads within thread groups execute the same code and are allowed to cooperate on \textit{local memory} addresses.
-We can group together threads by their work groups to align histories of memory accesses of threads in a group to analyse memory accesses performed by thread groups at each logical time step in the symbolic execution of the code.
+	In the OpenCL programming model, threads within a work groups execute the same code and are allowed to cooperate on \textit{local memory} addresses.
+	We can group together memory accesses of threads in a work group at each logical time step in the symbolic execution of the code.
 	We argue that analysing the accesses occuring at the same logical time step is an accurate, conceptually sound method of predicting the extent of memory access coalescing and bank conflicts we are likely to find upon executing the code on a GPU.
-
-	The precise definition of the proposed metric is provided in the following section.
-
-	\subsection{A Parallel Spatial Locality Metric}
 
@@
 
-	\textit{Record}: we first record memory accesses performed by each thread in an OpenCL thread group as described above to achieve a global ordering of all memory accesses performed in a thread group.
-This ordering is collected in the form of logical timestamps ($t_0 ..
-t_{last}$) at which memory accesses occur.
+	\textit{Record}: we first record memory accesses performed by each thread in an OpenCL work group as described above to achieve a global ordering of all memory accesses performed by the group.
+	This ordering is collected in the form of logical timestamps ($t_0 .. t_{last}$) at which memory accesses occur.
 
@@
 
-	The proposed metric is a parallel computing analogue for MICA's data stride metric that measures the distance between consecutive data accesses in a single threaded environment. 
+	The proposed metric is a parallel computing analogue for MICA's data stride metric that measures the distance between consecutive data accesses in a single-threaded environment. 
 	In parallel programs, to accurately measure spatial locality of accesses, we must consider memory accesses performed by multiple threads in a close temporal scope. 
-	The proposed metric calculates the locality of accesses in each time step of the program's execution and steeper drop-offs in $n$-bits dropped parallel spatial locality scores will be observed in programs that often access nearby memory addresses within the same timestamp.
+	The proposed metric calculates the locality of accesses in each time step of the program's execution and steeper reductions in $n$-bits dropped parallel spatial locality scores will be observed in programs that often access nearby memory addresses within the same timestamp.
 	Such programs will perform better on GPUs, as they will make better use of both global memory access coalescing and shared memory bank structures.
-	To a lesser extent, the proposed metric captures performance-critical memory access patterns on CPUs, as pulling a single cacheline from global memory services the threads running on all CPU cores.
-%TODO: Josh, Beau -- Please check this paragraph.
+	To a lesser extent, the proposed metric reflects performance-critical memory access patterns on CPUs, as pulling a single cache line from global memory into last-level cache may improve memory access times for all CPU cores.
 
@@
 	In figure \ref{fig: matmul parallel spatial locality} we see the proposed parallel spatial locality metric obtained using the AIWC tool for each of the matrix multiplication kernels from section \ref{motivating example}.
-In particular we observe that the \texttt{coalescedABT} and \texttt{alignedABT} kernels have the steepest drop-offs in entropy as the number of bits skipped is increased, which correlates to better locality of parallel memory accesses.
+	In particular, we observe that the \texttt{coalescedABT} and \texttt{alignedABT} kernels have the steepest reductions in entropy as the number of bits skipped is increased, which correlates to better locality of parallel memory accesses.
 	It is expected for these kernels to exhibit better parallel spatial locality than \texttt{simpleMultiply}, as \texttt{coalescedABT} and \texttt{alignedABT} make use of local memory, where accesses tend to be localized simply due to the small size of shared on-chip memory typically available on GPUs.
 	Further we find that the proposed metric successfully distinguishes between the \texttt{coalescedAB} and \texttt{coalescedABT} kernels.
-It accurately depicts a steeper drop-off for the more optimized \texttt{coalescedABT} kernel, where a larger proportion of parallel memory accesses make better use of the memory bank structure of GPU shared memory than all previous kernels.
+	It accurately depicts a steeper reduction for the more optimized \texttt{coalescedABT} kernel, where a larger proportion of parallel memory accesses make better use of the memory bank structure of GPU shared memory than all previous kernels.
 	This is a significant improvement over the state-of-the-art AIWC metrics in characterizing how codes localize simultaneous memory accesses to better use the hardware provided.
@@
 	The workloads presented are not optimized for any specific architecture -- hence optimizations using OpenCL local memory (which translates to CUDA shared memory) are not performed. 
-	Many of these benchmarks can be run with up to four \emph{problem sizes} based on the sizes of caches found in modern CPU memory heirarchies \cite{johnston18opendwarfs}.
+	Many of these benchmarks can be run with up to four problem sizes based on the sizes of caches found in modern CPU memory hierarchies~\cite{johnston18opendwarfs}.
 	We present the results of running the AIWC tool on selected benchmarks from the Extended OpenDwarfs benchmark suite \cite{opendwarfs2020head,johnston18opendwarfs} with a problem size setting of \emph{medium}, except for the GEM benchmark, which is run at a setting of \emph{tiny}.
@@
 Each OpenCL work-item operates on a single surface vertex, finding the electrostatic potential generated by looping through every atom in the biomolecule in a global order.
-	The computation pattern is highly regular and memory accesses are perfectly synchronized. Atom data is accessed consecutively, with all work-items simultaneously accessing each single atom's data.
+	The computation pattern is highly regular and memory accesses are perfectly synchronized. Atom data is accessed consecutively, with all work-items simultaneously accessing each atom's data.
 	The parallel spatial locality metric reflects this pattern of efficient memory utilization (single loads from global memory servicing all OpenCL work-items).
@@
 This rightly indicates that memory addresses accessed at each logical timestamp are very distant.
-On GPUs this translates to poor utilization of both memory access coalescing and caching \cite{krommydas2016opendwarfs}.
+On GPUs, this translates to poor utilization of both memory access coalescing and caching \cite{krommydas2016opendwarfs}.
 This trend in the parallel spatial locality metric suggests that a possible improvement for GPU performance would be to load blocks of the input matrix into on-chip local memory to reduce the number of global memory requests -- this is typically performed in GPU optimized implementations of the Needleman-Wunsch algorithm \cite{opendwarfs2020head}.
@@
 
-	To the best of our knowledge this work is the first to critically examine AIWC's measurement of memory access patterns to guide hardware-specific optimization. 
-	We implemented and justified the addition of relative local memory usage to AIWC to help characterize memory-based GPU code optimizations.
-	We proposed the concept of parallel spatial locality to capture the idea of closeness of memory accesses made by parallel OpenCL workloads, and implemented a new parallel spatial locality metric to AIWC.
-	We ran the modified AIWC tool on matrix multiply kernels and selected OpenDwarfs benchmarks, presenting results and analysis to validate our methodology. 
+	To the best of our knowledge, this work is the first to propose the use of architecture-independent metrics of parallel memory access to guide hardware-specific optimization. 
+	We implemented and evaluated a new feature, relative local memory usage, to help characterize memory-based GPU code optimizations.
+	We also implemented a new parallel spatial locality metric to capture the idea of closeness of memory accesses made by parallel OpenCL workloads.
+	We ran the enhanced AIWC tool on matrix multiply kernels and selected OpenDwarfs benchmarks, presenting results and analysis to validate our methodology. 
 	%Our work is the first to measure memory access patterns of parallel codes in an architecture-independent, easily interpreted manner.
@@
 
-	Our work in this paper intended to guide optimization efforts.
-Another potential use case for AIWC is providing architecture-independent performance predictions for OpenCL kernels by generating machine learning models based on AIWC metrics \cite{beauperformanceprediction}.
+	While our work in this paper intended to guide optimization efforts, another potential use case for AIWC is providing architecture-independent performance predictions for OpenCL kernels by generating machine learning models based on AIWC metrics \cite{beauperformanceprediction}.
 Future work would modify the presented metrics and develop new memory metrics with the specific intent of being fed to machine learning models to predict kernel performance on arbitrary and novel architectures.

### 2020-01-19 paper/paper.tex
fix minor bib issues; move Fig 3 to top of corresponding page

@@
 	Our aim in this paper is to develop a framework to guide HPC software engineers in hardware-dependent code optimization -- specifically by guiding the improvement of memory access patterns.
-	We provide examples of manufacturer-recommended code optimizations to improve memory access patterns on the target architecture, and examine these using the architecture-independent workload characterization (AIWC)~\cite{beauaiwc} plugin for Oclgrind.
+	We provide examples of manufacturer-recommended code optimizations to improve memory access patterns on the target architecture, and examine these using the architecture-independent workload characterization (AIWC)~\cite{beauaiwc} plugin for Oclgrind~\cite{price:15}.
 	Our work highlights the benefits and challenges arising from an architecture-independent analysis of memory-based program characteristics.
@@
 	Multiple global memory loads and stores are coalesced into a single transaction when certain device-specific conditions are met. 
-	On most NVIDIA GPUs, data accesses are coalesced when multiple requests are made for memory locations from the same cache line in global memory \cite{cudamanual}.
-Appendix \ref{coalescedAMultiply} contains the \texttt{coalescedAMultiply} kernel, which coalesces accesses to matrix $A$ by storing \emph{tiles}, or square blocks, of $A$'s values into shared memory (OpenCL \texttt{\_\_local} memory).
+	On most NVIDIA GPUs, data accesses are coalesced when multiple requests are made for memory locations from the same cache line in global memory~\cite{cudamanual}.
+	Appendix~\ref{coalescedAMultiply} contains the \texttt{coalescedAMultiply} kernel, which coalesces accesses to matrix $A$ by storing \emph{tiles}, or square blocks, of $A$'s values into shared memory (OpenCL \texttt{\_\_local} memory).
 
@@
 
-	\section{Methodology} \label{method}
-
-	To provide memory access pattern metrics that represent vendor recommended optimizations on GPUs, we improved AIWC by adding our first new metric: relative local memory usage.
-This measures the proportion of all memory accesses from the symbolic execution of the kernel that occurred to memory allocated as \texttt{\_\_local}.
-On NVIDIA GPUs, this memory address space is mapped to fast on-chip shared memory.
-
-	Relative local memory usage is an example of a metric that is useful to measure performance critical access patterns on some architectures such as GPUs, and not others, such as CPUs.
-CPUs do not typically have a notion of user controlled on-chip memory shared between hardware threads such as NVIDIA GPUs' shared memory.
-This is a natural consequence of programming for a heterogeneous system.
-Specific code patterns may translate to performance improvements only on certain hardware.
-
-	In the original AIWC tool \cite{beau_johnston_2017_1134175}, global and local memory address entropy (MAE) were calculated using physical addresses of memory used by the Oclgrind simulator back-end of AIWC.
-This caused irregularities in entropy calculations across multiple runs of the same simulation.
-We improved the calculation of memory address entropy by using \textit{virtual addresses} to calculate MAE values using an abstract ideal address space on which all memory accesses by the kernel occur.
-This allows AIWC to accurately abstract over the hardware and ISA-specific differences in memory layouts across the diverse hardware targets.
-
-	\section{Preliminary Results} \label{preliminary results}
-
-	\begin{figure*}[h!]
+	\begin{figure*}[ht!]
 		\centering
@@
 
+	\section{Methodology} \label{method}
+
+	To provide memory access pattern metrics that represent vendor recommended optimizations on GPUs, we improved AIWC by adding our first new metric: relative local memory usage.
+This measures the proportion of all memory accesses from the symbolic execution of the kernel that occurred to memory allocated as \texttt{\_\_local}.
+On NVIDIA GPUs, this memory address space is mapped to fast on-chip shared memory.
+
+	Relative local memory usage is an example of a metric that is useful to measure performance critical access patterns on some architectures such as GPUs, and not others, such as CPUs.
+CPUs do not typically have a notion of user controlled on-chip memory shared between hardware threads such as NVIDIA GPUs' shared memory.
+This is a natural consequence of programming for a heterogeneous system.
+Specific code patterns may translate to performance improvements only on certain hardware.
+
+	In the original AIWC tool \cite{beau_johnston_2017_1134175}, global and local memory address entropy (MAE) were calculated using physical addresses of memory used by the Oclgrind simulator back-end of AIWC.
+This caused irregularities in entropy calculations across multiple runs of the same simulation.
+We improved the calculation of memory address entropy by using \textit{virtual addresses} to calculate MAE values using an abstract ideal address space on which all memory accesses by the kernel occur.
+This allows AIWC to accurately abstract over the hardware and ISA-specific differences in memory layouts across the diverse hardware targets.
+
+	\section{Preliminary Results} \label{preliminary results}
+
 	Figure \ref{fig: matmul runtime} shows that across varying problem sizes, the vendor recommended optimizations to the matrix multiplication code lead to increased performance. 
@@
 		LMAE \#bits=10                & 9.02   & 5.18       & 1.78        & 1.78         & 1.78       \\ %\hline
-		Relative Local Memory Usage   & 0      & 49.91    & 94.28       & 94.28        & 94.28      \\ \bottomrule
+		Relative Local Memory Usage   & 0      & 0.50       & 0.94       &  0.94         & 0.94      \\ \bottomrule
 	\end{tabular}
@@
 
-	\textit{Relative local memory usage}: As reliance on NVIDIA GPUs' shared memory increases in each kernel from \texttt{simpleMultiply} to \texttt{coalescedAB}, we find that the ratio of OpenCL local memory usage increases as expected.
+	\textit{Relative local memory usage}: As reliance on NVIDIA GPUs' shared memory increases in each kernel from \texttt{simpleMultiply} to \texttt{coalescedAB}, we find that the proportion of OpenCL local memory usage increases as expected.
 

### 2020-01-19 paper/paper.tex
Minor changes

--- 2020-01-19 paper/paper.tex
@@
 	All the examples above are an incomplete listing of optimization strategies developers can apply to code targeted at NVIDIA GPUs.
-	In section~\ref{results}, we will compare the collected AIWC metrics with the expected effect of each optimization, to examine how effectively our methodology uncovers underlying bottlenecks and guides optimization efforts.
+	In section~\ref{preliminary results}, we will compare the collected AIWC metrics with the expected effect of each optimization to examine how effectively our methodology uncovers underlying bottlenecks and guides optimization efforts.
 
@@
 
-	In the original AIWC tool \cite{beau_johnston_2017_1134175}, global and local memory address entropy (MAE) were calculated using physical addresses of memory used by the Oclgrind simulator back-end of AIWC.
+	In the original AIWC tool \cite{beau_johnston_2017_1134175}, global and local memory address entropy (MAE) were calculated using physical addresses of memory used by the Oclgrind simulator~\cite{price:15} back-end of AIWC.
 This caused irregularities in entropy calculations across multiple runs of the same simulation.
@@
 
-	The Extended OpenDwarfs benchmark suite is a set of OpenCL workload classified according to the computation patterns or Dwarfs. 
-	First presented in 2006, the 13 Berkeley Dwarfs are common computational and communication patterns that aim to capture the landscape of parallel computing workloads. 
+	The Extended OpenDwarfs benchmark suite~\cite{johnston18opendwarfs, krommydas2016opendwarfs} is a set of OpenCL workload classified according to the computation patterns or Dwarfs. 
+	First presented in 2006, the 13 Berkeley Dwarfs are common computational and communication patterns that aim to capture the landscape of parallel computing workloads \cite{asanovic2006landscape}. 
 	The Extended OpenDwarfs benchmark suite presents a set of highly diverse workloads that each fit the programming paradigm of one of the 13 Berkeley Dwarfs.

### 2020-01-19 paper/paper.tex
name the PSL metric earlier in relevant section

--- 2020-01-19 paper/paper.tex
@@
 Further, we find that all current AIWC metrics do not measure any sizeable difference between \texttt{coalescedAB}, \texttt{coalescedABT} and \texttt{alignedABT} codes.
-We address the root cause of this -- a lack of temporal information about memory accesses, in the next section.%TODO READ THIS SECTION
+	We address this by proposing another new metric for locality of memory accesses in the following section.
 
@@
 
-	We propose a new architecture-independent metric to measure memory access patterns of parallel programs.
+	We propose a new architecture-independent metric, \emph{parallel spatial locality}, to measure memory access patterns of parallel programs.
 	The proposed metric is inspired by CuMAPz' direct approaches to measuring optimization specific characteristics of CUDA codes.
@@
 	\textit{Calculate}: for each timestamp $t = t_0 ..
-t_{last}$, calculate the n-bits dropped entropy of memory addresses accessed by all threads in a work group within the timestamp $t$.
+t_{last}$, calculate the $n$-bits-dropped entropy of memory addresses accessed by all threads in a work group within the timestamp $t$.
 Here $n$ can range between 0 and 10 as was the case for $LMAE$.
@@
 	\textit{Summarize}: average the collected entropy values across all the timestamps to calculate the parallel spatial locality metrics for one thread group.
-We then average the n-bits dropped entropy summaries across thread groups to obtain the \textit{n-bits dropped parallel spatial locality metrics} for the kernel's execution.
+We then average the $n$-bits-dropped entropy summaries across thread groups to obtain the $n$-bits-dropped parallel spatial locality metrics for the kernel's execution.
 
@@
 
-%	With $E_{n,t}$ the n-bits dropped entropy at timestamp $t$.
+%	With $E_{n,t}$ the n-bits-dropped entropy at timestamp $t$.
 
@@
 	In parallel programs, to accurately measure spatial locality of accesses, we must consider memory accesses performed by multiple threads in a close temporal scope. 
-	The proposed metric calculates the locality of accesses in each time step of the program's execution and steeper reductions in $n$-bits dropped parallel spatial locality scores will be observed in programs that often access nearby memory addresses within the same timestamp.
+	The proposed metric calculates the locality of accesses in each time step of the program's execution and steeper reductions in $n$-bits-dropped parallel spatial locality scores will be observed in programs that often access nearby memory addresses within the same timestamp.
 	Such programs will perform better on GPUs, as they will make better use of both global memory access coalescing and shared memory bank structures.

### 2020-01-20 paper/paper.tex
wording of abstract; start of Section 6

@@
 	%This paper addresses the challenge of profiling a code base to help programmers effectively optimize their codes for high performance.
-	HPC developers are faced with the challenge of optimizing OpenCL workloads for high performance on novel architectures.
-	The Architecture Independent Workload Characterisation (AIWC) tool is a plugin for the Oclgrind OpenCL simulator that gathers metrics of OpenCL programs that can be used to understand and predict program performance on an arbitrary given hardware architecture.
+	High-performance computing developers are faced with the challenge of optimizing the performance of OpenCL workloads on diverse architectures.
+	The Architecture-Independent Workload Characterization (AIWC) tool is a plugin for the Oclgrind OpenCL simulator that gathers metrics of OpenCL programs that can be used to understand and predict program performance on an arbitrary given hardware architecture.
 	However, AIWC metrics are not always easily interpreted and do not reflect some important memory access patterns affecting efficiency across architectures.
-	We propose the concept of parallel spatial locality -- the closeness of memory accesses simultaneously issued by OpenCL work-items (threads).
+	We propose a new metric of parallel spatial locality -- the closeness of memory accesses simultaneously issued by OpenCL work-items (threads).
 	We implement the parallel spatial locality metric in the AIWC framework, and analyse gathered results on matrix multiply and the Extended OpenDwarfs OpenCL benchmarks.
-	The differences in the obtained parallel spatial locality metric across implementations of matrix multiply reflect the optimizations performed.
+	The differences in the observed parallel spatial locality metric across implementations of matrix multiply reflect the optimizations performed.
 	The new metric can be used to distinguish between the OpenDwarfs benchmarks based on the memory access patterns affecting their performance on various architectures.
@@
 	%To this end we propose, implement and analyse the addition of two new metrics to AIWC and present a new approach to profile spatial locality intrinsic to parallel codes.
-	The improvements suggested to AIWC will help HPC developers better understand memory access patterns of complex codes, a crucial step in guiding optimization of codes for arbitrary hardware targets.
+	The improvements suggested to AIWC will help HPC developers better understand memory access patterns of complex codes and guide optimization of codes for arbitrary hardware targets.
 	\end{abstract}
@@
 	In section~\ref{motivating example} we use the example of matrix multiplication on a GPU to showcase the specific vendor-recommended optimizations our work aims to measure.
-	In section~\ref{related work} we discuss how AIWC and its precursors profiled memory access patterns in an architecture-independent fashion, and relevant architecture dependent approaches to memory access profiling.
+	In section~\ref{related work} we discuss how AIWC and its precursors profiled memory access patterns in an architecture-independent fashion and also consider relevant architecture-dependent approaches to memory access profiling.
 	In section \ref{method} we present a methodology to critically examine the architecture-independent workload characterization tool, which we use in section~\ref{preliminary results} to measure the accuracy of AIWC in profiling the impact of performance optimizations to various OpenCL codes.
@@
 
-	AIWC metrics are inherently lossy.
-Given the set of metrics calculated for an unknown kernel, we can not reconstruct the kernel based on these metrics.
-Specifically to memory access pattern analysis, the calculation of memory address entropy of a kernel relies only on the frequency distribution of memory accesses to all addresses accessed by the kernel.
-We observe that calculation of memory address entropy does not make use of temporal information.
-The order of sequential memory accesses performed by each thread, as well as addresses requested in parallel by multiple threads in an OpenCL thread group, are both vital in accurately characterizing parallel codes.
+	Aggregate metrics of the kind presented by AIWC necessarily present a simplified view of program behaviour, omitting many details.
+	Different ways of aggregating program measurements lead to different features of program execution being emphasised in the final metrics.
+	For example, the calculation of memory address entropy described in section~\ref{related work} relies only on the frequency distribution of memory accesses to all addresses accessed by the kernel, and discards temporal information.
+	The order of sequential memory accesses performed by each thread, as well the relationship between work items in an OpenCL work group, are both vital in accurately characterizing parallel codes.
 
-	We propose a new architecture-independent metric, \emph{parallel spatial locality}, to measure memory access patterns of parallel programs.
+	We propose a new architecture-independent metric, \emph{parallel spatial locality}, to measure memory access patterns in parallel programs.
 	The proposed metric is inspired by CuMAPz' direct approaches to measuring optimization specific characteristics of CUDA codes.

### 2020-01-20 paper/paper.tex
captions for Figure 3; describe how we collected these numbers

@@
 	On most NVIDIA GPUs, data accesses are coalesced when multiple requests are made for memory locations from the same cache line in global memory~\cite{cudamanual}.
-	Appendix~\ref{coalescedAMultiply} contains the \texttt{coalescedAMultiply} kernel, which coalesces accesses to matrix $A$ by storing \emph{tiles}, or square blocks, of $A$'s values into shared memory (OpenCL \texttt{\_\_local} memory).
+	Appendix~\ref{coalescedAMultiply} contains the \texttt{coalescedA} kernel, which coalesces accesses to matrix $A$ by storing \emph{tiles}, or square blocks, of $A$'s values into shared memory (OpenCL \texttt{\_\_local} memory).
 
@@
 			ymin=0,
-			ylabel={Execution time ($\mu s$)}, xlabel={Kernel},
+			ylabel={Execution time (ms)}, xlabel={Kernel},
 			ylabel style={at={(axis description cs:0.1,.5)},anchor=south},
@@
 			\addplot coordinates {%[mark=*] coordinates {
-				(simple,31)
-				(coalA,19)
-				(coalAB,20)
-				(coalABT,18)
-				(alignABT,18)
+				(simple,.031)
+				(coalA,.019)
+				(coalAB,.020)
+				(coalABT,.018)
+				(alignABT,.018)
 			};
 			\addplot coordinates {%[mark=star] coordinates {
-				(simple,325)
-				(coalA,75)
-				(coalAB,81)
-				(coalABT,55)
-				(alignABT,55)
+				(simple,.325)
+				(coalA,.075)
+				(coalAB,.081)
+				(coalABT,.055)
+				(alignABT,.055)
 			};	
@@
 			ymin=0,
-			ylabel={Execution time ($\mu s$)}, xlabel={Kernel},
+			ylabel={Execution time (ms)}, xlabel={Kernel},
 			ylabel style={at={(axis description cs:0.1,.5)},anchor=south},
 			legend entries={$N=1408$,$N=2048$},
-			legend style={at={(0.6,0.45)},anchor=west}
+			legend style={at={(0.6,0.45)},anchor=west},
+			scaled ticks=false
 			]
@@
 			\addplot [mark=*, color=brown] coordinates {%[mark=triangle*] coordinates {
-				(simple,46920)
-				(coalA,8783)
-				(coalAB,9237.5)
-				(coalABT,5307)
-				(alignABT,5272)
+				(simple,46.920)
+				(coalA,8.783)
+				(coalAB,9.238)
+				(coalABT,5.307)
+				(alignABT,5.272)
 			};
 			\addplot [mark=star] coordinates {%[mark=square] coordinates {
-				(simple,130682.0)
-				(coalA,28602)
-				(coalAB,27939)
-				(coalABT,16179)
-				(alignABT,16071)
+				(simple,130.682)
+				(coalA,28.602)
+				(coalAB,27.939)
+				(coalABT,16.179)
+				(alignABT,16.071)
 			};	
@@
 		}
-		\caption{NVIDIA Tesla P100 Runtime performance of section \ref{motivating example} codes averaged over 100 runs}
+		\caption{Execution time of matrix multiplication kernels (NVIDIA Tesla P100)}
 		\label{fig: matmul runtime}
@@
 
-	Figure \ref{fig: matmul runtime} shows that across varying problem sizes, the vendor recommended optimizations to the matrix multiplication code lead to increased performance. 
+	Figure \ref{fig: matmul runtime} shows execution times for the matrix multiplication kernels presented in section~\ref{motivating example} for $N \times N$ matrices of different sizes.
+	We recorded kernel execution time exclusive of data transfer on an NVIDIA Tesla P100 using NVIDIA OpenCL 1.2 CUDA 10.1.236 (driver 418.87), and took the average of 100 runs.
+	Across varying problem sizes, the vendor recommended optimizations to the matrix multiplication code lead to increased performance. 
 	%Note that a matrix size of 80 is small -- overheads attached to optimization techniques may be causing the observed decrease in performance as some optimizations are performed.
@@
 
-	\textit{Relative local memory usage}: As reliance on NVIDIA GPUs' shared memory increases in each kernel from \texttt{simpleMultiply} to \texttt{coalescedAB}, we find that the proportion of OpenCL local memory usage increases as expected.
+	\textit{Relative local memory usage}: As reliance on NVIDIA GPUs' shared memory increases in each kernel from \texttt{simple} to \texttt{coalescedAB}, we find that the proportion of OpenCL local memory usage increases as expected.
 
-	\textit{Global and local MAE}: Entropy measurements decrease from \texttt{simpleMultiply} to \texttt{coalescedAB}, as the optimizations reduce the number of reads from matrices $A$ and $B$ in global memory, replacing these with reads from smaller tiles in local memory.
-	We observe an almost completely uniform distribution of memory accesses in \texttt{simpleMultiply}, where the program makes $N$ loads to each element of $A$ and $B$ with $N$ the dimension of the matrices.
-	The distribution of memory accesses becomes increasingly non-uniform as we perform fewer accesses to matrices $A$ and $B$ and more accesses to smaller local memory buffers, resulting in decreases in local and global memory entropy values from \texttt{simpleMultiply} to \texttt{coalescedAB}.
+	\textit{Global and local MAE}: Entropy measurements decrease from \texttt{simple} to \texttt{coalescedAB}, as the optimizations reduce the number of reads from matrices $A$ and $B$ in global memory, replacing these with reads from smaller tiles in local memory.
+	We observe an almost completely uniform distribution of memory accesses in \texttt{simple}, where the program makes $N$ loads to each element of $A$ and $B$ with $N$ the dimension of the matrices.
+	The distribution of memory accesses becomes increasingly non-uniform as we perform fewer accesses to matrices $A$ and $B$ and more accesses to smaller local memory buffers, resulting in decreases in local and global memory entropy values from \texttt{simple} to \texttt{coalescedAB}.
 
-	\textit{Memory Footprint}: Similar to trends in global and local MAE, we find that the ratio of 90\% memory footprint to total memory footprint decreases from $60.12\%$ for \texttt{simpleMultiply} to $0.25\%$ (\texttt{coalescedAB}).
+	\textit{Memory Footprint}: Similar to trends in global and local MAE, we find that the ratio of 90\% memory footprint to total memory footprint decreases from $60.12\%$ for \texttt{simple} to $0.25\%$ (\texttt{coalescedAB}).
 Increased utilization of local memory in the optimized kernels means that the local memory buffers make up a greater proportion of total memory accesses in the program.
@@
 
-	AIWC currently collects a list of all memory accessed by each thread of execution.
-	In the OpenCL programming model, threads within a work groups execute the same code and are allowed to cooperate on \textit{local memory} addresses.
+	During simulation, AIWC collects a list of all memory accesses by each thread of execution.
+	In the OpenCL programming model, threads within a work groups execute the same code and share access to \textit{local memory}.
 	We can group together memory accesses of threads in a work group at each logical time step in the symbolic execution of the code.
-	We argue that analysing the accesses occuring at the same logical time step is an accurate, conceptually sound method of predicting the extent of memory access coalescing and bank conflicts we are likely to find upon executing the code on a GPU.
+	On a GPU, memory accesses executed at the same time by different threads in a work group are likely to interact, determining the extent of memory access coalescing and bank conflicts.
 
@@
 		ylabel style={at={(axis description cs:0.05,.5)},anchor=south},
-		legend entries={\texttt{simpleMultiply},\texttt{coalescedAMultiply},
+		legend entries={\texttt{simple},\texttt{coalescedA},
 			\texttt{coalescedAB}, \texttt{coalescedABT}, \texttt{alignedABT}},
@@
 	In figure \ref{fig: matmul parallel spatial locality} we see the proposed parallel spatial locality metric obtained using the AIWC tool for each of the matrix multiplication kernels from section \ref{motivating example}.
-	In particular, we observe that the \texttt{coalescedABT} and \texttt{alignedABT} kernels have the steepest reductions in entropy as the number of bits skipped is increased, which correlates to better locality of parallel memory accesses.
-	It is expected for these kernels to exhibit better parallel spatial locality than \texttt{simpleMultiply}, as \texttt{coalescedABT} and \texttt{alignedABT} make use of local memory, where accesses tend to be localized simply due to the small size of shared on-chip memory typically available on GPUs.
+	We observe that the \texttt{coalescedABT} and \texttt{alignedABT} kernels have the steepest reductions in entropy as the number of bits skipped is increased, which correlates to better locality of parallel memory accesses.
+	It is expected for these kernels to exhibit better parallel spatial locality than \texttt{simple}, as \texttt{coalescedABT} and \texttt{alignedABT} make use of local memory, where accesses tend to be localized simply due to the small size of shared on-chip memory typically available on GPUs.
 	Further we find that the proposed metric successfully distinguishes between the \texttt{coalescedAB} and \texttt{coalescedABT} kernels.
@@
 	Memory access patterns in SRAD are statically determined and relatively localized. 
-	Similar to the LUD and \texttt{simpleMultiply} kernels, both SRAD kernels observe a rapid drop-off in parallel spatial locality (figure \ref{fig: opendwarfs parallel spatial locality}) as the number of bits skipped is increased, with the metric stabilizing at approximately half its original value when 0 bits are skipped.
+	Similar to the LUD and \texttt{simple} kernels, both SRAD kernels observe a rapid drop-off in parallel spatial locality (figure \ref{fig: opendwarfs parallel spatial locality}) as the number of bits skipped is increased, with the metric stabilizing at approximately half its original value when 0 bits are skipped.
 
@@
 
-		\section{\texttt{simpleMultiply} kernel} \label{simpleMultiply}
+		\section{\texttt{simple} kernel} \label{simpleMultiply}
 
@@
 
-		\section{\texttt{coalescedAMultiply} kernel} \label{coalescedAMultiply}
+		\section{\texttt{coalescedA} kernel} \label{coalescedAMultiply}
 
@@
 		\begin{lstlisting}
-		__kernel void coalescedAB(
+		__kernel void coalescedABMultiply(
 				const __global float* A,

### 2020-01-20 paper/paper.tex
minor wording changes; ensure Appendix is own own page

@@
 
-	The Extended OpenDwarfs benchmark suite~\cite{johnston18opendwarfs, krommydas2016opendwarfs} is a set of OpenCL workload classified according to the computation patterns or Dwarfs. 
-	First presented in 2006, the 13 Berkeley Dwarfs are common computational and communication patterns that aim to capture the landscape of parallel computing workloads \cite{asanovic2006landscape}. 
-	The Extended OpenDwarfs benchmark suite presents a set of highly diverse workloads that each fit the programming paradigm of one of the 13 Berkeley Dwarfs.
+	The Extended OpenDwarfs benchmark suite~\cite{johnston18opendwarfs, krommydas2016opendwarfs} is a set of diverse OpenCL workloads.
+	Each benchmark is assigned to one of the 13 Berkeley Dwarfs, common computational and communication patterns which aim to capture the landscape of parallel computing workloads \cite{asanovic2006landscape}.
 	The workloads presented are not optimized for any specific architecture -- hence optimizations using OpenCL local memory (which translates to CUDA shared memory) are not performed. 
@@
 	We present the results of running the AIWC tool on selected benchmarks from the Extended OpenDwarfs benchmark suite \cite{opendwarfs2020head,johnston18opendwarfs} with a problem size setting of \emph{medium}, except for the GEM benchmark, which is run at a setting of \emph{tiny}.
-	For benchmarks such as BFS with multiple kernel invocations per run, the AIWC parallel spatial locality metrics of the invocation with the highest executed LLVM-SPIR instructions are presented for analysis.
-Note that the presented benchmarks use 32 bit numbers (OpenCL \texttt{int} and \texttt{float} types), and so dropping up to 2 bits of the memory addresses accessed will not change the parallel spatial locality, since any addresses accessed are at least 4 bytes apart.
+	For benchmarks such as BFS that have multiple kernel invocations per run, we present the AIWC parallel spatial locality metrics for the invocation with the highest number of LLVM-SPIR instructions executed.
+	Note that the presented benchmarks use 32-bit numeric types (OpenCL \texttt{int} and \texttt{float}), so dropping up to 2 bits of the memory addresses accessed will not change the parallel spatial locality, since any addresses accessed are at least 4 bytes apart.
 
@@
 	% TODO
-	The OpenDwarfs benchmark GEM computes the electrostatic potential of a biomolecule by calculating the sum of charges contributed by all atoms in the biomolecule at each specific surface vertex.
+	The GEM benchmark computes the electrostatic potential of a biomolecule by calculating the sum of charges contributed by all atoms in the biomolecule at each specific surface vertex.
 	This is an embarrassingly parallel problem.
-Each OpenCL work-item operates on a single surface vertex, finding the electrostatic potential generated by looping through every atom in the biomolecule in a global order.
+Each OpenCL work-item operates on a single surface vertex, finding the electrostatic potential generated by looping through every atom in the biomolecule in global order.
 	The computation pattern is highly regular and memory accesses are perfectly synchronized. Atom data is accessed consecutively, with all work-items simultaneously accessing each atom's data.
@@
 The recorded parallel spatial locality approaches the theoretical limit of 0 (0.0124 at 10 bits dropped).
-This indicates that almost all memory accesses made by the OpenCL kernel are perfectly synchronized between OpenCL threads.
-
-	Performance results \cite{johnston18opendwarfs, krommydas2016opendwarfs} show that GPUs perform significantly better than CPUs for GEM, as memory unit stalls are at low levels for both CPUs and GPUs due to the highly efficient memory utilization of this benchmark.
-	As memory operations do not present a major slowdown, superior FPU compute capability of GPUs leads to far better performance on GPUs than CPUs \cite{krommydas2016opendwarfs}.
+This indicates that almost all memory accesses made by the kernel are perfectly synchronized between OpenCL threads.
+	Performance results \cite{johnston18opendwarfs, krommydas2016opendwarfs} show that GEM performs significantly better on GPUs than on CPUs, as memory unit stalls are at low levels for both CPUs and GPUs due to the highly efficient memory utilization of this benchmark.
+	As memory operations do not present a bottleneck, this benchmark is able to take advantage of the superior floating-point compute capability of GPUs~\cite{krommydas2016opendwarfs}.
 
@@
 	Memory access patterns in a dense linear algebra workload such as LUD are typically highly regular and deterministic for each OpenCL work item, based on the matrix dimension and offset parameters.
-The OpenDwarfs implementation of LUD \cite{opendwarfs2020head} splits the LUD computation into three kernels: \texttt{LUD perimeter} and \texttt{LUD diagonal} kernels spawn work-items in a single work dimension; while \texttt{LUD internal} performs work in both dimensions of the input matrix.
+The OpenDwarfs implementation of LUD \cite{opendwarfs2020head} splits the LUD computation into three kernels: \texttt{LUD perimeter} and \texttt{LUD diagonal} kernels spawn work-items in a single work dimension; while \texttt{LUD internal} is decomposed in both dimensions of the input.
 
@@
 
-	Needleman-Wunsch is a dynamic programming algorithm used to perform protein sequence alignment by identifying the similarity level between two strings of Amino acids.
-Elements in a matrix are computed, with each element dependent on its west, north and north-west neighbours.
+	Needleman-Wunsch is a dynamic programming algorithm used to perform protein sequence alignment by identifying the similarity between two strings of amino acids.
+	The computation of each element in the similarity matrix depends on its west, north and north-west neighbours.
 	This dependency enforces a wavefront computation pattern, travelling along the main diagonal of the matrix.
@@
 	Graph traversal algorithms require pointer chasing operations to traverse nodes of a graph and perform calculations. 
-	The Breadth First Search implemented in OpenDwarfs calls two OpenCL kernels to traverse nodes immediately connected to the list of nodes at each depth starting from the root node. 
+	The Breadth-First Search implemented in OpenDwarfs calls two OpenCL kernels to traverse nodes immediately connected to the list of nodes at each depth starting from the root node. 
 	BFS is characterized by an imbalanced workload per kernel launch depending on the degree of the nodes being operated on, with only a proportion of launched nodes performing meaningful work. 
@@
 	Work-items fetch discontinuous memory locations attributed to any particular node in the graph, depending on the connectivity of the node they operate on. 
-	The program structure involves multiple levels of pointer-chasing and thus the precise memory address accessed are runtime dependent. 
+	The program structure involves multiple levels of pointer-chasing and thus the precise memory addresses accessed are data-dependent. 
 	This leads to poor parallel spatial locality. Thus, we see a slow dropoff in parallel spatial locality for \texttt{BFS kernel 1}.
@@
 
-	Compressed Sparse Row Matrix-Vector Multiplication (CSR) computes the sum of each of a matrix's row's elements after it is multiplied by a given vector.
+	Compressed Sparse Row Matrix-Vector Multiplication (CSR) computes the product of a sparse matrix and a dense vector.
 	The matrix is stored in a compressed row storage sparse matrix format, which is very efficient for storage when the number of zero elements is far greater than the number of non-zero elements.
@@
 
-	In this particular implementation, a single OpenCL work-item is assigned to compute over each row of the input sparse matrix. 
+	In the OpenDwarfs implementation, each row of the input sparse matrix is assigned to a separate OpenCL work item. 
 	The locations of matrix data read by each work-item are dependent on the number and position of non-zero values in the sparse-matrix, which are decided by the values in $Aj$ and $Ap$. 
@@
 	\bibliography{bib/bibliography.bib}
+
+	\pagebreak
 

### 2020-01-20 paper/paper.tex
reorganize related work and combine non-parallel methodology and results

@@
 	In section~\ref{related work} we discuss how AIWC and its precursors profiled memory access patterns in an architecture-independent fashion and also consider relevant architecture-dependent approaches to memory access profiling.
-	In section \ref{method} we present a methodology to critically examine the architecture-independent workload characterization tool, which we use in section~\ref{preliminary results} to measure the accuracy of AIWC in profiling the impact of performance optimizations to various OpenCL codes.
+	In section \ref{method} we consider non-parallel memory metrics collected by the architecture-independent workload characterization tool and evaluate the suitability of these metrics for capturing the impact of performance optimizations to various OpenCL codes.
 	In section~\ref{discussion} we discuss these preliminary results, and propose and implement our parallel spatial locality metric into the AIWC framework.
@@
 	All the examples above are an incomplete listing of optimization strategies developers can apply to code targeted at NVIDIA GPUs.
-	In section~\ref{preliminary results}, we will compare the collected AIWC metrics with the expected effect of each optimization to examine how effectively our methodology uncovers underlying bottlenecks and guides optimization efforts.
+	In section~\ref{method}, we will compare the collected AIWC metrics with the expected effect of each optimization to examine how effectively our methodology uncovers underlying bottlenecks and guides optimization efforts.
 
@@
 	For example, the causes of high cache miss rates in the execution of a program are complex and depend on microarchitecture specific features such as cache size, prefetch behaviour and cache placement policies.
-	We argue that a HPC developer tasked with optimizing code for a given hardware target would benefit from architecture-independent metrics of the code that can be conceptually correlated to hardware specific optimization patterns.
-Such architecture-independent metrics will effectively guide the HPC developer's workflow in both finding and fixing bottlenecks in code performance.
-	% TODO this argument isn't very clear
+	A HPC developer tasked with optimizing code for a given hardware target would benefit from architecture-independent metrics of the code that can be used to measure the effect of code modifications.
+    These metrics would help to guide the developer in both finding and fixing performance bottlenecks.
 
-	The Architecture-Independent Workload Characterization (AIWC) tool \cite{beauaiwc} relies on a set of instruction set architecture (ISA)-independent features determined by Shao and Brooks \cite{shao2013isa}.
-AIWC measures memory based metrics to characterize the target code's memory behaviours.
-AIWC first collects aggregate simple counts such as the (i) total memory footprint, the total number of unique addresses accessed; and (ii) 90\% memory footprint, the number of unique addresses covering 90\% of memory accesses.
-A small ratio of 90\% memory footprint to total memory footprint correlates to programs accessing the same small subset of memory addresses repeatedly -- which can be linked to increased performance in a cached memory hierarchy.
+	To address limitations of conventional microarchitecture-dependent characteristics, Hoste and Eeckout~\cite{hoste2007microarchitecture} developed the Microarchitecture-Independent Workload Characterization tool (MICA).
+	They observed that performance counter based approaches to profiling codes often failed to find underlying program features that map to improved or worsened usage of performance critical hardware features of the target architecture.
+	The MICA framework is a holistic characterization tool, and thus collects features including instruction mix, instruction-level parallelism, register traffic, data stream strides and branch predictability.
+
+	Of these metrics, data stream strides are of particular interest in memory access pattern profiling.
+MICA's stride length metric measures the distance between consecutive memory accesses in a single threaded application.
+For CPU architectures running single threaded applications, this metric correlates to the spatial locality of memory accesses -- a measure of how closely bunched are memory access in nearby times.
+This is directly correlated to cache reuse rates, critical to code performance on CPUs \cite{inteloptimisation}.
+%Intel optimization manual.
+
+	The MICA approach was tailored for single-threaded applications as the metrics collected rely heavily on Pin instrumentation \cite{luk2005pin}.
+	As such, MICA is unsuited to analysing HPC workloads with heavy use of parallelism.
+	The Workload ISA-Independent Characterization for Applications (WIICA)~\cite{shao2013isa} extends MICA to present a framework to analyse single-threaded programs independent of the instruction set architecture (ISA).
+
+	Kim and Shrivastava \cite{kim2011cumapz} present CuMAPz, a CUDA memory access profiling tool to guide NVIDIA GPU optimizations.
+	CuMAPz focuses on the problem of improving CUDA application performance using NVIDIA memory-based optimizations.
+	CuMAPz analyses CUDA codes structurally and simulates code execution on the memory hierarchy of specific NVIDIA GPU models.
+
+	During simulation of the target code, CuMAPz records all memory accesses in various buffers to simulate the global and shared memories on NVIDIA GPUs.
+	Using this detailed simulation data, CuMAPz can estimate the performance critical (i) shared memory data reuse profit, (ii) profit from coalesced access (iii) memory channel skew cost and (iv) bank conflict cost characteristics of the target code.
+
+	The simulation environment used by CuMAPz and the attached analysis framework is highly specific to CUDA enabled GPUs.
+	Replicating the CuMAPz framework for all target architectures is challenging.
+	However, CuMAPz is an interesting simulator from the standpoint of this paper's work since it adheres to the core parts of NVIDIA GPUs' memory models in its analysis, while allowing the user to specify their GPU model specific hardware information.
+
+	Our approach is an architecture-independent analysis of memory access patterns to provide metrics correlating to similar performance critical memory access optimizations as CuMAPz.
+	We aim to further the state of the art by providing a framework to guide developers in optimizing OpenCL codes for any given target architecture.
+	To the best our knowledge, none of the previous works present a set of performance metrics that accurately characterize memory access patterns of parallel applications independent of the target architecture.
+
+	The Architecture-Independent Workload Characterization (AIWC) tool \cite{beauaiwc} collects a set of instruction set architecture (ISA)-independent features based on those identified by Shao and Brooks \cite{shao2013isa}.
+	AIWC runs as a plugin to the Oclgrind~\cite{price:15} framework, which simulates OpenCL kernel execution on an ideal device according to the OpenCL execution and memory models.
+	AIWC collects metrics of kernel memory access, including simple counts such as the (i) total memory footprint, the total number of unique addresses accessed; and (ii) 90\% memory footprint, the number of unique addresses covering 90\% of memory accesses.
+	While these metrics are architecture-independent, they are correlated to program performance on typical architectures.
+	For example, a small ratio of 90\% memory footprint to total memory footprint indicates that a program accesses a small subset of memory addresses frequently, which is highly beneficial for performance in a cached memory hierarchy.
 
@@
 A steeper drop in entropy with increasing number of bits may correlate to more localized memory accesses over the program's execution.
-
-	Limitations of the memory footprint and local memory address entropy values as memory access pattern metrics are further discussed by this work. %TODO: REMOVE?
-
-	\subsection{Advancements in Memory Access Pattern Characterization}
-
-	We provide a brief summary of past work in architecture independent program characterization and Memory Access Pattern analysis:
-
-	\begin{itemize}
-		\item The Microarchitecture Independent Workload Characterization Tool (MICA) \cite{hoste2007microarchitecture} was the first microarchitecture independent workload characterization tool for single threaded programs.
-		\item The Workload ISA-Independent Characterization for Applications (WIICA) \cite{shao2013isa} extends MICA to present a framework to analyse single threaded programs independent of the ISA.
-		\item CuMAPz: CuMAPz \cite{kim2011cumapz} is a CUDA memory access profiling tool and the first to present metrics for CUDA codes which can be generalized to any given NVIDIA GPU.
-	\end{itemize}
-
-	\subsection{MICA}
-
-	The MICA framework was developed by Hoste and Eeckout \cite{hoste2007microarchitecture} in response to their findings on the limitations of conventional microarchitecture dependent characteristics.
-Hoste and Eeckout discovered that performance counter based approaches to profiling codes often struggled in finding underlying program features that map to improved or worsened usage of performance critical hardware features of the target architecture.
-The MICA framework is a holistic characterization tool, and thus collects features including instruction mix, instruction-level parallelism, register traffic, data stream strides and branch predictability.
-
-	Of these metrics, data stream strides are of particular interest in memory access pattern profiling.
-MICA's stride length metric measures the distance between consecutive memory accesses in a single threaded application.
-For CPU architectures running single threaded applications, this metric correlates to the spatial locality of memory accesses -- a measure of how closely bunched are memory access in nearby times.
-This is directly correlated to cache reuse rates, critical to code performance on CPUs \cite{inteloptimisation}.
-%Intel optimization manual.
-
-	The MICA approach was tailored for single-threaded applications as the metrics collected rely heavily on Pin instrumentation \cite{luk2005pin}.
-As such, MICA is unsuited to analysing HPC workloads with heavy use of parallelism.
-
-	\subsection{CuMAPz}
-
-	Kim and Shrivastava \cite{kim2011cumapz} present CuMAPz, a CUDA memory access pattern analysis tool to guide NVIDIA GPU optimizations.
-CuMAPz focuses on the problem of improving CUDA application performance using NVIDIA memory-based optimizations.
-CuMAPz analyses CUDA codes structurally and simulates code execution on the memory hierarchy of specific NVIDIA GPU models.
-
-	During the simulated run of the target code, CuMAPz records all memory accesses in various buffers to simulate the global and shared memories on NVIDIA GPUs.
-Using this detailed simulation data, CuMAPz can estimate the performance critical (i) shared memory data reuse profit, (ii) profit from coalesced access (iii) memory channel skew cost and (iv) bank conflict cost characteristics of the target code.
-
-	The simulation environment described in CuMAPz and the attached analysis framework is highly specific to CUDA enabled GPUs.
-Replicating the CuMAPz framework for all target architectures is challenging.
-However, CuMAPz is an interesting simulator from the standpoint of this paper's work since it adheres to the core parts of NVIDIA GPUs' memory models in its analysis, while allowing the user to specify their GPU model specific hardware information.
-
-	Our approach is an architecture independent analysis of memory access patterns to provide metrics correlating to similar performance critical memory access optimizations as CuMAPz.
-We aim to further the state of the art by providing a framework to guide developers in optimizing OpenCL codes for any given target architecture.
-To the best our knowledge, none of the previous works present a set of performance metrics that accurately characterize memory access patterns of parallel applications independent of the target architecture.
 
@@
 
-	\section{Methodology} \label{method}
+	\section{Non-Parallel Memory Metrics} \label{method}
 
-	To provide memory access pattern metrics that represent vendor recommended optimizations on GPUs, we improved AIWC by adding our first new metric: relative local memory usage.
+	We first consider AIWC memory metrics that do not take into account the interaction between work items executing in parallel.
+	We compare these metrics for the incrementally optimized versions of the matrix multiplication example presented in section~\ref{motivating example}.
+
+	In addition to the memory metrics reported in \cite{beauaiwc}, we added a new metric of relative local memory usage.
 	This measures the proportion of all memory accesses from the symbolic execution of the kernel that occurred to memory allocated as \texttt{\_\_local}.
 	On NVIDIA GPUs, this memory address space is mapped to fast on-chip shared memory.
-
 	Relative local memory usage is an example of a metric that is useful to measure performance critical access patterns on some architectures such as GPUs, and not others, such as CPUs.
@@
 This allows AIWC to accurately abstract over the hardware and ISA-specific differences in memory layouts across the diverse hardware targets.
-
-	\section{Preliminary Results} \label{preliminary results}
 

### 2020-01-20 paper/paper.tex
results -> Evaluation: EOD

@@
 	Our work highlights the benefits and challenges arising from an architecture-independent analysis of memory-based program characteristics.
-	We propose two new metrics to AIWC to better characterize these memory-based optimizations.
+	We propose two new metrics for AIWC to better characterize these memory-based optimizations.
 	We measure the presented codes using the new metrics and demonstrate the metrics' effectiveness in capturing the essence of the optimizations performed.
@@
 	In section~\ref{related work} we discuss how AIWC and its precursors profiled memory access patterns in an architecture-independent fashion and also consider relevant architecture-dependent approaches to memory access profiling.
-	In section \ref{method} we consider non-parallel memory metrics collected by the architecture-independent workload characterization tool and evaluate the suitability of these metrics for capturing the impact of performance optimizations to various OpenCL codes.
-	In section~\ref{discussion} we discuss these preliminary results, and propose and implement our parallel spatial locality metric into the AIWC framework.
-	In section~\ref{results} we present the collected metric for the matrix multiplication kernels and selected OpenDwarfs benchmarks to validate our methodology in this paper.
+	In section \ref{method} we consider non-parallel memory metrics collected by AIWC and evaluate the suitability of these metrics for capturing the impact of performance optimizations to various OpenCL codes.
+	In section~\ref{discussion} we propose a new parallel spatial locality metric and implement it in the AIWC framework.
+	In section~\ref{results} we present the collected metric for selected benchmarks from the Extended OpenDwarfs benchmark suite to validate our methodology in this paper.
 	Finally, in section~\ref{future work} we discuss further avenues to extend our work and conclude.
@@
 
-	\section{Results} \label{results}
-
-	In figure \ref{fig: matmul parallel spatial locality} we see the proposed parallel spatial locality metric obtained using the AIWC tool for each of the matrix multiplication kernels from section \ref{motivating example}.
+	Figure~\ref{fig: matmul parallel spatial locality} shows the proposed parallel spatial locality metric as measured by AIWC for each of the matrix multiplication kernels from section~\ref{motivating example}.
 	We observe that the \texttt{coalescedABT} and \texttt{alignedABT} kernels have the steepest reductions in entropy as the number of bits skipped is increased, which correlates to better locality of parallel memory accesses.
@@
 	This is a significant improvement over the state-of-the-art AIWC metrics in characterizing how codes localize simultaneous memory accesses to better use the hardware provided.
+
+	\section{Evaluation: Extended OpenDwarfs Benchmark Suite} \label{results}
 

### 2020-01-20 paper/paper.tex
minor grammar fixes

@@
 
-	Hoste and Eeckout \cite{hoste2007microarchitecture} show that while conventional architecture dependent characteristics are useful in locating performance bottlenecks, they can hide the underlying, inherent program behaviour causing performance bottlenecks.
+	Hoste and Eeckout \cite{hoste2007microarchitecture} show that while conventional architecture-dependent characteristics are useful in locating performance bottlenecks, they can hide the underlying, inherent program behaviour causing performance bottlenecks.
 	A conceptual understanding of the performance characteristics of complex codes is necessary for the programmer to effectively optimize these codes.
-Architecture dependent characteristics typically include instructions per cycle (IPC), cache and branch prediction miss rates, page faults and DRAM bus data transfer rates.
+	Architecture-dependent characteristics typically include instructions per cycle (IPC), cache and branch prediction miss rates, page faults and DRAM bus data transfer rates.
 These are typically collected using hardware performance counters available on most target architectures.
 These performance counters do not serve to guide optimization beyond highlighting potential bottlenecks \cite{hoste2007microarchitecture, ganesan2008performance}.
-Further, in many cases, architecture dependent characteristics usually cannot be directly correlated to specific code patterns.
+	Further, in many cases, architecture-dependent characteristics cannot be directly correlated to specific code patterns.
 	For example, the causes of high cache miss rates in the execution of a program are complex and depend on microarchitecture specific features such as cache size, prefetch behaviour and cache placement policies.
@@
 
-	To address limitations of conventional microarchitecture-dependent characteristics, Hoste and Eeckout~\cite{hoste2007microarchitecture} developed the Microarchitecture-Independent Workload Characterization tool (MICA).
-	They observed that performance counter based approaches to profiling codes often failed to find underlying program features that map to improved or worsened usage of performance critical hardware features of the target architecture.
+	To address the limitations of conventional microarchitecture-dependent characteristics, Hoste and Eeckout~\cite{hoste2007microarchitecture} developed the Microarchitecture-Independent Workload Characterization tool (MICA).
+	They observed that performance counter-based approaches to profiling codes often failed to find underlying program features that map to improved or worsened usage of performance-critical hardware features of the target architecture.
 	The MICA framework is a holistic characterization tool, and thus collects features including instruction mix, instruction-level parallelism, register traffic, data stream strides and branch predictability.
@@
 	Of these metrics, data stream strides are of particular interest in memory access pattern profiling.
-MICA's stride length metric measures the distance between consecutive memory accesses in a single threaded application.
-For CPU architectures running single threaded applications, this metric correlates to the spatial locality of memory accesses -- a measure of how closely bunched are memory access in nearby times.
-This is directly correlated to cache reuse rates, critical to code performance on CPUs \cite{inteloptimisation}.
-%Intel optimization manual.
+	MICA's stride length metric measures the distance between consecutive memory accesses in a single-threaded application.
+	For CPU architectures running single-threaded applications, this metric correlates to the spatial locality of memory accesses -- a measure of how closely bunched are memory access in nearby times.
+	This is directly correlated to cache reuse rates, critical to code performance on CPUs \cite{inteloptimisation}. %Intel optimization manual.
 
@@
 	During simulation of the target code, CuMAPz records all memory accesses in various buffers to simulate the global and shared memories on NVIDIA GPUs.
-	Using this detailed simulation data, CuMAPz can estimate the performance critical (i) shared memory data reuse profit, (ii) profit from coalesced access (iii) memory channel skew cost and (iv) bank conflict cost characteristics of the target code.
+	Using this detailed simulation data, CuMAPz can estimate the performance-critical (i) shared memory data reuse profit, (ii) profit from coalesced access (iii) memory channel skew cost and (iv) bank conflict cost characteristics of the target code.
 
@@
 	We aim to further the state of the art by providing a framework to guide developers in optimizing OpenCL codes for any given target architecture.
-	To the best our knowledge, none of the previous works present a set of performance metrics that accurately characterize memory access patterns of parallel applications independent of the target architecture.
+	To the best our knowledge, none of the previous works presents a set of performance metrics that accurately characterize memory access patterns of parallel applications independent of the target architecture.
 
@@
 	On NVIDIA GPUs, this memory address space is mapped to fast on-chip shared memory.
-	Relative local memory usage is an example of a metric that is useful to measure performance critical access patterns on some architectures such as GPUs, and not others, such as CPUs.
-CPUs do not typically have a notion of user controlled on-chip memory shared between hardware threads such as NVIDIA GPUs' shared memory.
-This is a natural consequence of programming for a heterogeneous system.
-Specific code patterns may translate to performance improvements only on certain hardware.
+	Relative local memory usage is an example of a metric that is useful to measure performance-critical access patterns on some architectures such as GPUs, and not others, such as CPUs.
+	CPUs do not typically have a notion of user-controlled on-chip memory shared between hardware threads such as NVIDIA GPUs' shared memory.
+	This is a natural consequence of programming for a heterogeneous system. Specific code patterns may translate to performance improvements only on certain hardware.
 

### 2020-01-20 paper/paper.tex
why do we use EOD?

@@
 	Each benchmark is assigned to one of the 13 Berkeley Dwarfs, common computational and communication patterns which aim to capture the landscape of parallel computing workloads \cite{asanovic2006landscape}.
+	To show that the proposed parallel spatial locality metric is useful for understanding performance properties of a wide range of application codes, we present the results of running the AIWC tool on selected benchmarks from the suite~\cite{opendwarfs2020head}.
 	The workloads presented are not optimized for any specific architecture -- hence optimizations using OpenCL local memory (which translates to CUDA shared memory) are not performed.
+
 	Many of these benchmarks can be run with up to four problem sizes based on the sizes of caches found in modern CPU memory hierarchies~\cite{johnston18opendwarfs}.
-	We present the results of running the AIWC tool on selected benchmarks from the Extended OpenDwarfs benchmark suite \cite{opendwarfs2020head,johnston18opendwarfs} with a problem size setting of \emph{medium}, except for the GEM benchmark, which is run at a setting of \emph{tiny}.
+	For these results we used a problem size setting of \emph{medium}, except for the GEM benchmark, which is run at a setting of \emph{tiny}.
 	For benchmarks such as BFS that have multiple kernel invocations per run, we present the AIWC parallel spatial locality metrics for the invocation with the highest number of LLVM-SPIR instructions executed.

### 2020-01-20 paper/paper.tex
fix intro

@@
 	In section \ref{method} we consider non-parallel memory metrics collected by AIWC and evaluate the suitability of these metrics for capturing the impact of performance optimizations to various OpenCL codes.
-	In section~\ref{discussion} we propose a new parallel spatial locality metric and implement it in the AIWC framework.
+	In section~\ref{discussion} we propose a new parallel spatial locality metric, implement it in the AIWC framework, and evaluate it over the matrix multiplication example from section~\ref{motivating example}.
 	In section~\ref{results} we present the collected metric for selected benchmarks from the Extended OpenDwarfs benchmark suite to validate our methodology in this paper.

### 2020-03-13 paper/paper.tex
Working on improvements for camera ready

@@
 
+	\begin{comment}
 	\subsection{Alignment of Memory Allocation}
@@
 	\end{lstlisting}
-
+	\end{comment}
 	All the examples above are an incomplete listing of optimization strategies developers can apply to code targeted at NVIDIA GPUs.
@@
 
+	%TODO: Are we making our work sound too much like CuMAPz? This is of course not true and we should fix Reviewer 2's confusion as to what we are doing differently.
+
 	Our approach is an architecture-independent analysis of memory access patterns to provide metrics correlating to similar performance critical memory access optimizations as CuMAPz.
 	We aim to further the state of the art by providing a framework to guide developers in optimizing OpenCL codes for any given target architecture.
-	To the best our knowledge, none of the previous works presents a set of performance metrics that accurately characterize memory access patterns of parallel applications independent of the target architecture.
+	To the best of our knowledge, none of the previous works presents a set of performance metrics that accurately characterize memory access patterns of parallel applications independent of the target architecture.
 
@@
 			\begin{axis}[symbolic x coords={simple,coalA,coalAB,coalABT,alignABT},
-			xtick={simple,coalA,coalAB,coalABT,alignABT},
+			xtick={simple,coalA,coalAB,coalABT},%alignABT},
 			xticklabel style={text height=2ex},
@@
 				(coalABT,.018)
-				(alignABT,.018)
+				%(alignABT,.018)
 			};
@@
 				(coalABT,.055)
-				(alignABT,.055)
+				%(alignABT,.055)
 			};	
@@
 			\begin{tikzpicture}
-			\begin{axis}[symbolic x coords={simple,coalA,coalAB,coalABT,alignABT},
-			xtick={simple,coalA,coalAB,coalABT,alignABT},
+			\begin{axis}[symbolic x coords={simple,coalA,coalAB,coalABT},%alignABT},
+			xtick={simple,coalA,coalAB,coalABT},%alignABT},
 			xticklabel style={text height=2ex},
@@
 				(coalABT,5.307)
-				(alignABT,5.272)
+				%(alignABT,5.272)
 			};
@@
 				(coalABT,16.179)
-				(alignABT,16.071)
+				%(alignABT,16.071)
 			};	
@@
 		\toprule
-		& simple & coalescedA & coalescedAB & coalescedABT & alignedABT \\ \midrule 
-		Total memory footprint        & 196608 & 196608     & 196608      & 196608       & 196608     \\ %\hline
-		90\% Memory Footprint         & 118196 & 56176      & 489         & 489          & 489        \\ %\hline
-		Global MAE                    & 17.02  & 13.18      & 9.78        & 9.78         & 9.78       \\ %\hline
-		LMAE \#bits=3                 & 16.02  & 12.18      & 8.78        & 8.78         & 8.78       \\ %\hline
-		LMAE \#bits=10                & 9.02   & 5.18       & 1.78        & 1.78         & 1.78       \\ %\hline
-		Relative Local Memory Usage   & 0      & 0.50       & 0.94       &  0.94         & 0.94      \\ \bottomrule
+		& simple & coalescedA & coalescedAB & coalescedABT \\ \midrule %& alignedABT \\ \midrule 
+		Total memory footprint        & 196608 & 196608     & 196608      & 196608      \\ %\hline
+		90\% Memory Footprint         & 118196 & 56176      & 489         & 489         \\ %\hline
+		Global MAE                    & 17.02  & 13.18      & 9.78        & 9.78        \\ %\hline
+		LMAE \#bits=3                 & 16.02  & 12.18      & 8.78        & 8.78        \\ %\hline
+		LMAE \#bits=10                & 9.02   & 5.18       & 1.78        & 1.78        \\ %\hline
+		Relative Local Memory Usage   & 0      & 0.50       & 0.94       &  0.94        \\ \bottomrule
 	\end{tabular}
@@
 	The proposed relative local memory usage metric is the first to correspond to a recommended optimization strategy of using fast on-chip shared memory. 
-Further, we find that all current AIWC metrics do not measure any sizeable difference between \texttt{coalescedAB}, \texttt{coalescedABT} and \texttt{alignedABT} codes.
+	Further, we find that all current AIWC metrics do not measure any sizeable difference between \texttt{coalescedAB} and \texttt{coalescedABT} codes.%and \texttt{alignedABT} codes.
 	We address this by proposing another new metric for locality of memory accesses in the following section.
@@
 		legend entries={\texttt{simple},\texttt{coalescedA},
-			\texttt{coalescedAB}, \texttt{coalescedABT}, \texttt{alignedABT}},
+			\texttt{coalescedAB}, \texttt{coalescedABT}},%\texttt{alignedABT}},
 		legend style={at={(0.55,0.75)},anchor=west}
@@
 
-		\addplot [mark=*, color=red] coordinates {%[mark=square] coordinates {
+		\addplot coordinates {%[mark=square] coordinates {
 			(0,4.145)
@@
 
-		\addplot coordinates {%[mark=o] coordinates {
-			(0,4.145)
-			(1,4.145)
-			(2,4.145)
-			(3,3.2)
-			(4,2.254)
-			(5,1.308)
-			(6,0.3621)
-			(7,0.318)
-			(8,0.2738)
-			(9,0.2296)	
-			(10,0.1854)
-		};
 		\end{axis}
@@
 	Figure~\ref{fig: matmul parallel spatial locality} shows the proposed parallel spatial locality metric as measured by AIWC for each of the matrix multiplication kernels from section~\ref{motivating example}.
-	We observe that the \texttt{coalescedABT} and \texttt{alignedABT} kernels have the steepest reductions in entropy as the number of bits skipped is increased, which correlates to better locality of parallel memory accesses.
-	It is expected for these kernels to exhibit better parallel spatial locality than \texttt{simple}, as \texttt{coalescedABT} and \texttt{alignedABT} make use of local memory, where accesses tend to be localized simply due to the small size of shared on-chip memory typically available on GPUs.
+	We observe that the \texttt{coalescedABT} kernel has the steepest reductions in entropy as the number of bits skipped is increased, which correlates to better locality of parallel memory accesses.
+	It is expected for this kernel to exhibit better parallel spatial locality than \texttt{simple}, as \texttt{coalescedABT} makes use of local memory, where accesses tend to be localized simply due to the small size of shared on-chip memory typically available on GPUs.
 	Further we find that the proposed metric successfully distinguishes between the \texttt{coalescedAB} and \texttt{coalescedABT} kernels.
@@
 		\addplot coordinates {%[mark=*] coordinates {
-			(0, 0.1139)
-			(1, 0.1139)
-			(2, 0.1139)
-			(3, 0.0989)
-			(4, 0.08402)
-			(5, 0.06914)
-			(6, 0.05431)
-			(7, 0.03938)
-			(8, 0.02442)
-			(9, 0.0124)
-			(10, 0.00038)
+			(0, 0.08604)
+			(1, 0.08604)
+			(2, 0.08604)
+			(3, 0.07516)
+			(4, 0.06431)
+			(5, 0.05346)
+			(6, 0.04262)
+			(7, 0.03176)
+			(8, 0.021250)
+			(9, 0.01073)
+			(10, 0.000218)
 		};
@@
 		\addplot coordinates {%[mark=star] coordinates {
-			(0, 2.175)
-			(1, 2.175)
-			(2, 2.175)
-			(3, 2.175)
-			(4, 2.175)
-			(5, 2.175)
-			(6, 2.175)
-			(7, 2.175)
-			(8, 2.175)
-			(9, 2.175)
-			(10, 2.175)
+			(0, .5315)
+			(1, .5315)
+			(2, .5315)
+			(3, .5315)
+			(4, .5315)
+			(5, .5315)
+			(6, .5315)
+			(7, .5315)
+			(8, .5315)
+			(9, .5315)
+			(10, .5315)
+		};
+
+		\addplot coordinates {%[mark=star] coordinates {
+			(0, .8963)
+			(1, .8963)
+			(2, .8963)
+			(3, .7871)
+			(4, .6768)
+			(5, .5663)
+			(6, .4557)
+			(7, .4521)
+			(8, .4503)
+			(9, .4494)
+			(10, .4489)
 		};
@@
 		\addplot coordinates {%[mark=*] coordinates {
-			(0, 7.176)
-			(1, 7.176)
-			(2, 7.176)
-			(3, 6.301)
-			(4, 5.418)
-			(5, 4.533)
-			(6, 3.648)
-			(7, 3.619)
-			(8, 3.605)
-			(9, 3.598)
-			(10, 3.594)
-		};
-
-		\addplot coordinates {%[mark=star] coordinates {
-			(0, 7.097)
-			(1, 7.097)
-			(2, 7.097)
-			(3, 6.223)
-			(4, 5.344)
-			(5, 4.464)
-			(6, 3.583)
-			(7, 3.567)
-			(8, 3.559)
-			(9, 3.554)
-			(10, 3.553)
+			(0, .8866)
+			(1, .8866)
+			(2, .8866)
+			(3, .7773)
+			(4, .6675)
+			(5, .5576)
+			(6, .4476)
+			(7, .4456)
+			(8, .4446)
+			(9, .4441)
+			(10, .4439)
 		};
@@
 		\addplot coordinates {%[mark=triangle*] coordinates {
-			(0, 9.605)
-			(1, 9.605)
-			(2, 9.605)
-			(3, 9.478)
-			(4, 9.293)
-			(5, 9.049)
-			(6, 8.761)
-			(7, 8.447)
-			(8, 8.12)
-			(9,7.233)
-			(10,6.287)
+			(0, .9605)
+			(1, .9605)
+			(2, .9605)
+			(3, .9478)
+			(4, .9293)
+			(5, .9049)
+			(6, .8761)
+			(7, .8447)
+			(8, .812)
+			(9,.7233)
+			(10,.6287)
 		};
@@
 		\addplot coordinates { %[mark=square] coordinates {
-			(0,4.933)	
-			(1,4.933)
-			(2,4.933)
-			(3,4.89)
-			(4,4.795)
-			(5,4.652)
-			(6,4.456)
-			(7,4.215)
-			(8,3.93)
-			(9,3.608)
-			(10,3.253)
+			(0, .6163)	
+			(1, .6163)
+			(2, .6163)
+			(3, .6109)
+			(4, .599)
+			(5, .5812)
+			(6, .5567)
+			(7, .5266)
+			(8, .491)
+			(9, .4508)
+			(10, .4064)
 		};
@@
 		\addplot coordinates { %[mark=square] coordinates {
-			(0,4.691)
-			(1,4.691)
-			(2,4.691)
-			(3,4.276)
-			(4,3.760)
-			(5,3.171)
-			(6,2.552)
-			(7,1.922)
-			(8,1.286)
-			(9,0.648)
-			(10,0.00776)
+			(0,.5859)
+			(1,.5859)
+			(2,.5859)
+			(3,.5341)
+			(4,.4696)
+			(5,.3961)
+			(6,.3188)
+			(7,.2401)
+			(8,.1607)
+			(9,.08088)
+			(10,.004861)
+		};
+
+		\addplot coordinates { %[mark=star] coordinates {
+			(0,.3787)
+			(1,.3787)
+			(2,.3787)
+			(3,.3192)
+			(4,.2671)
+			(5,.2296)
+			(6,.2152)
+			(7,.2152)
+			(8,.2152)
+			(9,.2152)
+			(10,.2152)
 		};
@@
 		\addplot coordinates { %[mark=o] coordinates {
-			(0,1.548)
-			(1,1.548)
-			(2,1.548)
-			(3,1.305)
-			(4,1.092)
-			(5,0.9385)
-			(6,0.8796)
-			(7,0.8796)
-			(8,0.8796)
-			(9,0.8796)
-			(10,0.8796)
-		};
-
-	    \addplot coordinates { %[mark=star] coordinates {
-	    	(0,3.803)
-	    	(1,3.803)
-	    	(2,3.803)
-	    	(3,3.473)
-	    	(4,3.144)
-	    	(5,2.814)
-	    	(6,2.485)
-	    	(7,2.485)
-	    	(8,2.485)
-	    	(9,2.485)
-	    	(10,2.485)
+			(0,.7538)
+			(1,.7538)
+			(2,.7538)
+			(3,.6885)
+			(4,.6231)
+			(5,.5578)
+			(6,.4925)
+			(7,.4925)
+			(8,.4925)
+			(9,.4925)
+			(10,.4925)
 		};
@@
    		\addplot coordinates { %[mark=square*] coordinates {
-   			(0,4.101)
-   			(1,4.101)
-   			(2,4.101)
-   			(3,3.589)
-   			(4,3.077)
-   			(5,2.565)
-   			(6,2.053)
-   			(7,2.053)
-   			(8,2.053)
-   			(9,2.053)
-   			(10,2.053)
+   			(0,.5115)
+   			(1,.5115)
+   			(2,.5115)
+   			(3,.4477)
+   			(4,.3838)
+   			(5,.3200)
+   			(6,.2561)
+   			(7,.2561)
+   			(8,.2561)
+   			(9,.2561)
+   			(10,.2561)
    		};
@@
 	Similar to the presented matrix multiply codes, LUD can be optimized for GPUs by effective utilization of shared memory \cite{opendwarfs2020head}.
-
-	%The use of creates imbalanced workload and divergent control flow within work-items in each kernel iteration.
-While the presence of conditional statements cause some thread divergence in the LUD diagonal kernel, the use of barriers following conditional statements synchronizes the threads.
-LUD diagonal and perimeter kernels , while LUD.
-	%TODO: Divergence in the kernels. Make a diagram/code simplification for LUD.
 
@@
 
+	\section{Guidelines: Using the Parallel Spatial Locality Metric}
+
+	-- Notes
+
+	 GEM: The recorded parallel spatial locality approaches the theoretical limit of 0 (0.0124 at 10 bits dropped).
+This indicates that almost all memory accesses made by the kernel are perfectly synchronized between OpenCL threads.
+
+	 LUD: Codes with a mix of favourable and unfavourable access patterns (typically both dimensions of matrices) will have a steep dropoff to ~0.5 of the starting PSL metric, and flattens out at that point.
+
+	%TODO: Compile list of guidelines
+
+	\begin{comment}
+	\section{Predicting OpenCL Kernel Performance using AIWC}
+
+	\begin{comment}
+	- AIWC performance prediction paper
+	- Important for workload schedulers and compiler writers.
+	- Used in this paper to demonstrate PSL is inherently linked to performance-prediction.
+
+	-- Methodology:
+		- Leave-one-out predictive accuracy
+		- CLgen?
+			- intro
+			- use to generate kernels for performance prediction
+
+	In~\cite{beauperformanceprediction} we show that AIWC features can be used to build performance prediction models for the runtime of OpenCL codes on diverse heterogeneous platforms.
+	This is useful for workload schedulers at supercomputer facilities and compiler engineers who must predict performance of codes on architectures without gathering performance data directly.
+	In this paper, we test whether the implemented parallel spatial locality metric is linked to code performance by building a predictive model
+
+	- Used random forest model (ranger).
+
+	- 
+	\end{comment}
+
 	\section{Conclusions and Future Work} \label{future work}
 
-	To the best of our knowledge, this work is the first to propose the use of architecture-independent metrics of parallel memory access to guide hardware-specific optimization. 
+	To the best of our knowledge, this work is the first to propose the use of architecture-independent metrics of parallel memory access to characterise hardware-specific optimization. 
 	We implemented and evaluated a new feature, relative local memory usage, to help characterize memory-based GPU code optimizations.
@@
 Future work would modify the presented metrics and develop new memory metrics with the specific intent of being fed to machine learning models to predict kernel performance on arbitrary and novel architectures.
-%FIXME: I want to say that we have been focusing on _guiding optimization_.
+%FIXME: I want to say that we have been focusing on _characterising optimization_.
 We can do more work in making the AIWC metrics better for _performance prediction_.

### 2020-03-20 paper/paper.tex
Finished adding content to paper. Ready for a final check and then figure mainpulation

@@
 \affiliation{%
+	\institution{Oak Ridge National Laboratory}
 	\institution{Australian National University}}
@@
 	Each thread of the kernel updates the $(globalRow, globalCol)$ element of matrix $C$ by computing the dot product of the corresponding row of $A$ and column of $B$.
+With square matrices of order $2048$, we observe an effective performance of 131 GFLOP/s on an NVIDIA Tesla P100 GPU.
 
@@
 	Appendix~\ref{coalescedAMultiply} contains the \texttt{coalescedA} kernel, which coalesces accesses to matrix $A$ by storing \emph{tiles}, or square blocks, of $A$'s values into shared memory (OpenCL \texttt{\_\_local} memory).
+	We see an improvement to 600 GFLOP/s is effective performance for order 2048 matrix multiplication.
 
@@
 	Appendix~\ref{coalescedABMultiply} contains the \texttt{coalescedAB} kernel which performs this optimization.
+	Performance is further improved to 614 GLOP/s after shared memory optimisation for Matrix B.
 
@@
 	\end{lstlisting}
+
+	We achieve 1061 GFLOP/s after applying all the recommended GPU memory optimisations to the simple matrix multiply kernel.
 
@@
 			\addplot [mark=*, color=brown] coordinates {%[mark=triangle*] coordinates {
-				(simple,46.920)
-				(coalA,8.783)
-				(coalAB,9.238)
-				(coalABT,5.307)
+				(simple,46.920) %118.94
+				(coalA,8.783) %635.39
+				(coalAB,9.238) %	604.10
+				(coalABT,5.307) %404.45356548
 				%(alignABT,5.272)
@@
 			\addplot [mark=star] coordinates {%[mark=square] coordinates {
-				(simple,130.682)
-				(coalA,28.602)
-				(coalAB,27.939)
-				(coalABT,16.179)
+				(simple,130.682) %131.431
+				(coalA,28.602) %600.506
+				(coalAB,27.939) %614.75
+				(coalABT,16.179) %1061.60
 				%(alignABT,16.071)
@@
 We then average the $n$-bits-dropped entropy summaries across thread groups to obtain the $n$-bits-dropped parallel spatial locality metrics for the kernel's execution.
+The theoretical maximum for an $n$-bits dropped parallel spatial locality value is $\log_2(group\_size)$.
+We divide the calculated average entropy values by $\log_2 (group\_size)$ to normalise the parallel spatial locality metric.
+This allows us to derive trends on the normalised metrics allowing comparison across separate OpenCL codes with different work distribution structures.
 
@@
 		\end{tikzpicture}
-		\caption{Parallel spatial locality metric for selected OpenDwarfs benchmark kernels}
+		\caption{Normalised parallel spatial locality metric for selected OpenDwarfs benchmark kernels}
 		\label{fig: opendwarfs parallel spatial locality}
@@
 	This is reflected in the drop in parallel spatial locality for the LUD kernels to approximately half the value at 0 bits skipped.
-Taking \texttt{LUD internal} as an example, the 0-bits skipped parallel spatial locality metric is 4.101, while the 10-bits skipped parallel spatial locality metric is 2.053.
-	The swift decline of the metric to 2.053 from 2 to 6-bits skipped parallel spatial locality indicates that approximately half the parallel memory accesses in \texttt{LUD internal} are highly localized.
+Taking \texttt{LUD internal} as an example, the 0-bits skipped parallel spatial locality metric is 0.5115, while the 10-bits skipped parallel spatial locality metric is 0.2561.
+	The swift decline of the metric to 0.2561 from 2 to 6-bits skipped parallel spatial locality indicates that approximately half the parallel memory accesses in \texttt{LUD internal} are highly localized.
 This occurs when work-items simultaneously access contiguous memory along a matrix row.
@@
 
-	-- Notes
+	 The recorded parallel spatial locality in GEM starts at a low value of 0.08604 and approaches the theoretical lower limit of 0 (2.18e-4 at 10 bits dropped).
+This indicates that almost all memory accesses made by the kernel are perfectly synchronized between OpenCL threads.
+From a memory performance standpoint, this is the best possible scenario as single memory accesses service all the active synchronised threads on the GPU.
 
-	 GEM: The recorded parallel spatial locality approaches the theoretical limit of 0 (0.0124 at 10 bits dropped).
-This indicates that almost all memory accesses made by the kernel are perfectly synchronized between OpenCL threads.
+	 Codes with a mix of favourable and unfavourable access patterns (typically involving two-dimensional matrices such as LUD and SRAD) will have a steep dropoff in PSL to about half of the starting PSL metric as $n$ increases, and the curve flattens out at that point.
 
-	 LUD: Codes with a mix of favourable and unfavourable access patterns (typically both dimensions of matrices) will have a steep dropoff to ~0.5 of the starting PSL metric, and flattens out at that point.
+	 In the extreme case, when parallel memory accesses are entirely disparate as for Needleman-Wuntsch, we observe a flat parallel spatial locality graph. 
+	 The worst possible memory access pattern: all threads in a workgroup accessing results in a flat PSL metric of 1. 
+	 The PSL metric for Needleman-Wuntsch is a flat line at 0.53. This indicates that parallel memory accesses are entirely disparate. 
+	 However, the workload imbalance across threads in each OpenCL work group eases pressure on the memory unit when not all threads are active.
+
+	 Codes with irregular or data-driven memory access patterns like CSR and BFS in the OpenDwarfs Suite have a slow dropoff in the PSL metric at low $n$ values.
+This is reflective of the mostly disparate parallel memory access pattern in these codes.
+The increasing steepness of the dropoff at higher $n$ in codes with data-driven memory accesses is reflective of the size of buffers being operated upon as the data structures used during computation cause localisation of memory accesses.
 
@@
 
-	While our work in this paper intended to guide optimization efforts, another potential use case for AIWC is providing architecture-independent performance predictions for OpenCL kernels by generating machine learning models based on AIWC metrics \cite{beauperformanceprediction}.
+	%While our work in this paper intended to guide optimization efforts, another potential use case for AIWC is providing architecture-independent performance predictions for OpenCL kernels by generating machine learning models based on AIWC metrics \cite{beauperformanceprediction}.
 Future work would modify the presented metrics and develop new memory metrics with the specific intent of being fed to machine learning models to predict kernel performance on arbitrary and novel architectures.

### 2020-03-20 paper/paper.tex
Fixing typos

@@
 	In section~\ref{results} we present the collected metric for selected benchmarks from the Extended OpenDwarfs benchmark suite to validate our methodology in this paper.
+	In section~\ref{guidelines} we summarize our findings and present guidelines for interpreting the new parallel spatial metric for OpenCL codes.
 	Finally, in section~\ref{future work} we discuss further avenues to extend our work and conclude.
@@
 	Appendix~\ref{coalescedAMultiply} contains the \texttt{coalescedA} kernel, which coalesces accesses to matrix $A$ by storing \emph{tiles}, or square blocks, of $A$'s values into shared memory (OpenCL \texttt{\_\_local} memory).
-	We see an improvement to 600 GFLOP/s is effective performance for order 2048 matrix multiplication.
+	We see an improvement to 600 GFLOP/s in effective performance for order 2048 matrix multiplication.
 
@@
 	Appendix~\ref{coalescedABMultiply} contains the \texttt{coalescedAB} kernel which performs this optimization.
-	Performance is further improved to 614 GLOP/s after shared memory optimisation for Matrix B.
+	Performance is further improved to 614 GLOP/s after this optimisation.
 
@@
 	Aggregate metrics of the kind presented by AIWC necessarily present a simplified view of program behaviour, omitting many details.
-	Different ways of aggregating program measurements lead to different features of program execution being emphasised in the final metrics.
+	Different ways of aggregating program measurements lead to different features of program execution being emphasized in the final metrics.
 	For example, the calculation of memory address entropy described in section~\ref{related work} relies only on the frequency distribution of memory accesses to all addresses accessed by the kernel, and discards temporal information.
@@
 The theoretical maximum for an $n$-bits dropped parallel spatial locality value is $\log_2(group\_size)$.
-We divide the calculated average entropy values by $\log_2 (group\_size)$ to normalise the parallel spatial locality metric.
-This allows us to derive trends on the normalised metrics allowing comparison across separate OpenCL codes with different work distribution structures.
+We divide the calculated average entropy values by $\log_2 (group\_size)$ to normalize the parallel spatial locality metric.
+This allows us to derive trends on the normalized metrics allowing comparison across separate OpenCL codes with different work distribution structures.
 
@@
 
-	\section{Guidelines: Using the Parallel Spatial Locality Metric}
+	\section{Guidelines: Using the Parallel Spatial Locality Metric} \label{guidelines}
 

### 2020-03-21 paper/paper.tex
Fixed typos and restructured figures

@@
 	In section~\ref{results} we present the collected metric for selected benchmarks from the Extended OpenDwarfs benchmark suite to validate our methodology in this paper.
-	In section~\ref{guidelines} we summarize our findings and present guidelines for interpreting the new parallel spatial metric for OpenCL codes.
+	In section~\ref{guidelines} we summarize our findings and present guidelines for interpreting the new parallel spatial locality metric for OpenCL codes.
 	Finally, in section~\ref{future work} we discuss further avenues to extend our work and conclude.
@@
 	\section{Motivating Example} \label{motivating example}
+
+	\begin{figure*}[h!]
+		\centering
+		\subfloat{
+			\centering
+			\begin{tikzpicture}
+			\begin{axis}[symbolic x coords={simple,coalA,coalAB,coalABT,alignABT},
+			xtick={simple,coalA,coalAB,coalABT},%alignABT},
+			xticklabel style={text height=2ex},
+			ymin=0,
+			ylabel={Execution time (ms)}, xlabel={Kernel},
+			ylabel style={at={(axis description cs:0.1,.5)},anchor=south},
+			legend entries={$N=80\ \ \ \ $,$N=256\ \ $},
+			legend style={at={(0.6,0.45)},anchor=west}
+			]
+
+			\addplot coordinates {%[mark=*] coordinates {
+				(simple,.031)
+				(coalA,.019)
+				(coalAB,.020)
+				(coalABT,.018)
+				%(alignABT,.018)
+			};
+			\addplot coordinates {%[mark=star] coordinates {
+				(simple,.325)
+				(coalA,.075)
+				(coalAB,.081)
+				(coalABT,.055)
+				%(alignABT,.055)
+			};	
+			\end{axis}
+			\end{tikzpicture} 
+		}
+		\subfloat{
+			\begin{tikzpicture}
+			\begin{axis}[symbolic x coords={simple,coalA,coalAB,coalABT},%alignABT},
+			xtick={simple,coalA,coalAB,coalABT},%alignABT},
+			xticklabel style={text height=2ex},
+			ymin=0,
+			ylabel={Execution time (ms)}, xlabel={Kernel},
+			ylabel style={at={(axis description cs:0.1,.5)},anchor=south},
+			legend entries={$N=1408$,$N=2048$},
+			legend style={at={(0.6,0.45)},anchor=west},
+			scaled ticks=false
+			]
+
+			\addplot [mark=*, color=brown] coordinates {%[mark=triangle*] coordinates {
+				(simple,46.920) %118.94
+				(coalA,8.783) %635.39
+				(coalAB,9.238) %	604.10
+				(coalABT,5.307) %404.45356548
+				%(alignABT,5.272)
+			};
+			\addplot [mark=star] coordinates {%[mark=square] coordinates {
+				(simple,130.682) %131.431
+				(coalA,28.602) %600.506
+				(coalAB,27.939) %614.75
+				(coalABT,16.179) %1061.60
+				%(alignABT,16.071)
+			};	
+			\end{axis}
+			\end{tikzpicture} 
+		}
+		\caption{Execution time of matrix multiplication kernels (NVIDIA Tesla P100)}
+		\label{fig: matmul runtime}
+	\end{figure*}
 
@@
 	Appendix~\ref{coalescedABMultiply} contains the \texttt{coalescedAB} kernel which performs this optimization.
-	Performance is further improved to 614 GLOP/s after this optimisation.
+	Performance is further improved to 614 GLOP/s after this optimization.
 
@@
 
-	We achieve 1061 GFLOP/s after applying all the recommended GPU memory optimisations to the simple matrix multiply kernel.
+	We achieve 1061 GFLOP/s after applying all the recommended GPU memory optimizations to the simple matrix multiply kernel.
 
@@
 
-	\flushbottom
 	\section{Related Work} \label{related work}
@@
 
-	\begin{figure*}[ht!]
+	\section{Non-Parallel Memory Metrics} \label{method}
+
+	\begin{table*}[h!]
 		\centering
-		\subfloat{
-			\centering
-			\begin{tikzpicture}
-			\begin{axis}[symbolic x coords={simple,coalA,coalAB,coalABT,alignABT},
-			xtick={simple,coalA,coalAB,coalABT},%alignABT},
-			xticklabel style={text height=2ex},
-			ymin=0,
-			ylabel={Execution time (ms)}, xlabel={Kernel},
-			ylabel style={at={(axis description cs:0.1,.5)},anchor=south},
-			legend entries={$N=80\ \ \ \ $,$N=256\ \ $},
-			legend style={at={(0.6,0.45)},anchor=west}
-			]
-
-			\addplot coordinates {%[mark=*] coordinates {
-				(simple,.031)
-				(coalA,.019)
-				(coalAB,.020)
-				(coalABT,.018)
-				%(alignABT,.018)
-			};
-			\addplot coordinates {%[mark=star] coordinates {
-				(simple,.325)
-				(coalA,.075)
-				(coalAB,.081)
-				(coalABT,.055)
-				%(alignABT,.055)
-			};	
-			\end{axis}
-			\end{tikzpicture} 
-		}
-		\subfloat{
-			\begin{tikzpicture}
-			\begin{axis}[symbolic x coords={simple,coalA,coalAB,coalABT},%alignABT},
-			xtick={simple,coalA,coalAB,coalABT},%alignABT},
-			xticklabel style={text height=2ex},
-			ymin=0,
-			ylabel={Execution time (ms)}, xlabel={Kernel},
-			ylabel style={at={(axis description cs:0.1,.5)},anchor=south},
-			legend entries={$N=1408$,$N=2048$},
-			legend style={at={(0.6,0.45)},anchor=west},
-			scaled ticks=false
-			]
-
-			\addplot [mark=*, color=brown] coordinates {%[mark=triangle*] coordinates {
-				(simple,46.920) %118.94
-				(coalA,8.783) %635.39
-				(coalAB,9.238) %	604.10
-				(coalABT,5.307) %404.45356548
-				%(alignABT,5.272)
-			};
-			\addplot [mark=star] coordinates {%[mark=square] coordinates {
-				(simple,130.682) %131.431
-				(coalA,28.602) %600.506
-				(coalAB,27.939) %614.75
-				(coalABT,16.179) %1061.60
-				%(alignABT,16.071)
-			};	
-			\end{axis}
-			\end{tikzpicture} 
-		}
-		\caption{Execution time of matrix multiplication kernels (NVIDIA Tesla P100)}
-		\label{fig: matmul runtime}
-	\end{figure*}
-
-	\section{Non-Parallel Memory Metrics} \label{method}
+		\begin{tabular}[t]{lrrrrr}
+			\toprule
+			& simple & coalescedA & coalescedAB & coalescedABT \\ \midrule %& alignedABT \\ \midrule 
+			Total memory footprint        & 196608 & 196608     & 196608      & 196608      \\ %\hline
+			90\% Memory Footprint         & 118196 & 56176      & 489         & 489         \\ %\hline
+			Global MAE                    & 17.02  & 13.18      & 9.78        & 9.78        \\ %\hline
+			LMAE \#bits=3                 & 16.02  & 12.18      & 8.78        & 8.78        \\ %\hline
+			LMAE \#bits=10                & 9.02   & 5.18       & 1.78        & 1.78        \\ %\hline
+			Relative Local Memory Usage   & 0      & 0.50       & 0.94       &  0.94        \\ \bottomrule
+		\end{tabular}
+		\caption{Selection of AIWC\cite{beauaiwc} metrics for 256 $\times$ 256 matrix multiplication}
+		\label{matmul aiwc table}
+	\end{table*}
 
@@
 	%Note that a matrix size of 80 is small -- overheads attached to optimization techniques may be causing the observed decrease in performance as some optimizations are performed.
-
-	\begin{table*}[h!]
-	\centering
-	\begin{tabular}[t]{lrrrrr}
-		\toprule
-		& simple & coalescedA & coalescedAB & coalescedABT \\ \midrule %& alignedABT \\ \midrule 
-		Total memory footprint        & 196608 & 196608     & 196608      & 196608      \\ %\hline
-		90\% Memory Footprint         & 118196 & 56176      & 489         & 489         \\ %\hline
-		Global MAE                    & 17.02  & 13.18      & 9.78        & 9.78        \\ %\hline
-		LMAE \#bits=3                 & 16.02  & 12.18      & 8.78        & 8.78        \\ %\hline
-		LMAE \#bits=10                & 9.02   & 5.18       & 1.78        & 1.78        \\ %\hline
-		Relative Local Memory Usage   & 0      & 0.50       & 0.94       &  0.94        \\ \bottomrule
-	\end{tabular}
-	\caption{Selection of AIWC\cite{beauaiwc} metrics for 256 $\times$ 256 matrix multiplication}
-	\label{matmul aiwc table}
-	\end{table*}
 
@@
 
+	\flushbottom
+
 	\section{A Parallel Spatial Locality Metric} \label{discussion}
-
-	Aggregate metrics of the kind presented by AIWC necessarily present a simplified view of program behaviour, omitting many details.
-	Different ways of aggregating program measurements lead to different features of program execution being emphasized in the final metrics.
-	For example, the calculation of memory address entropy described in section~\ref{related work} relies only on the frequency distribution of memory accesses to all addresses accessed by the kernel, and discards temporal information.
-	The order of sequential memory accesses performed by each thread, as well the relationship between work items in an OpenCL work group, are both vital in accurately characterizing parallel codes.
-
-	We propose a new architecture-independent metric, \emph{parallel spatial locality}, to measure memory access patterns in parallel programs.
-	The proposed metric is inspired by CuMAPz' direct approaches to measuring optimization specific characteristics of CUDA codes.
-
-	During simulation, AIWC collects a list of all memory accesses by each thread of execution.
-	In the OpenCL programming model, threads within a work groups execute the same code and share access to \textit{local memory}.
-	We can group together memory accesses of threads in a work group at each logical time step in the symbolic execution of the code.
-	On a GPU, memory accesses executed at the same time by different threads in a work group are likely to interact, determining the extent of memory access coalescing and bank conflicts.
 
@@
 		\end{tikzpicture}
-		\caption{Parallel spatial locality metric obtained from AIWC for matrix multiply kernels for 256 $\times$ 256 matrix multiplication}
+		\caption{Parallel spatial locality metric for matrix multiply kernels for 256 $\times$ 256 matrix multiplication}
 		\label{fig: matmul parallel spatial locality}
 	\end{figure*}
+
+	Aggregate metrics of the kind presented by AIWC necessarily present a simplified view of program behaviour, omitting many details.
+	Different ways of aggregating program measurements lead to different features of program execution being emphasized in the final metrics.
+	For example, the calculation of memory address entropy described in section~\ref{related work} relies only on the frequency distribution of memory accesses to all addresses accessed by the kernel, and discards temporal information.
+	The order of sequential memory accesses performed by each thread, as well the relationship between work items in an OpenCL work group, are both vital in accurately characterizing parallel codes.
+
+	We propose a new architecture-independent metric, \emph{parallel spatial locality}, to measure memory access patterns in parallel programs.
+	The proposed metric is inspired by CuMAPz' direct approaches to measuring optimization specific characteristics of CUDA codes.
+
+	During simulation, AIWC collects a list of all memory accesses by each thread of execution.
+	In the OpenCL programming model, threads within a work groups execute the same code and share access to \textit{local memory}.
+	We can group together memory accesses of threads in a work group at each logical time step in the symbolic execution of the code.
+	On a GPU, memory accesses executed at the same time by different threads in a work group are likely to interact, determining the extent of memory access coalescing and bank conflicts.
 
@@
 		\end{tikzpicture}
-		\caption{Normalised parallel spatial locality metric for selected OpenDwarfs benchmark kernels}
+		\caption{Normalized parallel spatial locality metric for selected OpenDwarfs benchmark kernels}
 		\label{fig: opendwarfs parallel spatial locality}
@@
 	Thus we observe that while cache hit rates are typically low on this benchmark, particularly on GPUs \cite{krommydas2016opendwarfs}, GPUs can hide the latency of global memory accesses through memory access coalescing to some extent.
-
 	\subsection{Graph Traversal: BFS}
@@
 	This leads to poor parallel spatial locality. Thus, we see a slow dropoff in parallel spatial locality for \texttt{BFS kernel 1}.
-
 	\subsection{Sparse Linear Algebra: CSR}
@@
 Figure \ref{fig: opendwarfs parallel spatial locality} shows the gradual decline in parallel spatial locality as the number of bits dropped is increased.
-
 	\section{Guidelines: Using the Parallel Spatial Locality Metric} \label{guidelines}
@@
 This indicates that almost all memory accesses made by the kernel are perfectly synchronized between OpenCL threads.
-From a memory performance standpoint, this is the best possible scenario as single memory accesses service all the active synchronised threads on the GPU.
+From a memory performance standpoint, this is the best possible scenario as single memory accesses service all the active synchronized threads on the GPU.
 
-	 Codes with a mix of favourable and unfavourable access patterns (typically involving two-dimensional matrices such as LUD and SRAD) will have a steep dropoff in PSL to about half of the starting PSL metric as $n$ increases, and the curve flattens out at that point.
+	 Codes with a mix of favourable and unfavourable access patterns (typically involving two-dimensional matrices such as in LUD and SRAD) will have a steep dropoff in PSL to about half of the starting PSL metric as $n$ increases, and the curve flattens out at that point.
 
-	 In the extreme case, when parallel memory accesses are entirely disparate as for Needleman-Wuntsch, we observe a flat parallel spatial locality graph. 
-	 The worst possible memory access pattern: all threads in a workgroup accessing results in a flat PSL metric of 1. 
-	 The PSL metric for Needleman-Wuntsch is a flat line at 0.53. This indicates that parallel memory accesses are entirely disparate. 
+	 In the extreme case, when parallel memory accesses are entirely disparate as for Needleman-Wunsch, we observe a flat parallel spatial locality graph. 
+	 The worst possible memory access pattern occurs when all threads in a workgroup access disparate memory locations. This results in a flat PSL metric of 1. 
+	 The PSL metric for Needleman-Wunsch is a flat line at 0.53. This indicates that parallel memory accesses are entirely disparate. 
 	 However, the workload imbalance across threads in each OpenCL work group eases pressure on the memory unit when not all threads are active.
@@
 This is reflective of the mostly disparate parallel memory access pattern in these codes.
-The increasing steepness of the dropoff at higher $n$ in codes with data-driven memory accesses is reflective of the size of buffers being operated upon as the data structures used during computation cause localisation of memory accesses.
+The increasing steepness of the dropoff at higher $n$ in codes with data-driven memory accesses is reflective of the size of buffers being operated upon as the data structures used during computation cause localization of memory accesses.
 
@@
 	\end{comment}
-
 	\section{Conclusions and Future Work} \label{future work}
 
-	To the best of our knowledge, this work is the first to propose the use of architecture-independent metrics of parallel memory access to characterise hardware-specific optimization. 
+	To the best of our knowledge, this work is the first to propose the use of architecture-independent metrics of parallel memory access to characterize hardware-specific optimization. 
 	We implemented and evaluated a new feature, relative local memory usage, to help characterize memory-based GPU code optimizations.

### 2020-03-21 paper/paper.tex
changed ize, consistant hyphen in work-item/work-group, and removed a few stray commas

--- 2020-03-20 paper/paper.tex
@@
 	We propose a new metric of parallel spatial locality -- the closeness of memory accesses simultaneously issued by OpenCL work-items (threads).
-	We implement the parallel spatial locality metric in the AIWC framework, and analyse gathered results on matrix multiply and the Extended OpenDwarfs OpenCL benchmarks.
+	We implement the parallel spatial locality metric in the AIWC framework and analyse gathered results on matrix multiply and the Extended OpenDwarfs OpenCL benchmarks.
 	The differences in the observed parallel spatial locality metric across implementations of matrix multiply reflect the optimizations performed.
@@
 	The OpenCL programming language is designed to support modern HPC software engineers in writing code that executes on multiple hardware targets. 
-	%It is possible for an engineer to write highly thread-divergent logical analysis code aimed at traditional CPUs; parallelisable data crunching kernels aimed at GPUs; and code to satisfy specialized computational requirements aimed at FPGAs -- all in a single code base using OpenCL.
-	%The language provides support for individual \emph{kernels}, portions of parallelisable codes that make up a HPC code base, to be executed independently on diverse hardware targets such as traditional CPUs, GPUs and FPGAs \cite{khronosopenclworkinggroup2011opencl}.
+	%It is possible for an engineer to write highly thread-divergent logical analysis code aimed at traditional CPUs; parallelizable data crunching kernels aimed at GPUs; and code to satisfy specialized computational requirements aimed at FPGAs -- all in a single code base using OpenCL.
+	%The language provides support for individual \emph{kernels}, portions of parallelizable codes that make up a HPC code base, to be executed independently on diverse hardware targets such as traditional CPUs, GPUs and FPGAs \cite{khronosopenclworkinggroup2011opencl}.
 	This gives HPC software developers greater flexibility by allowing codes aimed at a range of hardware targets to be written in a single programming language environment.
@@
 	To mitigate this latency, modern CPU designs have employed a wide range of cache technologies to reduce main memory accesses using the principle of \emph{spatial locality}: the observation that data that a program accesses close together in time tend to also be close together in memory.
-	On the other hand, GPUs rely on hardware multithreading to hide memory latency, and their architectures favour ALU capability over sophisticated logic to manage a cache hierarchy and out-of-order execution.
+	On the other hand, GPUs rely on hardware multi-threading to hide memory latency, and their architectures favour ALU capability over sophisticated logic to manage a cache hierarchy and out-of-order execution.
 As a result, the performance of kernels on CPUs and GPUs alike is strongly dependent on memory access patterns intrinsic to the code.
@@
 	Appendix~\ref{coalescedAMultiply} contains the \texttt{coalescedA} kernel, which coalesces accesses to matrix $A$ by storing \emph{tiles}, or square blocks, of $A$'s values into shared memory (OpenCL \texttt{\_\_local} memory).
-	We see an improvement to 600 GFLOP/s in effective performance for order 2048 matrix multiplication.
+	We see an improvement to 600 GFLOP/s in the effective performance for order 2048 matrix multiplication.
 
@@
 	While in the previous kernel each thread reads only a single element of matrix $A$ from global memory, each thread reads a full column of matrix $B$.
-	The repeated reads of elements of matrix $B$ can be shared between threads in a work group by reading tiles from matrix $B$ into shared memory.
+	The repeated reads of elements of matrix $B$ can be shared between threads in a work-group by reading tiles from matrix $B$ into shared memory.
 	Appendix~\ref{coalescedABMultiply} contains the \texttt{coalescedAB} kernel which performs this optimization.
@@
 
-	During simulation of the target code, CuMAPz records all memory accesses in various buffers to simulate the global and shared memories on NVIDIA GPUs.
+	During the simulation of the target code, CuMAPz records all memory accesses in various buffers to simulate the global and shared memories on NVIDIA GPUs.
 	Using this detailed simulation data, CuMAPz can estimate the performance-critical (i) shared memory data reuse profit, (ii) profit from coalesced access (iii) memory channel skew cost and (iv) bank conflict cost characteristics of the target code.
@@
 To calculate this, AIWC collects a frequency distribution of all non-register memory accesses by all threads in the target kernel.
-Using the collected frequency distribution, AIWC calculates 10 separate local memory address entropy (LMAE) values according to increasing number of least significant bits (LSB) skipped using
+Using the collected frequency distribution, AIWC calculates 10 separate local memory address entropy (LMAE) values according to increasing the number of least significant bits (LSB) skipped, using
 	%in the following fashion.
@@
 
-	We first consider AIWC memory metrics that do not take into account the interaction between work items executing in parallel.
+	We first consider AIWC memory metrics that do not take into account the interaction between work-items executing in parallel.
 	We compare these metrics for the incrementally optimized versions of the matrix multiplication example presented in section~\ref{motivating example}.
@@
 Increased utilization of local memory in the optimized kernels means that the local memory buffers make up a greater proportion of total memory accesses in the program.
-As the local memory buffers are small and reused within a workgroup, the memory footprint of the local memory buffers is also smaller.
+As the local memory buffers are small and reused within a work-group, the memory footprint of the local memory buffers is also smaller.
 
@@
 	For example, the calculation of memory address entropy described in section~\ref{related work} relies only on the frequency distribution of memory accesses to all addresses accessed by the kernel, and discards temporal information.
-	The order of sequential memory accesses performed by each thread, as well the relationship between work items in an OpenCL work group, are both vital in accurately characterizing parallel codes.
+	The order of sequential memory accesses performed by each thread, as well the relationship between work-items in an OpenCL work-group, are both vital in accurately characterizing parallel codes.
 
@@
 	During simulation, AIWC collects a list of all memory accesses by each thread of execution.
-	In the OpenCL programming model, threads within a work groups execute the same code and share access to \textit{local memory}.
-	We can group together memory accesses of threads in a work group at each logical time step in the symbolic execution of the code.
-	On a GPU, memory accesses executed at the same time by different threads in a work group are likely to interact, determining the extent of memory access coalescing and bank conflicts.
+	In the OpenCL programming model, threads within a work-group execute the same code and share access to \textit{local memory}.
+	We can group together memory accesses of threads in a work-group at each logical time step in the symbolic execution of the code.
+	On a GPU, memory accesses executed at the same time by different threads in a work-group are likely to interact, determining the extent of memory access coalescing and bank conflicts.
 
@@
 
-	\textit{Record}: we first record memory accesses performed by each thread in an OpenCL work group as described above to achieve a global ordering of all memory accesses performed by the group.
+	\textit{Record}: we first record memory accesses performed by each thread in an OpenCL work-group as described above to achieve a global ordering of all memory accesses performed by the group.
 	This ordering is collected in the form of logical timestamps ($t_0 .. t_{last}$) at which memory accesses occur.
@@
 	\textit{Calculate}: for each timestamp $t = t_0 ..
-t_{last}$, calculate the $n$-bits-dropped entropy of memory addresses accessed by all threads in a work group within the timestamp $t$.
+t_{last}$, calculate the $n$-bits-dropped entropy of memory addresses accessed by all threads in a work-group within the timestamp $t$.
 Here $n$ can range between 0 and 10 as was the case for $LMAE$.
@@
 	LUD in OpenDwarfs is a program to decompose an input $N\times N$ matrix as the product of one lower and one upper diagonal matrix. 
-	Memory access patterns in a dense linear algebra workload such as LUD are typically highly regular and deterministic for each OpenCL work item, based on the matrix dimension and offset parameters.
+	Memory access patterns in a dense linear algebra workload such as LUD are typically highly regular and deterministic for each OpenCL work-item, based on the matrix dimension and offset parameters.
 The OpenDwarfs implementation of LUD \cite{opendwarfs2020head} splits the LUD computation into three kernels: \texttt{LUD perimeter} and \texttt{LUD diagonal} kernels spawn work-items in a single work dimension; while \texttt{LUD internal} is decomposed in both dimensions of the input.
@@
 	The computation of each element in the similarity matrix depends on its west, north and north-west neighbours.
-	This dependency enforces a wavefront computation pattern, travelling along the main diagonal of the matrix.
-Each iteration of the kernels computes over an antidiagonal of the matrix, starting at its top left corner, and finishing at its bottom right corner.
+	This dependency enforces a wave-front computation pattern, travelling along the main diagonal of the matrix.
+Each iteration of the kernels computes over an anti-diagonal of the matrix, starting at its top left corner, and finishing at its bottom right corner.
 
-	In particular, the wavefront pattern of computations causes each work-item to request distant memory addresses in a parallel fashion.
+	In particular, the wave-front pattern of computations causes each work-item to request distant memory addresses in a parallel fashion.
 Thus, parallel memory requests into the input matrix prohibit locality.
-We note the lack of any dropoff in parallel spatial locality as the number of bits dropped increases.
+We note the lack of any drop-off in parallel spatial locality as the number of bits dropped increases.
 This rightly indicates that memory addresses accessed at each logical timestamp are very distant.
@@
 	The program structure involves multiple levels of pointer-chasing and thus the precise memory addresses accessed are data-dependent. 
-	This leads to poor parallel spatial locality. Thus, we see a slow dropoff in parallel spatial locality for \texttt{BFS kernel 1}.
+	This leads to poor parallel spatial locality. Thus, we see a slow drop-off in parallel spatial locality for \texttt{BFS kernel 1}.
 
@@
 
-	In the OpenDwarfs implementation, each row of the input sparse matrix is assigned to a separate OpenCL work item. 
+	In the OpenDwarfs implementation, each row of the input sparse matrix is assigned to a separate OpenCL work-item.
 	The locations of matrix data read by each work-item are dependent on the number and position of non-zero values in the sparse-matrix, which are decided by the values in $Aj$ and $Ap$. 
@@
 This indicates that almost all memory accesses made by the kernel are perfectly synchronized between OpenCL threads.
-From a memory performance standpoint, this is the best possible scenario as single memory accesses service all the active synchronised threads on the GPU.
+From a memory performance standpoint, this is the best possible scenario as single memory accesses service all the active synchronized threads on the GPU.
 
-	 Codes with a mix of favourable and unfavourable access patterns (typically involving two-dimensional matrices such as LUD and SRAD) will have a steep dropoff in PSL to about half of the starting PSL metric as $n$ increases, and the curve flattens out at that point.
+	 Codes with a mix of favourable and unfavourable access patterns (typically involving two-dimensional matrices such as LUD and SRAD) will have a steep drop-off in PSL to about half of the starting PSL metric as $n$ increases, and the curve flattens out at that point.
 
-	 In the extreme case, when parallel memory accesses are entirely disparate as for Needleman-Wuntsch, we observe a flat parallel spatial locality graph. 
-	 The worst possible memory access pattern: all threads in a workgroup accessing results in a flat PSL metric of 1. 
-	 The PSL metric for Needleman-Wuntsch is a flat line at 0.53. This indicates that parallel memory accesses are entirely disparate. 
-	 However, the workload imbalance across threads in each OpenCL work group eases pressure on the memory unit when not all threads are active.
+	 In the extreme case, when parallel memory accesses are entirely disparate as for Needleman–Wunsch, we observe a flat parallel spatial locality graph.
+	 The worst possible memory access pattern: all threads in a work-group accessing results in a flat PSL metric of 1.
+	 The PSL metric for Needleman–Wunsch is a flat line at 0.53. This indicates that parallel memory accesses are entirely disparate.
+	 However, the workload imbalance across threads in each OpenCL work-group eases pressure on the memory unit when not all threads are active.
 
-	 Codes with irregular or data-driven memory access patterns like CSR and BFS in the OpenDwarfs Suite have a slow dropoff in the PSL metric at low $n$ values.
+	 Codes with irregular or data-driven memory access patterns like CSR and BFS in the OpenDwarfs Suite have a slow drop-off in the PSL metric at low $n$ values.
 This is reflective of the mostly disparate parallel memory access pattern in these codes.
-The increasing steepness of the dropoff at higher $n$ in codes with data-driven memory accesses is reflective of the size of buffers being operated upon as the data structures used during computation cause localisation of memory accesses.
+The increasing steepness of the drop-off at higher $n$ in codes with data-driven memory accesses is reflective of the size of buffers being operated upon as the data structures used during computation cause localisation of memory accesses.
 

### 2020-03-21 paper/paper.tex

@@
 	The OpenCL programming language is designed to support modern HPC software engineers in writing code that executes on multiple hardware targets. 
-	%It is possible for an engineer to write highly thread-divergent logical analysis code aimed at traditional CPUs; parallelizable data crunching kernels aimed at GPUs; and code to satisfy specialized computational requirements aimed at FPGAs -- all in a single code base using OpenCL.
-	%The language provides support for individual \emph{kernels}, portions of parallelizable codes that make up a HPC code base, to be executed independently on diverse hardware targets such as traditional CPUs, GPUs and FPGAs \cite{khronosopenclworkinggroup2011opencl}.
+	%It is possible for an engineer to write highly thread-divergent logical analysis code aimed at traditional CPUs; parallelisable data crunching kernels aimed at GPUs; and code to satisfy specialized computational requirements aimed at FPGAs -- all in a single code base using OpenCL.
+	%The language provides support for individual \emph{kernels}, portions of parallelisable codes that make up a HPC code base, to be executed independently on diverse hardware targets such as traditional CPUs, GPUs and FPGAs \cite{khronosopenclworkinggroup2011opencl}.
 	This gives HPC software developers greater flexibility by allowing codes aimed at a range of hardware targets to be written in a single programming language environment.
@@
 
-	We aim to capture the essence of vendor-recommended optimizations for target architectures in our metrics.
-	We demonstrate the optimization strategies using an OpenCL kernel which multiplies square matrices of order $N$.
-	We start with a simple unoptimized kernel, and then improve the memory access patterns of this kernel by incrementally performing the optimizations recommended in the CUDA Optimization Handbook \cite{cudaoptimisation} for NVIDIA GPUs.
-	These incrementally optimized OpenCL codes are then analysed using AIWC to determine the accuracy in profiling favourable memory access patterns of the current framework and the proposed extensions to AIWC.
-
-	\subsection{Simple Unoptimized Matrix Multiplication}
-
-	The unoptimized matrix multiplication kernel presented in Appendix~\ref{simpleMultiply} is used as a baseline to validate the performance improvements measured from each following optimization.
-	Each thread of the kernel updates the $(globalRow, globalCol)$ element of matrix $C$ by computing the dot product of the corresponding row of $A$ and column of $B$.
-With square matrices of order $2048$, we observe an effective performance of 131 GFLOP/s on an NVIDIA Tesla P100 GPU.
-
-	\subsection{Using Shared Memory to Coalesce Global Memory Access to Matrix A}
-
-	We first notice that the number of global memory accesses to matrices A and B increases as $O(N^3)$ with respect to matrix size $N$.
-	Global memory is typically located off-chip and accesses induce large delays.
-	NVIDIA GPUs coalesce global memory loads and stores issued within thread-groups into as few DRAM transactions as possible.
-	Multiple global memory loads and stores are coalesced into a single transaction when certain device-specific conditions are met. 
-	On most NVIDIA GPUs, data accesses are coalesced when multiple requests are made for memory locations from the same cache line in global memory~\cite{cudamanual}.
-	Appendix~\ref{coalescedAMultiply} contains the \texttt{coalescedA} kernel, which coalesces accesses to matrix $A$ by storing \emph{tiles}, or square blocks, of $A$'s values into shared memory (OpenCL \texttt{\_\_local} memory).
-	We see an improvement to 600 GFLOP/s in the effective performance for order 2048 matrix multiplication.
-
-	% 9.2.1.2: Misaligned access pattern
-
-	\subsection{Using Shared Memory to store Tiles of Matrix B}
-
-	The code is further optimized by improving the locality of reads from Matrix $B$.
-	While in the previous kernel each thread reads only a single element of matrix $A$ from global memory, each thread reads a full column of matrix $B$.
-	The repeated reads of elements of matrix $B$ can be shared between threads in a work-group by reading tiles from matrix $B$ into shared memory.
-	Appendix~\ref{coalescedABMultiply} contains the \texttt{coalescedAB} kernel which performs this optimization.
-	Performance is further improved to 614 GLOP/s after this optimisation.
-
-	\subsection{Optimizing Handling of Shared Memory}
-
-	NVIDIA shared memory is divided into multiple banks -- stored in independent memory modules -- to allow parallel memory access.
-	Bank conflicts occur when shared memory in the same bank is accessed concurrently.
-	The code in \texttt{coalescedAB} kernel is further optimized by implicitly transposing tiles of Matrix $A$ while loading from global memory.
-This improves memory bank utilization during reads to shared memory.
-The \texttt{coalescedABT} kernel demonstrates this optimization by modifying lines 20,21 of \texttt{coalescedAB} to:
-
-	\begin{lstlisting}[firstnumber=20]
-		ASub[localCol][localRow] = A[tiledRow];
-		BSub[localRow][localCol] = B[tiledCol];
-	\end{lstlisting}
-
-	We achieve 1061 GFLOP/s after applying all the recommended GPU memory optimisations to the simple matrix multiply kernel.
-
-	\begin{comment}
-	\subsection{Alignment of Memory Allocation}
-
-	Memory access alignment is important to best utilize all parts of GPU memory architecture.
-Global memory buffer alignment can allow threads to access blocks of global memory aligned to the nearest cache line.
-This enables coalescing of memory accesses.
-If buffers are misaligned, parallel memory requests may cross over cache lines, and may double the number of slow global memory accesses needed as demonstrated in figures \ref{fig: Simple memory alignment} and \ref{fig: Simple memory misalignment}.
-
-	\begin{figure}[h]
-		\includegraphics[width=\linewidth]{aligned_access_coalescing.pdf}
-		\caption{All threads access memory aligned to nearest cache line in parallel \cite{cudaoptimisation}}
-		\label{fig: Simple memory alignment}
-	\end{figure}
-
-	\begin{figure}[h]
-		\includegraphics[width=\linewidth]{misaligned_access_coalescing.pdf}
-		\caption{Unaligned sequential memory addresses fit in two cache lines \cite{cudaoptimisation}}
-		\label{fig: Simple memory misalignment}
-	\end{figure}
-
-	A similar principle applies when using shared memory.
-Bank conflicts may be reduced by aligning allocations of shared memory buffers.
-The \texttt{alignedABT} kernel improves the alignment of shared memory tiles in \texttt{coalescedABT}.
-	An arbitrary large alignment value of 4096 is chosen as it is larger than the cache line size of modern hardware and attributed to the local array declarations in the existing code as shown below.
-
-	\begin{lstlisting}
-		__local float aTile[TILE_DIM][TILE_DIM] __attribute__ ((aligned(4096)));
-		__local float bTile[TILE_DIM][TILE_DIM] __attribute__ ((aligned(4096)));
-	\end{lstlisting}
-	\end{comment}
-	All the examples above are an incomplete listing of optimization strategies developers can apply to code targeted at NVIDIA GPUs.
-	In section~\ref{method}, we will compare the collected AIWC metrics with the expected effect of each optimization to examine how effectively our methodology uncovers underlying bottlenecks and guides optimization efforts.
-
-	\flushbottom
-	\section{Related Work} \label{related work}
-
-	Hoste and Eeckout \cite{hoste2007microarchitecture} show that while conventional architecture-dependent characteristics are useful in locating performance bottlenecks, they can hide the underlying, inherent program behaviour causing performance bottlenecks.
-	A conceptual understanding of the performance characteristics of complex codes is necessary for the programmer to effectively optimize these codes.
-	Architecture-dependent characteristics typically include instructions per cycle (IPC), cache and branch prediction miss rates, page faults and DRAM bus data transfer rates.
-These are typically collected using hardware performance counters available on most target architectures.
-These performance counters do not serve to guide optimization beyond highlighting potential bottlenecks \cite{hoste2007microarchitecture, ganesan2008performance}.
-	Further, in many cases, architecture-dependent characteristics cannot be directly correlated to specific code patterns.
-	For example, the causes of high cache miss rates in the execution of a program are complex and depend on microarchitecture specific features such as cache size, prefetch behaviour and cache placement policies.
-	A HPC developer tasked with optimizing code for a given hardware target would benefit from architecture-independent metrics of the code that can be used to measure the effect of code modifications.
-    These metrics would help to guide the developer in both finding and fixing performance bottlenecks.
-
-	To address the limitations of conventional microarchitecture-dependent characteristics, Hoste and Eeckout~\cite{hoste2007microarchitecture} developed the Microarchitecture-Independent Workload Characterization tool (MICA).
-	They observed that performance counter-based approaches to profiling codes often failed to find underlying program features that map to improved or worsened usage of performance-critical hardware features of the target architecture.
-	The MICA framework is a holistic characterization tool, and thus collects features including instruction mix, instruction-level parallelism, register traffic, data stream strides and branch predictability.
-
-	Of these metrics, data stream strides are of particular interest in memory access pattern profiling.
-	MICA's stride length metric measures the distance between consecutive memory accesses in a single-threaded application.
-	For CPU architectures running single-threaded applications, this metric correlates to the spatial locality of memory accesses -- a measure of how closely bunched are memory access in nearby times.
-	This is directly correlated to cache reuse rates, critical to code performance on CPUs \cite{inteloptimisation}. %Intel optimization manual.
-
-	The MICA approach was tailored for single-threaded applications as the metrics collected rely heavily on Pin instrumentation \cite{luk2005pin}.
-	As such, MICA is unsuited to analysing HPC workloads with heavy use of parallelism.
-	The Workload ISA-Independent Characterization for Applications (WIICA)~\cite{shao2013isa} extends MICA to present a framework to analyse single-threaded programs independent of the instruction set architecture (ISA).
-
-	Kim and Shrivastava \cite{kim2011cumapz} present CuMAPz, a CUDA memory access profiling tool to guide NVIDIA GPU optimizations.
-	CuMAPz focuses on the problem of improving CUDA application performance using NVIDIA memory-based optimizations.
-	CuMAPz analyses CUDA codes structurally and simulates code execution on the memory hierarchy of specific NVIDIA GPU models.
-
-	During the simulation of the target code, CuMAPz records all memory accesses in various buffers to simulate the global and shared memories on NVIDIA GPUs.
-	Using this detailed simulation data, CuMAPz can estimate the performance-critical (i) shared memory data reuse profit, (ii) profit from coalesced access (iii) memory channel skew cost and (iv) bank conflict cost characteristics of the target code.
-
-	The simulation environment used by CuMAPz and the attached analysis framework is highly specific to CUDA enabled GPUs.
-	Replicating the CuMAPz framework for all target architectures is challenging.
-	However, CuMAPz is an interesting simulator from the standpoint of this paper's work since it adheres to the core parts of NVIDIA GPUs' memory models in its analysis, while allowing the user to specify their GPU model specific hardware information.
-
-	%TODO: Are we making our work sound too much like CuMAPz? This is of course not true and we should fix Reviewer 2's confusion as to what we are doing differently.
-
-	Our approach is an architecture-independent analysis of memory access patterns to provide metrics correlating to similar performance critical memory access optimizations as CuMAPz.
-	We aim to further the state of the art by providing a framework to guide developers in optimizing OpenCL codes for any given target architecture.
-	To the best of our knowledge, none of the previous works presents a set of performance metrics that accurately characterize memory access patterns of parallel applications independent of the target architecture.
-
-	The Architecture-Independent Workload Characterization (AIWC) tool \cite{beauaiwc} collects a set of instruction set architecture (ISA)-independent features based on those identified by Shao and Brooks \cite{shao2013isa}.
-	AIWC runs as a plugin to the Oclgrind~\cite{price:15} framework, which simulates OpenCL kernel execution on an ideal device according to the OpenCL execution and memory models.
-	AIWC collects metrics of kernel memory access, including simple counts such as the (i) total memory footprint, the total number of unique addresses accessed; and (ii) 90\% memory footprint, the number of unique addresses covering 90\% of memory accesses.
-	While these metrics are architecture-independent, they are correlated to program performance on typical architectures.
-	For example, a small ratio of 90\% memory footprint to total memory footprint indicates that a program accesses a small subset of memory addresses frequently, which is highly beneficial for performance in a cached memory hierarchy.
-
-	AIWC also records the global memory address entropy (GMAE), a positive real number corresponding to the randomness of the memory access distribution of a program.
-To measure locality of memory accesses, AIWC collects the local memory address entropy (LMAE) of memory addresses accessed after dropping $n$ least significant bits of all memory addresses accessed by the program.
-To calculate this, AIWC collects a frequency distribution of all non-register memory accesses by all threads in the target kernel.
-Using the collected frequency distribution, AIWC calculates 10 separate local memory address entropy (LMAE) values according to increasing the number of least significant bits (LSB) skipped, using
-	%in the following fashion.
-The information entropy of the collection of memory addresses accessed, after dropping $n$ LSBs and accounting for the number of memory accesses performed to each address is counted as the $n$-bits skipped LMAE.
-Given the frequency distribution of memory addresses accessed by the program,
-	the explicit formula for the $n$-bits skipped LMAE:
-
-	\begin{equation}
-	LMAE_{n-bits} = \sum_{a \in A_n}p_a \log_2(p_a^{-1})
-	\end{equation}
-
-	\begin{itemize}
-		\item $A_n$ is the set of all addresses accessed after skipping $n$ LSBs of each address.
-		\item $p_a := \frac{\#access_a}{\#access_{total}}$ is the probability (calculated as relative frequency) at which each memory address is accessed.
-	\end{itemize}
-
-	LMAE measures the locality of memory accesses performed over the full execution of a program.
-A steeper drop in entropy with increasing number of bits may correlate to more localized memory accesses over the program's execution.
-
-	\begin{figure*}[ht!]
+	\begin{figure*}[h!]
 		\centering
@@
 
+	We aim to capture the essence of vendor-recommended optimizations for target architectures in our metrics.
+	We demonstrate the optimization strategies using an OpenCL kernel which multiplies square matrices of order $N$.
+	We start with a simple unoptimized kernel, and then improve the memory access patterns of this kernel by incrementally performing the optimizations recommended in the CUDA Optimization Handbook \cite{cudaoptimisation} for NVIDIA GPUs.
+	These incrementally optimized OpenCL codes are then analysed using AIWC to determine the accuracy in profiling favourable memory access patterns of the current framework and the proposed extensions to AIWC.
+
+	\subsection{Simple Unoptimized Matrix Multiplication}
+
+	The unoptimized matrix multiplication kernel presented in Appendix~\ref{simpleMultiply} is used as a baseline to validate the performance improvements measured from each following optimization.
+	Each thread of the kernel updates the $(globalRow, globalCol)$ element of matrix $C$ by computing the dot product of the corresponding row of $A$ and column of $B$.
+With square matrices of order $2048$, we observe an effective performance of 131 GFLOP/s on an NVIDIA Tesla P100 GPU.
+
+	\subsection{Using Shared Memory to Coalesce Global Memory Access to Matrix A}
+
+	We first notice that the number of global memory accesses to matrices A and B increases as $O(N^3)$ with respect to matrix size $N$.
+	Global memory is typically located off-chip and accesses induce large delays.
+	NVIDIA GPUs coalesce global memory loads and stores issued within thread-groups into as few DRAM transactions as possible.
+	Multiple global memory loads and stores are coalesced into a single transaction when certain device-specific conditions are met. 
+	On most NVIDIA GPUs, data accesses are coalesced when multiple requests are made for memory locations from the same cache line in global memory~\cite{cudamanual}.
+	Appendix~\ref{coalescedAMultiply} contains the \texttt{coalescedA} kernel, which coalesces accesses to matrix $A$ by storing \emph{tiles}, or square blocks, of $A$'s values into shared memory (OpenCL \texttt{\_\_local} memory).
+	We see an improvement to 600 GFLOP/s in the effective performance for order 2048 matrix multiplication.
+
+	% 9.2.1.2: Misaligned access pattern
+
+	\subsection{Using Shared Memory to store Tiles of Matrix B}
+
+	The code is further optimized by improving the locality of reads from Matrix $B$.
+	While in the previous kernel each thread reads only a single element of matrix $A$ from global memory, each thread reads a full column of matrix $B$.
+	The repeated reads of elements of matrix $B$ can be shared between threads in a work-group by reading tiles from matrix $B$ into shared memory.
+	Appendix~\ref{coalescedABMultiply} contains the \texttt{coalescedAB} kernel which performs this optimization.
+	Performance is further improved to 614 GLOP/s after this optimization.
+
+	\subsection{Optimizing Handling of Shared Memory}
+
+	NVIDIA shared memory is divided into multiple banks -- stored in independent memory modules -- to allow parallel memory access.
+	Bank conflicts occur when shared memory in the same bank is accessed concurrently.
+	The code in \texttt{coalescedAB} kernel is further optimized by implicitly transposing tiles of Matrix $A$ while loading from global memory.
+This improves memory bank utilization during reads to shared memory.
+The \texttt{coalescedABT} kernel demonstrates this optimization by modifying lines 20,21 of \texttt{coalescedAB} to:
+
+	\begin{lstlisting}[firstnumber=20]
+		ASub[localCol][localRow] = A[tiledRow];
+		BSub[localRow][localCol] = B[tiledCol];
+	\end{lstlisting}
+
+	We achieve 1061 GFLOP/s after applying all the recommended GPU memory optimizations to the simple matrix multiply kernel.
+
+	\begin{comment}
+	\subsection{Alignment of Memory Allocation}
+
+	Memory access alignment is important to best utilize all parts of GPU memory architecture.
+Global memory buffer alignment can allow threads to access blocks of global memory aligned to the nearest cache line.
+This enables coalescing of memory accesses.
+If buffers are misaligned, parallel memory requests may cross over cache lines, and may double the number of slow global memory accesses needed as demonstrated in figures \ref{fig: Simple memory alignment} and \ref{fig: Simple memory misalignment}.
+
+	\begin{figure}[h]
+		\includegraphics[width=\linewidth]{aligned_access_coalescing.pdf}
+		\caption{All threads access memory aligned to nearest cache line in parallel \cite{cudaoptimisation}}
+		\label{fig: Simple memory alignment}
+	\end{figure}
+
+	\begin{figure}[h]
+		\includegraphics[width=\linewidth]{misaligned_access_coalescing.pdf}
+		\caption{Unaligned sequential memory addresses fit in two cache lines \cite{cudaoptimisation}}
+		\label{fig: Simple memory misalignment}
+	\end{figure}
+
+	A similar principle applies when using shared memory.
+Bank conflicts may be reduced by aligning allocations of shared memory buffers.
+The \texttt{alignedABT} kernel improves the alignment of shared memory tiles in \texttt{coalescedABT}.
+	An arbitrary large alignment value of 4096 is chosen as it is larger than the cache line size of modern hardware and attributed to the local array declarations in the existing code as shown below.
+
+	\begin{lstlisting}
+		__local float aTile[TILE_DIM][TILE_DIM] __attribute__ ((aligned(4096)));
+		__local float bTile[TILE_DIM][TILE_DIM] __attribute__ ((aligned(4096)));
+	\end{lstlisting}
+	\end{comment}
+	All the examples above are an incomplete listing of optimization strategies developers can apply to code targeted at NVIDIA GPUs.
+	In section~\ref{method}, we will compare the collected AIWC metrics with the expected effect of each optimization to examine how effectively our methodology uncovers underlying bottlenecks and guides optimization efforts.
+
+	\section{Related Work} \label{related work}
+
+	Hoste and Eeckout \cite{hoste2007microarchitecture} show that while conventional architecture-dependent characteristics are useful in locating performance bottlenecks, they can hide the underlying, inherent program behaviour causing performance bottlenecks.
+	A conceptual understanding of the performance characteristics of complex codes is necessary for the programmer to effectively optimize these codes.
+	Architecture-dependent characteristics typically include instructions per cycle (IPC), cache and branch prediction miss rates, page faults and DRAM bus data transfer rates.
+These are typically collected using hardware performance counters available on most target architectures.
+These performance counters do not serve to guide optimization beyond highlighting potential bottlenecks \cite{hoste2007microarchitecture, ganesan2008performance}.
+	Further, in many cases, architecture-dependent characteristics cannot be directly correlated to specific code patterns.
+	For example, the causes of high cache miss rates in the execution of a program are complex and depend on microarchitecture specific features such as cache size, prefetch behaviour and cache placement policies.
+	A HPC developer tasked with optimizing code for a given hardware target would benefit from architecture-independent metrics of the code that can be used to measure the effect of code modifications.
+    These metrics would help to guide the developer in both finding and fixing performance bottlenecks.
+
+	To address the limitations of conventional microarchitecture-dependent characteristics, Hoste and Eeckout~\cite{hoste2007microarchitecture} developed the Microarchitecture-Independent Workload Characterization tool (MICA).
+	They observed that performance counter-based approaches to profiling codes often failed to find underlying program features that map to improved or worsened usage of performance-critical hardware features of the target architecture.
+	The MICA framework is a holistic characterization tool, and thus collects features including instruction mix, instruction-level parallelism, register traffic, data stream strides and branch predictability.
+
+	Of these metrics, data stream strides are of particular interest in memory access pattern profiling.
+	MICA's stride length metric measures the distance between consecutive memory accesses in a single-threaded application.
+	For CPU architectures running single-threaded applications, this metric correlates to the spatial locality of memory accesses -- a measure of how closely bunched are memory access in nearby times.
+	This is directly correlated to cache reuse rates, critical to code performance on CPUs \cite{inteloptimisation}. %Intel optimization manual.
+
+	The MICA approach was tailored for single-threaded applications as the metrics collected rely heavily on Pin instrumentation \cite{luk2005pin}.
+	As such, MICA is unsuited to analysing HPC workloads with heavy use of parallelism.
+	The Workload ISA-Independent Characterization for Applications (WIICA)~\cite{shao2013isa} extends MICA to present a framework to analyse single-threaded programs independent of the instruction set architecture (ISA).
+
+	Kim and Shrivastava \cite{kim2011cumapz} present CuMAPz, a CUDA memory access profiling tool to guide NVIDIA GPU optimizations.
+	CuMAPz focuses on the problem of improving CUDA application performance using NVIDIA memory-based optimizations.
+	CuMAPz analyses CUDA codes structurally and simulates code execution on the memory hierarchy of specific NVIDIA GPU models.
+
+	During the simulation of the target code, CuMAPz records all memory accesses in various buffers to simulate the global and shared memories on NVIDIA GPUs.
+	Using this detailed simulation data, CuMAPz can estimate the performance-critical (i) shared memory data reuse profit, (ii) profit from coalesced access (iii) memory channel skew cost and (iv) bank conflict cost characteristics of the target code.
+
+	The simulation environment used by CuMAPz and the attached analysis framework is highly specific to CUDA enabled GPUs.
+	Replicating the CuMAPz framework for all target architectures is challenging.
+	However, CuMAPz is an interesting simulator from the standpoint of this paper's work since it adheres to the core parts of NVIDIA GPUs' memory models in its analysis, while allowing the user to specify their GPU model specific hardware information.
+
+	%TODO: Are we making our work sound too much like CuMAPz? This is of course not true and we should fix Reviewer 2's confusion as to what we are doing differently.
+
+	Our approach is an architecture-independent analysis of memory access patterns to provide metrics correlating to similar performance critical memory access optimizations as CuMAPz.
+	We aim to further the state of the art by providing a framework to guide developers in optimizing OpenCL codes for any given target architecture.
+	To the best of our knowledge, none of the previous works presents a set of performance metrics that accurately characterize memory access patterns of parallel applications independent of the target architecture.
+
+	The Architecture-Independent Workload Characterization (AIWC) tool \cite{beauaiwc} collects a set of instruction set architecture (ISA)-independent features based on those identified by Shao and Brooks \cite{shao2013isa}.
+	AIWC runs as a plugin to the Oclgrind~\cite{price:15} framework, which simulates OpenCL kernel execution on an ideal device according to the OpenCL execution and memory models.
+	AIWC collects metrics of kernel memory access, including simple counts such as the (i) total memory footprint, the total number of unique addresses accessed; and (ii) 90\% memory footprint, the number of unique addresses covering 90\% of memory accesses.
+	While these metrics are architecture-independent, they are correlated to program performance on typical architectures.
+	For example, a small ratio of 90\% memory footprint to total memory footprint indicates that a program accesses a small subset of memory addresses frequently, which is highly beneficial for performance in a cached memory hierarchy.
+
+	AIWC also records the global memory address entropy (GMAE), a positive real number corresponding to the randomness of the memory access distribution of a program.
+To measure locality of memory accesses, AIWC collects the local memory address entropy (LMAE) of memory addresses accessed after dropping $n$ least significant bits of all memory addresses accessed by the program.
+To calculate this, AIWC collects a frequency distribution of all non-register memory accesses by all threads in the target kernel.
+Using the collected frequency distribution, AIWC calculates 10 separate local memory address entropy (LMAE) values according to increasing the number of least significant bits (LSB) skipped using
+	%in the following fashion.
+The information entropy of the collection of memory addresses accessed, after dropping $n$ LSBs and accounting for the number of memory accesses performed to each address is counted as the $n$-bits skipped LMAE.
+Given the frequency distribution of memory addresses accessed by the program,
+	the explicit formula for the $n$-bits skipped LMAE:
+
+	\begin{equation}
+	LMAE_{n-bits} = \sum_{a \in A_n}p_a \log_2(p_a^{-1})
+	\end{equation}
+
+	\begin{itemize}
+		\item $A_n$ is the set of all addresses accessed after skipping $n$ LSBs of each address.
+		\item $p_a := \frac{\#access_a}{\#access_{total}}$ is the probability (calculated as relative frequency) at which each memory address is accessed.
+	\end{itemize}
+
+	LMAE measures the locality of memory accesses performed over the full execution of a program.
+A steeper drop in entropy with increasing number of bits may correlate to more localized memory accesses over the program's execution.
+
 	\section{Non-Parallel Memory Metrics} \label{method}
+
+	\begin{table*}[h!]
+		\centering
+		\begin{tabular}[t]{lrrrrr}
+			\toprule
+			& simple & coalescedA & coalescedAB & coalescedABT \\ \midrule %& alignedABT \\ \midrule 
+			Total memory footprint        & 196608 & 196608     & 196608      & 196608      \\ %\hline
+			90\% Memory Footprint         & 118196 & 56176      & 489         & 489         \\ %\hline
+			Global MAE                    & 17.02  & 13.18      & 9.78        & 9.78        \\ %\hline
+			LMAE \#bits=3                 & 16.02  & 12.18      & 8.78        & 8.78        \\ %\hline
+			LMAE \#bits=10                & 9.02   & 5.18       & 1.78        & 1.78        \\ %\hline
+			Relative Local Memory Usage   & 0      & 0.50       & 0.94       &  0.94        \\ \bottomrule
+		\end{tabular}
+		\caption{Selection of AIWC\cite{beauaiwc} metrics for 256 $\times$ 256 matrix multiplication}
+		\label{matmul aiwc table}
+	\end{table*}
 
@@
 
+	\flushbottom
+
 	\section{A Parallel Spatial Locality Metric} \label{discussion}
-
-	Aggregate metrics of the kind presented by AIWC necessarily present a simplified view of program behaviour, omitting many details.
-	Different ways of aggregating program measurements lead to different features of program execution being emphasized in the final metrics.
-	For example, the calculation of memory address entropy described in section~\ref{related work} relies only on the frequency distribution of memory accesses to all addresses accessed by the kernel, and discards temporal information.
-	The order of sequential memory accesses performed by each thread, as well the relationship between work-items in an OpenCL work-group, are both vital in accurately characterizing parallel codes.
-
-	We propose a new architecture-independent metric, \emph{parallel spatial locality}, to measure memory access patterns in parallel programs.
-	The proposed metric is inspired by CuMAPz' direct approaches to measuring optimization specific characteristics of CUDA codes.
-
-	During simulation, AIWC collects a list of all memory accesses by each thread of execution.
-	In the OpenCL programming model, threads within a work-group execute the same code and share access to \textit{local memory}.
-	We can group together memory accesses of threads in a work-group at each logical time step in the symbolic execution of the code.
-	On a GPU, memory accesses executed at the same time by different threads in a work-group are likely to interact, determining the extent of memory access coalescing and bank conflicts.
 
@@
 		\end{tikzpicture}
-		\caption{Parallel spatial locality metric obtained from AIWC for matrix multiply kernels for 256 $\times$ 256 matrix multiplication}
+		\caption{Parallel spatial locality metric for matrix multiply kernels for 256 $\times$ 256 matrix multiplication}
 		\label{fig: matmul parallel spatial locality}
 	\end{figure*}
+
+	Aggregate metrics of the kind presented by AIWC necessarily present a simplified view of program behaviour, omitting many details.
+	Different ways of aggregating program measurements lead to different features of program execution being emphasized in the final metrics.
+	For example, the calculation of memory address entropy described in section~\ref{related work} relies only on the frequency distribution of memory accesses to all addresses accessed by the kernel, and discards temporal information.
+	The order of sequential memory accesses performed by each thread, as well the relationship between work-items in an OpenCL work-group, are both vital in accurately characterizing parallel codes.
+
+	We propose a new architecture-independent metric, \emph{parallel spatial locality}, to measure memory access patterns in parallel programs.
+	The proposed metric is inspired by CuMAPz' direct approaches to measuring optimization specific characteristics of CUDA codes.
+
+	During simulation, AIWC collects a list of all memory accesses by each thread of execution.
+	In the OpenCL programming model, threads within a work-groups execute the same code and share access to \textit{local memory}.
+	We can group together memory accesses of threads in a work-group at each logical time step in the symbolic execution of the code.
+	On a GPU, memory accesses executed at the same time by different threads in a work-group are likely to interact, determining the extent of memory access coalescing and bank conflicts.
 
@@
 	This dependency enforces a wave-front computation pattern, travelling along the main diagonal of the matrix.
-Each iteration of the kernels computes over an anti-diagonal of the matrix, starting at its top left corner, and finishing at its bottom right corner.
+Each iteration of the kernels computes over an antidiagonal of the matrix, starting at its top left corner, and finishing at its bottom right corner.
 
-	In particular, the wave-front pattern of computations causes each work-item to request distant memory addresses in a parallel fashion.
+	In particular, the wavefront pattern of computations causes each work-item to request distant memory addresses in a parallel fashion.
 Thus, parallel memory requests into the input matrix prohibit locality.
@@
 	This leads to poor parallel spatial locality. Thus, we see a slow drop-off in parallel spatial locality for \texttt{BFS kernel 1}.
-
 	\subsection{Sparse Linear Algebra: CSR}
@@
 
-	 Codes with a mix of favourable and unfavourable access patterns (typically involving two-dimensional matrices such as LUD and SRAD) will have a steep drop-off in PSL to about half of the starting PSL metric as $n$ increases, and the curve flattens out at that point.
+	 Codes with a mix of favourable and unfavourable access patterns (typically involving two-dimensional matrices such as in LUD and SRAD) will have a steep drop-off in PSL to about half of the starting PSL metric as $n$ increases, and the curve flattens out at that point.
 
-	 In the extreme case, when parallel memory accesses are entirely disparate as for Needleman–Wunsch, we observe a flat parallel spatial locality graph.
-	 The worst possible memory access pattern: all threads in a work-group accessing results in a flat PSL metric of 1.
-	 The PSL metric for Needleman–Wunsch is a flat line at 0.53. This indicates that parallel memory accesses are entirely disparate.
+	 In the extreme case, when parallel memory accesses are very distant as for Needleman-Wunsch, we observe a flat parallel spatial locality graph. 
+	 The worst possible memory access pattern occurs when all threads in a work-group access distant memory locations. This results in a flat PSL metric of 1. 
+	 The PSL metric for Needleman-Wunsch is a flat line at 0.53. This indicates that parallel memory accesses are entirely disparate. 
 	 However, the workload imbalance across threads in each OpenCL work-group eases pressure on the memory unit when not all threads are active.
@@
 This is reflective of the mostly disparate parallel memory access pattern in these codes.
-The increasing steepness of the drop-off at higher $n$ in codes with data-driven memory accesses is reflective of the size of buffers being operated upon as the data structures used during computation cause localisation of memory accesses.
+The increasing steepness of the drop-off at higher $n$ in codes with data-driven memory accesses is reflective of the size of buffers being operated upon as the data structures used during computation cause localization of memory accesses.
 

### 2020-03-21 paper/paper.tex

@@
 	This dependency enforces a wave-front computation pattern, travelling along the main diagonal of the matrix.
-Each iteration of the kernels computes over an antidiagonal of the matrix, starting at its top left corner, and finishing at its bottom right corner.
+Each iteration of the kernels computes over an anti-diagonal of the matrix, starting at its top left corner, and finishing at its bottom right corner.
 
-	In particular, the wavefront pattern of computations causes each work-item to request distant memory addresses in a parallel fashion.
+	In particular, the wave-front pattern of computations causes each work-item to request distant memory addresses in a parallel fashion.
 Thus, parallel memory requests into the input matrix prohibit locality.
